
<!-- Indicator Chapter for INDICATOR -->


```{r setup-INDICATOR, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
source("global.R")
# indicator <- ""

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# INDICATOR

<!-- Metadata -->
**Data Type:** 

**Spatial Scope:**

**Duration** 

**Source:** 

**Contact:** 



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->



<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-INDICATOR, fig.cap="", echo=TRUE}


```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-INDICATOR, include=F}


```



<!---------- Trends Text ----------->



<!--------- Trends Table--------->
### Summary Table


|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value  |  | |
|Recent Trend  |  |   |
|Overall Trend  |  |  |
|Highest Value |  | |
|Lowest Value |  |  |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->


<!----------- Column Definitions Section -------------->
## Variable Definitions



|variable   |description |unit |
|:----------|:-----------|:----|
|year       |            |     |
|region     |            |     |
|mean_value |            |     |

<!----------- Additional Data Section -------------->
## Additional Data




<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-INDICATOR, include=T, echo=T, eval=F}
library(marea)
data('INDICATOR')
plot(INDICATOR)
```


```{r}
# remove all created objects
rm(list = ls())
```
