
<!-- Indicator Chapter for amo: Atlantic MultiDecadal Oscillation Index -->


```{r setup-amo, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "amo"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Atlantic MultiDecadal Oscillation Index

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Northern Hemisphere (0-60N)

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator

The AMO describes long-duration changes in the sea surface temperature of the North Atlantic Ocean characterized by extended cool (negative) and warm (positive) phases that may last for 20-60 years [@knight2006climate]. AMO values are computed from detrended sea surface temperature anomalies, such that the resultant value offers a description of large-scale temperature variation that is independent of gradual warming as a result of climate change. 

AMO is linked with several oceanographic and atmospheric processes including air temperature and precipitation, storm strength, and variability in ocean currents. 




<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-amo, echo=TRUE}
library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1),
         smooth_1yr = zoo::rollapply(SSTA_value, mean, width = 12, partial = TRUE),
         smooth_10yr = zoo::rollapply(SSTA_value, mean, width = 120, partial = TRUE),
         overall_mean = mean(SSTA_value))

p <- plot_ly(plotly_df, x = ~date) %>%
  add_lines(y = ~SSTA_value,
            name = "Monthly Anomaly",
            line = list(color = "lightgrey"),
            hovertemplate = "Monthly anomaly: %{y:.2f}<extra></extra>"
           ) %>%
  add_lines(y = ~smooth_1yr,
            name = "1-yr Smooth",
            line = list(color = "blue", width = 2),
            hovertemplate = "1-yr smoothed: %{y:.2f}<extra></extra>",
            ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "red", width = 2),
            hovertemplate = "10-yr smoothed: %{y:.2f}<extra></extra>",
           ) %>%
    add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Overall Mean: %{y:.2f}<extra></extra>",
           ) %>%
  layout(
    title = "Atlantic Multidecadal Oscillation Index for Northern Hemisphere",
    xaxis = list(title = "Date"),
    yaxis = list(title = "AMO Index",
                 fixedrange = TRUE),
    hovermode = "x unified",
    margin = list( t = 80)
    
  ) %>%
  config(displayModeBar = FALSE) 

p


```


```{r, include=F}
decadal_means <- data@data %>%
  mutate(decade = stringr::str_sub(year, 1,3)) %>%
  group_by(decade) %>%
  mutate(decade_start = first(year),
         decade_end = last(year),
         decade_mean = mean(SSTA_value)) %>%
   mutate(date = lubridate::make_date(year, month, day = 1))
  
p %>%
  add_lines(data = decadal_means,
            x = ~date,
            y = ~decade_mean,
            split = ~decade,
            line = list(color = "black",
                        width = 3),
            showlegend= F,
            text = ~paste0("Decadal mean: ", round(decade_mean,2)),
            hoverinfo = "text") %>%
  layout( hovermode = "x unified")
```




<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-amo, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
current_vs_mean <- last_val - overall_mean
current_vs_mean_text <- ifelse(sign(current_vs_mean) == 1,"above","below")
last_different_sign <- switch(current_vs_mean_text,
                                  "above" = last(which(data@data[value_col] < overall_mean)),
                                  "below" = last(which(data@data[value_col] > overall_mean)))
last_date_different_sign <- ifelse(has_months,
                                       paste0(data@data[last_different_sign,"month"],"/",data@data[last_different_sign,"year"]),
                                       paste0(data@data[last_different_sign,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the AMO value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The AMO value is currently **`r current_vs_mean_text` the long-term mean**, and has been **since `r last_date_different_sign`**.
The AMO value has followed a(n) **`r current_trend` trend** in recent years.



<!--------- Trends Table--------->
### Summary Table
|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. The most recent value is `r abs(round(current_vs_mean,2))` `r current_vs_mean_text` the long-term mean.|
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, amo has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, amo has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Highest Value | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Lowest Value | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

AMO can both directly and indirectly affect marine ecosystem function [@nye2014ecosystem].  

In Canada's maritimes region, AMO is closely linked to several ecological trends with relevance to fisheries and stock assessments. AMO has shown negative correlations with primary productivity and fish biomass in Eastern Canada (NAFO zones 4X, 5YB, Western Scotian Shelf, and Bay of Fundy) [@araujo2012effects], and has been hypothesized a limitation on stock size despite moratoriums on fishing activity for commercially important species like Atlantic salmon [@condron2005multidecadal].

Still, specific effects of AMO on marine resources can vary aross space. For example, AMO has displayed a positive relationship with catch per unit effort of bluefin Tuna in Canadian regions (Gulf of St. Lawrence and Nova Scotia), and a contrastng negative correlation in the United States [@hansell2020incorporating]. Impacts of AMO on focal species are therefore likely related to thresholds of thermal performance of a focal species within an ecosystem. 



<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable   |description |unit |
|:----------|:-----------|:----|
|year       |year of model-fit value          |  |
|month      |month of model-fit value         |  |
|SSTA_value |AMO index         |  |



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-amo, include=T, echo=T, eval=F}
library(marea)
data('amo')
plot(amo)
```
