
<!-- Indicator Chapter for amo: Atlantic MultiDecadal Oscillation Index -->


```{r setup-amo, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "amo"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Atlantic MultiDecadal Oscillation Index

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Northern Hemisphere (0-60N)

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator

The AMO describes long-duration changes in the sea surface temperature of the North Atlantic Ocean characterized by extended cool (negative) and warm (positive) phases that may last for 20-60 years [@knight2006climate]. Positive AMO phases are associated with warm waters in the North Atlantic, reduced precipitation in the Southern US, and increased hurricane activity in the Tropical Atlantic; negative AMO phases are associated with roughly the opposite [@knight2006climate].

AMO values are computed from detrended sea surface temperature anomalies, such that the resultant value offers a description of large-scale temperature variation that is independent of gradual warming as a result of climate change. 

AMO is linked with several oceanographic and atmospheric processes including air temperature and precipitation, storm strength, and variability in ocean currents. 




<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-amo, fig.cap=paste0("AMO Index Value; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)) , echo=TRUE}
library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1),
         smooth_1yr = zoo::rollapply(SSTA_value, mean, width = 12, partial = TRUE),
         smooth_10yr = zoo::rollapply(SSTA_value, mean, width = 120, partial = TRUE),
         overall_mean = mean(SSTA_value))

p <- plot_ly(plotly_df, x = ~date) %>%
  add_lines(y = ~SSTA_value,
            name = "Monthly Anomaly",
            line = list(color = "lightgrey"),
            hovertemplate = "Monthly anomaly: %{y:.2f}<extra></extra>"
           ) %>%
  add_lines(y = ~smooth_1yr,
            name = "1-yr Smooth",
            line = list(color = "blue", width = 2),
            hovertemplate = "1-yr smoothed: %{y:.2f}<extra></extra>",
            ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "red", width = 2),
            hovertemplate = "10-yr smoothed: %{y:.2f}<extra></extra>",
           ) %>%
    add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Overall Mean: %{y:.2f}<extra></extra>",
           ) %>%
  layout(
    title = "Atlantic Multidecadal Oscillation Index for Northern Hemisphere",
    xaxis = list(title = "Date"),
    yaxis = list(title = "AMO Index",
                 fixedrange = TRUE),
    hovermode = "x unified",
    margin = list( t = 80)
    
  ) %>%
  config(displayModeBar = FALSE) 

p


```


```{r, include=F}
decadal_means <- data@data %>%
  mutate(decade = stringr::str_sub(year, 1,3)) %>%
  group_by(decade) %>%
  mutate(decade_start = first(year),
         decade_end = last(year),
         decade_mean = mean(SSTA_value)) %>%
   mutate(date = lubridate::make_date(year, month, day = 1))
  
p %>%
  add_lines(data = decadal_means,
            x = ~date,
            y = ~decade_mean,
            split = ~decade,
            line = list(color = "black",
                        width = 3),
            showlegend= F,
            text = ~paste0("Decadal mean: ", round(decade_mean,2)),
            hoverinfo = "text") %>%
  layout( hovermode = "x unified")
```




<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*

```{r calc-trends-amo, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current anomaly is in a positive or negative phase
last_anomaly <- last(data@data)[[value_col]]
current_phase <- ifelse(last_anomaly < 0,"negative","positive")

streak_start <- switch(
  current_phase,
  "negative" = last(which(data@data[value_col] > 0)),
  "positive" = last(which(data@data[value_col] < 0))) + 1
streak_start_date <- ifelse(
  has_months,
  paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
  paste0(data@data[streak_start,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r print_date(last_date)`**, the AMO value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The AMO is currently in a **`r current_phase` phase**, and has been since `r print_date(streak_start_date)`.
The AMO value has followed an **`r current_trend` trend** in recent years (Fig. \@ref(fig:plot-amo)).



<!--------- Trends Table--------->
### Summary Table

Summary values for the Atlantic Multidecadal Oscillation are found in the table below (Table \@ref(tab:amo-trends-table))

Table: (\#tab:amo-trends-table) Summary table of AMO values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r print_date(last_date)`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. |
|Timeseries Record High | `r highest_val` | The highest value in the timeseries was recorded on **`r print_date(highest_timepoint)`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Timeseries Record Low | `r lowest_val` | The lowest value in the timeseries was recorded on **`r print_date(lowest_timepoint)`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

AMO can both directly and indirectly affect marine ecosystem function [@nye2014ecosystem].  

In Canada's maritimes region, AMO is closely linked to several ecological trends with relevance to fisheries and stock assessments. AMO has shown negative correlations with primary productivity and fish biomass in Eastern Canada (NAFO divisions 4X, 5YB, Western Scotian Shelf, and Bay of Fundy) [@araujo2012effects], and has been hypothesized a limitation on stock size despite moratoriums on fishing activity for commercially important species like Atlantic salmon [@condron2005multidecadal].

Still, specific effects of AMO on marine resources can vary aross space. For example, AMO has displayed a positive relationship with catch per unit effort of bluefin Tuna in Canadian regions (Gulf of St. Lawrence and Nova Scotia), and a contrastng negative correlation in the United States [@hansell2020incorporating]. Impacts of AMO on focal species are therefore likely related to thresholds of thermal performance of a focal species within an ecosystem. 



<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:amo-variable-definitions) Column names and definitions in the AMO dataset.

|variable   |description |unit |
|:----------|:-----------|:----|
|year       |year of model-fit value          |  |
|month      |month of model-fit value         |  |
|SSTA_value |AMO index         |  |


<!----------- Additional Data Section -------------->
## Additional Data

No additional data for AMO.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-amo, include=T, echo=T, eval=F}
library(marea)
data('amo')
plot(amo)
```


```{r}
# remove all created objects
rm(list = ls())
```

