
<!-- Indicator Chapter for ao: Arctic Oscillation Index -->


```{r setup-ao, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "ao"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Arctic Oscillation Index

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Northern Hemisphere

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator

The Arctic Oscillation describes the variability in atmospheric pressure between the Arctic and surrounding mid-latitude regions of the Northern Hemisphere, and is linked to fluctuations in the strength of the polar vortex [@thompson1998arctic]. The trend of AO over time does not follow a particular periodicity.

Effects of strong AO anomalies are most pronounced in the Northern Hemisphere winter. Positive AO anomaly values represent lower-than-average pressure in Arctic regions and higher-than-average pressure in mid-latitudes, resulting in a stronger polar vortex and contained winter storm activity in the Arctic region. Negative AO values represent the opposite -- higher-than-average pressures in the Arctic and lower-than-average pressures in the mid-latitudes. In negative phases, the jet stream is weaker and can meander, leading to anomalous cold events in mid-latitudes, particularly in Eastern North America and Eurasia [@li2017circulation; @he2017impact].

The AO is highly correlated with the North Atlantic Oscillation (NAO) and some discussion exists around whether the two are distinct (refs), but the AO index better reflects changes occurring at high arctic latitudes. 

<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-ao, fig.cap=paste0("AO Index Value; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}
library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1),
         smooth_1yr = zoo::rollapply(anomaly_value, mean, width = 12, partial = TRUE),
         smooth_10yr = zoo::rollapply(anomaly_value, mean, width = 120, partial = TRUE),
         overall_mean = mean(anomaly_value))

p <- plot_ly(plotly_df, x = ~date) %>%
  add_lines(y = ~anomaly_value,
            name = "Monthly Anomaly",
            line = list(color = "lightgrey"),
            hovertemplate = "Anomaly Value: %{y:.2f} <extra></extra>"
  ) %>%
  add_lines(y = ~smooth_1yr,
            name = "1-yr Smooth",
            line = list(color = "blue", width = 2),
            hovertemplate = "1-yr Smooth: %{y:.2f} <extra></extra>"
  ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "red", width = 2),
            hovertemplate = "10-yr Smooth: %{y:.2f} <extra></extra>") %>%
  add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Timeseries Mean: %{y:.2f} <extra></extra>") %>%
  layout(
    title = "Arctic Oscillation Index for Northern Hemisphere",
    xaxis = list(title = "Date",
                 hoverformat = "%b %Y" ),
    yaxis = list(title = "AO Index",
                 fixedrange = T),
    hovermode = "x unified",
    margin = list(t = 80)
    
  ) %>%
  config(displayModeBar = FALSE)

p 

```

```{r, include=F}
decadal_means <- data@data %>%
  mutate(decade = stringr::str_sub(year, 1,3)) %>%
  group_by(decade) %>%
  mutate(decade_start = first(year),
         decade_end = last(year),
         decade_mean = mean(anomaly_value)) %>%
   mutate(date = lubridate::make_date(year, month, day = 1))
  
p %>%
  add_lines(data = decadal_means,
            x = ~date,
            y = ~decade_mean,
            split = ~decade,
            line = list(color = "black",
                        width = 3),
            showlegend= F,
            text = ~paste0("Decadal mean: ", round(decade_mean,2)),
            hoverinfo = "text") %>%
  layout( hovermode = "x unified")
```





<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*

```{r calc-trends-ao, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
last_anomaly <- last(data@data)[[value_col]]
current_phase <- ifelse(last_anomaly < 0,"negative","positive")

streak_start <- switch(
  current_phase,
  "negative" = last(which(data@data[value_col] > 0)),
  "positive" = last(which(data@data[value_col] < 0))) + 1
streak_start_date <- ifelse(
  has_months,
  paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
  paste0(data@data[streak_start,"year"]))


```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r print_date(last_date)`**, the AO value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The AO is **currently in a `r current_phase` phase**, and has been since `r print_date(streak_start_date)`.
The AO value has followed a **`r current_trend` trend** in recent years.



<!--------- Trends Table--------->
### Summary Table

Summary values for the Arctic Oscillation are found in the table below (Table \@ref(tab:ao-trends-table))

Table: (\#tab:ao-trends-table) Summary table of AO values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r print_date(last_date)`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values.|
<!-- note that reviewers suggested removing linear trends
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, ao has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, ao has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). | -->
|Timeseries Record High | `r highest_val` | The highest value in the timeseries was recorded on **`r print_date(highest_timepoint)`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Timeseries Record Low| `r lowest_val` | The lowest value in the timeseries was recorded on **`r print_date(lowest_timepoint)`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Direct examples of AO's effect on fisheries are less common compared to other atmospheric variables, but the index has potential effects to stocks through its association with sea ice coverage and freshwater input in the Arctic and North Atlantic oceans [@greene2013remote]. 

In Canada, some fisheries responses to AO have been identified. Positive springtime AO values have been associated with decreased recruitment and  of juvenile movement arctic cod in the Beaufort Sea, and positive AO years result in increased transport of fish towards eastern waters [@herbig2023climate]. AO has also shown an important predictor of exploitable snow crab biomass, with effects that vary across regions and timescales [@mullowney2023sub].

<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:ao-variable-definitions) Column names and definitions in the AO dataset.

|variable      |description |unit |
|:-------------|:-----------|:----|
|year          |Year of Model-projected value         |   |
|month         |Month of Model-projected value         |   |
|anomaly_value |Standardized index for AO anomalies    |   |

<!----------- Additional Data Section -------------->
## Additional Data

No additional data for AO. 



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-ao, include=T, echo=T, eval=F}
library(marea)
data('ao')
plot(ao)
```


```{r}
# remove all created objects
rm(list = ls())
```
