
<!-- Indicator Chapter for azmp_satellite_temperature: Surface Temperature from Satellite -->


```{r setup-azmp_satellite_temperature, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

global_cols2 <- global_cols2 %>%
  mutate(region = recode(region, "4X" = "4XSS")) 

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "azmp_satellite_temperature"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))

timespan <- marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)
```



<!------- Header Section -------------->
<!-- Title -->
# Surface Temperature from Satellite Data (AZMP) {#azmp-satellite-temperature}

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Scotian Shelf (4X, 4V, 4W)

**Duration** `r timespan`

**Source:** `r data@meta$source_citation` 

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Sea surface temperature from satellite data refers to continuous sea surface temperature (SST) fields derived from satellite sensors, representing the temperature of the ocean's surface. Here, continuous fields are summarized within non-mutually exclusive regions representing NAFO divisions or subzone divisions. 

Sea surface temperature is influenced by a variety of conditions spanning multiple spatial and temporal scales. In this region, long-term trends in SST are consistent with expectations of anthropogenic climate change [@greenan2019changes]. At the same time, decadal-to-interannual regional SST trends are influenced by atmospheric indicators like those described in this book, and short-term local values can be influenced by many oceanographic and climate factors, such as freshwater input, ocean currents, and weather events.

<!----------- Plots Section -------------->
## View Data

```{r plot-azmp-satellite-temperature, fig.cap=paste0("AZMP Satellite Surface Temperature in Scotian Shelf Regions; ",timespan), echo=T}
library(plotly)
plotly_df <- data@data %>%
  inner_join(global_cols2 )

p <- plot_ly()

regions_order <- c("4Vn","4Vs","4W","4XSS")

for(r in regions_order){

  df_sub <- plotly_df %>% filter(region == r)

  group_name <- unique(df_sub$region_group)

  p <- p %>%
    add_lines(
      data = df_sub,
      x = ~year,
      y = ~mean_value,
      name = r,
      legendgroup = group_name,
      legendgrouptitle = list(text =
        ifelse(r == "4Vn" & group_name == "ESS",
               "Eastern Scotian Shelf Zones",
        ifelse(r == "4XSS",
               "Western Scotian Shelf Zones",
               NA))
      ),
      line = list(color = unique(df_sub$color), width = 2),
      text = ~paste0("<b>",region,"</b>: ",round(mean_value,2),"°C"),
      hoverinfo = "x+text"
    )
}

p %>%
  layout(
    title = "AZMP SST by Division",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Temperature (°C)", fixedrange = TRUE),
    hovermode = "x unified",
    legend = list(
      tracegroupgap = 5,
      groupclick = "toggleitem",
      itemdoubleclick = FALSE
    )
  ) %>%
  config(displayModeBar = FALSE)

```


<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*

<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-azmp_satellite_temperature, include=F}

data@data <- data@data %>% inner_join(global_cols2)

 # get value colname
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]



# calculate mean per region
region_means <- data@data %>%
      group_by(region) %>%
      summarize(mean = mean(get(value_col)),
                sd = sd(get(value_col)),
                se = plotrix::std.error(get(value_col)),
                max_year = year[which.max(get(value_col))],
                max_val = max(get(value_col)),
                min_year = year[which.min(get(value_col))],
                min_val = min(get(value_col)),
                .groups = "drop") %>%
      mutate(region = forcats::fct_reorder(region, mean))

 # find overall mean
overall_mean <- data@data %>%
      summarize(mean = mean(get(value_col))) %>% pull(mean)



# find trends per region
data_split <-  data@data  %>% split(f =  data@data$region)
data_split_mods <- purrr::map(.x = data_split,
                                  .f = ~lm(.x, formula = get(value_col) ~ year))
# find coefficients per region
data_coefficients <- 
  purrr::map(.x = data_split_mods,
             .f = ~coefficients(.x)[[2]] %>% as_tibble) %>%
      bind_rows(.id = "region") %>%
      rename(trend = value)
# find confidence intervals
data_confints <- 
  purrr::map(.x = data_split_mods,
             .f = ~confint(.x)[2,] %>%
               setNames(c("lower","upper"))) %>%
  bind_rows(.id = "region")
# find p values per region
data_p_values <- purrr::map(.x = data_split_mods,
                                .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
      bind_rows(.id = "region") %>%
      rename(p_val = value)


# merge
merge <- region_means %>% 
  merge(data_coefficients) %>%
  merge(data_confints) %>%
  merge(data_p_values) %>%
  arrange(region) %>%
  left_join(global_cols2)

# plot
slopes_p2 <- merge %>%
  ggplot(aes(y = region, x = trend,
              xmin = lower, xmax = upper,
              color = mean)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(linewidth = 1) +
  geom_text(data = . %>% filter(p_val < 0.05),
            label = "*", 
            nudge_y =.15,
            size = 6, color = "black") +
  scale_color_viridis_c() +
  ggthemes::theme_few(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title.position = "plot") +
  labs(y = "NAFO Division",
       x = "Temperature Trend (°C/year)",
       color = "Mean\nTemperature\n(°C)",
       title = "Surface Temperature Trends per Region",
       subtitle = "Asterisks indicate significant trends (p < 0.05)") +
    facet_wrap(region_group_label~.,
               ncol = 1,
               scales = "free_y",
               space = "free_y")  


slopes_p2



# find number of regions increasing and decreasing
n_regions <- length(unique(data@data$region))
n_increasing <- sum(data_coefficients$trend > 0)
n_increasing_sig <- sum(merge$trend > 0 & merge$p_val < 0.05)
n_decreasing <- sum(data_coefficients$trend < 0)
n_significant <- sum(merge$p_val < 0.05)
majority_trend <- ifelse(n_increasing > n_decreasing, "increasing","decreasing")

# find highest and lowest regions
highest_region <- as.character(merge$region[which.max(merge$mean)])
highest_mean <- round(merge$mean[merge$region == highest_region],2)
lowest_region <- as.character(merge$region[which.min(merge$mean)])
lowest_mean <- round(merge$mean[merge$region == lowest_region],2)
highest_trend_region <-  as.character(merge$region[which.max(merge$trend)])
lowest_trend_region <-  as.character(merge$region[which.min(merge$trend)])


# find mean quantile of most recent value
last_year <- max(data@data$year)
data_split_last <- purrr::map(
      .x = data_split,
      .f = ~last(.x)
    )
last_percentile <- purrr::map2(
      .x = data_split,
      .y = data_split_last,
      .f = ~round(ecdf(.x[[value_col]])(.y[[value_col]]) * 100,2) %>%
        as_tibble
    ) %>%
      bind_rows(.id = "region") %>%
      rename(percentile = value)
last_percentile_groups <- last_percentile %>%
      summarize(n_high = sum(percentile > 70 ),
                n_mid = sum(percentile > 30 & percentile < 70),
                n_low = sum(percentile < 30))
last_percentile_mean <- last_percentile %>% summarize(mean = mean(percentile))

last10 <- data@data %>%
  group_by(region) %>%
  summarize(highest_year = year[mean_value == max(mean_value)],
            lowest_year = year[mean_value == min(mean_value)],
            max_year = max(year)) %>%
  ungroup() %>%
  summarize(last10_high = sum(highest_year >= max_year-10),
            last10_low = sum(lowest_year >= max_year-10))

ess_means <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"±", round(se,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_means <- merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"±", round(se,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_trends <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = 
           paste0("<b>",region,"</b>: ", 
                  round(trend,3)," ± ",
                  round((upper-lower)/2,2),
                  "°C/year ",
                  ifelse(p_val < 0.05,"\u2217",""))) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_trends <- merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = 
           paste0("<b>",region,"</b>: ", 
                  round(trend,3)," ± ",
                  round((upper-lower)/2,2),
                  "°C/year ",
                  ifelse(p_val < 0.05,"\u2217",""))) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_max_year <-  merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", max_year," (",round(max_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_max_year <-  merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", max_year," (",round(max_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_min_year <-  merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", min_year," (",round(min_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_min_year <-  merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", min_year," (",round(min_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")



```


Surface temperatures are less variable across the region compared to bottom temperature. Among regions in this dataset, mean surface temperatures range from **`r highest_mean`°C in `r highest_region`** to **`r lowest_mean`°C in `r lowest_region`**. `r last10$last10_high` `r paste0(ifelse(last10$last10_high == 1,"region had its series record","regions had their series records"))`  within the last 10 years of sampling (`r last_year-10`-`r last_year`), and `r last10$last10_low` `r paste0(ifelse(last10$last10_low == 1,"region had its series minimum","regions had their series minima"))` (Fig. \@ref(fig:plot-azmp-satellite-temperature)).

In the most recent year of sampling (`r last_year`), **`r last_percentile_groups[["n_high"]]` regions had values that are high** compared to their overall timeseries (>70th percentile), **`r last_percentile_groups[["n_mid"]]` regions had values that were moderate** (between 30th and 70th percentile), and **`r last_percentile_groups[["n_low"]]` regions had low values** (<30th percentile) within their timeseries.

```{r plot-trends-azmp-satellite-temperature, fig.cap="Coefficients of satellite surface temperature trends over time for NAFO divisions within Eastern and Western Scotian Shelf regions. Error bars represent 95% confidence intervals of model fits. Asterisks indicate significant trends (p < 0.05 for year term)", echo=F}
slopes_p2
```


The overall trend is **`r majority_trend` temperature** across regions. `r n_increasing` regional subdivisions have increasing satellite surface temperature trends (`r n_increasing_sig` statistically significant) and `r n_decreasing` have negative trends (Fig \@ref(fig:plot-trends-azmp-satellite-temperature)).

As with Bottom Temperature, average satellite surface temperature exhibited a general northward decrease between NAFO divisions (Fig \@ref(fig:plot-trends-azmp-satellite-temperature-map)), but warming trends were fairly consistent across zones (Fig. \@ref(fig:plot-trends-azmp-satellite-temperature),\@ref(fig:plot-trends-azmp-satellite-temperature-map)).

```{r plot-trends-azmp-satellite-temperature-map, fig.cap="Map of target study areas colored by mean satellite surface temperature (left) and warming trend over time (right).", echo=F, cache=F}
library(sf)
library(mapgl)

# get shapefiles for regions
#sf <- readRDS("data/derived_data/azmp_area.rds") %>%
#  filter(get(indicator) == T)


#sf_merge <- sf %>%
#  select(area, area_type, geometry) %>%
#  rename(region = area) %>%
#  # merge in modeled values
#  inner_join(merge) 

sf_merge <- readRDS("data/derived_data/nafo_regions_correct.rds") %>%
  inner_join(merge)


# make color scale for trends - symmetrical around zero
maxtrend <- max(abs(sf_merge$trend))
scale_data <- data.frame(trend = c(sf_merge$trend, -maxtrend, maxtrend))
trend_color_scale <- interpolate_palette(
  data = scale_data,
  column = "trend",
  method = "equal",
  n = 5, 
  palette = function(n) rev(RColorBrewer::brewer.pal(n, "RdBu"))  
)


# make color scale for means
mean_color_scale <- interpolate_palette(
  data = sf_merge,
  column = "mean",    
  method = "equal",
  n = 4,
  palette = viridisLite::viridis
)

# make popup label
sf_merge <- sf_merge %>%
  mutate(label = paste(
      "<span style='font-size: 12px;'><b>",region, "</b><br>",
      "<b>Timeseries Mean:</b>", round(mean,2),"°C<br>",
      "<b>Temperature Trend:</b>", round(trend,3),"°C/year<br>",
      "<b>p_val:</b>", ifelse(p_val<0.05,"<0.05", round(p_val,2)),
      "<br>"
    )) %>%
    mutate(label2 = paste0(
    '<div style="',
    'font-size: 11px; ',
    'line-height: 1.5em',
    '">',label,
    '</div>'))


m1 <- maplibre(center = c(-62, 44),
         zoom = 3.5)  |> 
  # add filled polygons
  add_fill_layer(
    id = "warming_trend",
    source = sf_merge,
    fill_color = trend_color_scale$expression,
    fill_opacity = 0.6,
    hover_options = list(fill_opacity = 1,
                         fill_color =  trend_color_scale$expression),
    tooltip = "label2"
  ) |>
  # add solid outlines for significant trends
  add_line_layer(
    "sig_polygons",
    source = sf_merge %>% filter(p_val <= 0.05),
    line_color = "black",
    line_width = 1
  ) |>
  # add dashed outlines for nonsig trends
  add_line_layer(
    "nonsig_polygons",
    source = sf_merge %>% filter(p_val > 0.05),
    line_color = "black",
    line_dasharray = c(5,2),
    line_width = 1
  ) |>
  # add legend
  add_legend(
    "Warming Trend (°C/year)<br><small>Solid outlines indicate significance</small>",
    values = get_legend_labels(trend_color_scale, digits = 2),
    colors = get_legend_colors(trend_color_scale),
    type = "continuous",
    position = "top-right"

  ) 


m2 <- maplibre(center = c(-62, 44),
               zoom = 3.5)  |> 
  # add filled polygons
  add_fill_layer(
    id = "mean_temp",
    source = sf_merge,
    fill_color = mean_color_scale$expression,
    fill_opacity = 0.5,
    hover_options = list(fill_opacity = 1,
                         fill_color =  mean_color_scale$expression),
    tooltip = "label2"
  ) |>
  # add outlines (all same)
  add_line_layer(
    "mean_temp_polygons",
    source = sf_merge %>% filter(p_val <= 0.05),
    line_color = "black",
    line_width = 1
  ) |>
  # add legend
  add_legend(
    legend_title = "Mean Temperature (°C)<br><small>Average value across timeseries</small>",
    values = get_legend_labels(mean_color_scale, digits = 2),
    colors = get_legend_colors(mean_color_scale),
    type = "continuous",
    position = "top-left"
  ) 


# make widget showing both maps with a slider
compare(m2, m1)
  

```

<div style="height: 2rem;"></div>
<div style="height: 2rem;"></div>

### Summary Table by Region

Summary values for the AZMP Satellite Temperature within each individual NAFO division in the Eastern and Western Scotian Shelf are shown below (Table \@ref(tab:azmp-satellite-temperature-trends-table))

Table: (\#tab:azmp-satellite-temperature-trends-table) Summary values of AZMP Satellite Temperature for NAFO divisions in the Eastern Scotian Shelf (4Vn, 4Vs, 4W), and Western Scotian Shelf (4X) values through time. For warming trends, asterisks indicate significant trends over time (p < 0.05 for the year term). 

|Metric      |Eastern Scotian Shelf     |Western Scotian Shelf  |
|:-----------|:------------------|:------------------|
| Timeseries Mean (±SE) |`r ess_means`      |`r wss_means`      |
| Warming Trend (`r timespan`) |`r ess_trends`  |`r wss_trends`     |
| Timeseries Record High |`r ess_max_year`  |`r wss_max_year`     |
| Timeseries Record Low |`r ess_min_year`  |`r wss_min_year`     |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Changes to sea surface temperature could have great impact on fisheries and stocks from both direct and indirect effects. 

Sea surface temperatures can directly impact fisheries via changes to the distribution and phenology of key fisheries species [@hutchings2012climate]. These changes are likely to result in the loss or decline of some commercially important species, but also the gain or increase of others. Atlantic Canada is projected to experience decreases in species and maximum catch in the Scotian Shelf and surrounding bioregions, but projected gains towards the Arctic [@cheung2010large]. These changes are consistent with expected poleward shifts if species track climatic envelopes towards the poles. Changes to surface temperature might also affect phenological processes of key species, including spawning times and predator/prey encounters.

Sea surface temperatures can also affect the physical and chemical properties of marine ecosystems, in turn, affecting species. Warming and freshening of surface waters can lead to increased stratification in coastal oceans, influencing primary productivity and nutrient cycling.



<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:azmp-satellite-temperature-variable-definitions) Column names and definitions in the AZMP_satellite_temperature dataset.

|variable   |description |unit |
|:----------|:-----------|:----|
|year       |Year of data collection         |  |
|region     |NAFO region or subregion         |  |
|mean_value |Regionally summarized sea surface temperature, derived from satellite data         |°C  |


<!----------- Additional Data Section -------------->
## Additional Data

AZMP Satellite Temperature also contains data for `r glue::glue_collapse(unique(azmp_satellite_temperature@data$region)[!unique(azmp_satellite_temperature@data$region) %in% merge$region], sep = ", ", last = " and ")`, which are not displayed on this page. 


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-azmp_satellite_temperature, include=T, echo=T, eval=F}
library(marea)
data('azmp_satellite_temperature')
plot(azmp_satellite_temperature)
```


```{r}
# remove all created objects
rm(list = ls())
```
