
<!-- Indicator Chapter for azmp_bottom_temperature: Bottom Temperature -->


```{r setup-azmp_bottom_temperature, include=FALSE}
library(marea)
library(dplyr)
library(ggplot2)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "azmp_bottom_temperature"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea')) 
data@data <- data@data %>% inner_join(global_cols2)

```



<!------- Header Section -------------->
<!-- Title -->
# Bottom Temperature (AZMP) {#azmp-bottom-temperature}

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Scotian Shelf (4X, 4V, 4W)

**Duration:** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

<!--**Contact:** ADD CONTACT--> 



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

AZMP Bottom Temperature data represent derived and regionally summarized data from the Atlantic Zone Monitoring Program (AZMP), accessed via the `azmpdata` R package. In `azmpdata`, raw temperature observations from fixed AZMP monitoring stations are quality controlled and aggregated into regional aggregations, generating annual timeseries of variables for as long as records exist. 

AZMP Bottom Temperature is derived from CTD profiles as the deepest reliable temperature measurement at each fixed monitoring station, and area-level values are calculated as the mean of station-level bottom temperatures within predefined spatial aggregations on an annual scale.


<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-azmp-bottom-temperature, fig.cap=paste0("AZMP Regional Bottom Temperature in Scotian Shelf Regions; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}
library(plotly)

plotly_df <- data@data %>%
  left_join(global_cols2) # join in colors and region_group df

p <- plot_ly()

regions_order <- c("4Vn","4Vs","4W","4X")

for(r in regions_order){

  df_sub <- plotly_df %>% filter(region == r)

  group_name <- unique(df_sub$region_group)

  p <- p %>%
    add_lines(
      data = df_sub,
      x = ~year,
      y = ~mean_value,
      name = r,
      legendgroup = group_name,
      legendgrouptitle = list(text =
        ifelse(r == "4Vn" & group_name == "ESS",
               "Eastern Scotian Shelf Zones",
        ifelse(r == "4X",
               "Western Scotian Shelf Zones",
               NA))
      ),
      line = list(color = unique(df_sub$color), width = 2),
      text = ~paste0("<b>",region,"</b>: ",round(mean_value,2),"°C"),
      hoverinfo = "x+text"
    )
}

p %>%
  layout(
    title = "AZMP Bottom Temperature by Region",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Sea Bottom Temperature (°C)", fixedrange = TRUE),
    hovermode = "x unified",
    legend = list(
      tracegroupgap = 5,
      groupclick = "toggleitem"
    )
  ) %>%
  config(displayModeBar = FALSE)

```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-azmp_bottom_temperature, include=F}

 # get value colname
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]


# calculate mean per region
region_means <- data@data %>%
      group_by(region) %>%
      summarize(mean = mean(get(value_col)),
                sd = sd(get(value_col)),
                max_year = year[which.max(get(value_col))],
                max_val = max(get(value_col)),
                min_year = year[which.min(get(value_col))],
                min_val = min(get(value_col)),
                .groups = "drop") %>%
      mutate(region = forcats::fct_reorder(region, mean))


# find trends per region
data_split <-  data@data  %>% split(f =  data@data$region)
data_split_mods <- purrr::map(.x = data_split,
                                  .f = ~lm(.x, formula = get(value_col) ~ year))
# find coefficients per region
data_coefficients <- purrr::map(.x = data_split_mods,
                                    .f = ~coefficients(.x)[[2]] %>% as_tibble) %>%
      bind_rows(.id = "region") %>%
      rename(trend = value)
# find confidence intervals per region
data_confints <- purrr::map(.x = data_split_mods,
                            .f = ~confint(.x)[2,] %>%
                              setNames(c("lower","upper"))) %>%
  bind_rows(.id = "region")
# find p values per region
data_p_values <- purrr::map(.x = data_split_mods,
                                .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
      bind_rows(.id = "region") %>%
      rename(p_val = value)


# merge
merge <- region_means %>% 
  merge(data_coefficients) %>%
  merge(data_confints) %>%
  merge(data_p_values) %>% 
  arrange(region) %>%
  merge(global_cols2)


# plot
slopes_p2 <- merge %>%
  ggplot(aes(y = region, x = trend,
              xmin = lower, xmax = upper,
              color = mean)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(linewidth = 1) +
  geom_text(data = . %>% filter(p_val < 0.05),
            label = "*", 
            nudge_y =.15,
            size = 6, color = "black") +
  scale_color_viridis_c() +
  ggthemes::theme_few(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title.position = "plot") +
  labs(y = "NAFO Divison",
       x = "Temperature Trend (°C/year)",
       color = "Mean\nTemperature\n(°C)",
       title = "Bottom Temperature Trends per Region",
       subtitle = "Asterisks indicate significant trends (p < 0.05)") +
    facet_wrap(region_group_label~.,
               ncol = 1,
               scales = "free_y",
               space = "free_y")  


slopes_p2

# find number of regions increasing and decreasing
n_regions <- length(unique(data@data$region))
n_increasing <- sum(data_coefficients$trend > 0)
n_decreasing <- sum(data_coefficients$trend < 0)
n_significant <- sum(merge$p_val < 0.05)
majority_trend <- ifelse(n_increasing > n_decreasing, "increasing","decreasing")

# find highest and lowest regions
highest_region <- as.character(merge$region[which.max(merge$mean)])
highest_mean <- round(merge$mean[merge$region == highest_region],2)
lowest_region <- as.character(merge$region[which.min(merge$mean)])
lowest_mean <- round(merge$mean[merge$region == lowest_region],2)
highest_trend_region <-  as.character(merge$region[which.max(merge$trend)])
lowest_trend_region <-  as.character(merge$region[which.min(merge$trend)])



# find mean quantile of most recent value
last_year <- max(data@data$year)
data_split_last <- purrr::map(
  .x = data_split,
  .f = ~last(.x)
)
last_percentile <- purrr::map2(
  .x = data_split,
  .y = data_split_last,
  .f = ~round(ecdf(.x[[value_col]])(.y[[value_col]]) * 100,2) %>%
    as_tibble
) %>%
  bind_rows(.id = "region") %>%
  rename(percentile = value)
last_percentile_groups <- last_percentile %>%
  summarize(n_high = sum(percentile > 70 ),
            n_mid = sum(percentile > 30 & percentile < 70),
            n_low = sum(percentile < 30))


last10 <- data@data %>%
  group_by(region) %>%
  summarize(highest_year = year[mean_value == max(mean_value)],
            max_year = max(year)) %>%
  ungroup() %>%
  summarize(last10 = sum(highest_year >= max_year-10))

ess_means <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_means <- merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_trends <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(trend,3),"°C/year ",ifelse(p_val < 0.05,"\\*",""))) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_trends <- merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(trend,3),"°C/year ",ifelse(p_val < 0.05,"\\*",""))) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_max_year <-  merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", max_year," (",round(max_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_max_year <-  merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", max_year," (",round(max_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_min_year <-  merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", min_year," (",round(min_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_min_year <-  merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", min_year," (",round(min_val,2), "°C) ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")


```



Mean AZMP Bottom Temperature varies between regions, with **NAFO Division `r highest_region` having the warmest mean temperature (`r highest_mean`°C), and NAFO Division `r lowest_region` having the lowest mean temperature (`r lowest_mean`°C)**. `r last10` regions had their series records within the last 10 years of sampling (`r last_year-10`-`r last_year`) (Fig. \@ref(fig:plot-azmp-bottom-temperature)).

In the most recent year of sampling (`r last_year`), **`r last_percentile_groups[["n_high"]]` regions had values that are high** compared to their overall timeseries (>70th percentile), **`r last_percentile_groups[["n_mid"]]` regions had values that were moderate** (between 30th and 70th percentile), and **`r last_percentile_groups[["n_low"]]` regions had low values** (<30th percentile) within their timeseries.


```{r plot-trends-azmp-bottom-temperature, fig.cap=paste0("Model coefficients of bottom temperature over time for NAFO zones in the Eastern and Western Scnotian Shelf Regions; ", marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan), ". Error bars represent 95% confidence intervals of model fits. Asterisks indicate significant trends (p < 0.05)."), echo=F}
slopes_p2
```



The overall trend is **`r majority_trend`** across regions. `r n_increasing` regional subdivisions have increasing Bottom Temperature trends (all of which are statistically significant) and `r n_decreasing` have negative trends (Fig \@ref(fig:plot-trends-azmp-bottom-temperature)).

Average bottom temperature exhibited a general northward decrease between NAFO zones (Fig \@ref(fig:plot-trends-azmp-bottom-temperature-map)), but warming trends were fairly consistent across zones (Fig. \@ref(fig:plot-trends-azmp-bottom-temperature),\@ref(fig:plot-trends-azmp-bottom-temperature-map)).

```{r plot-trends-azmp-bottom-temperature-map,fig.cap="Map of target study areas colored by mean temperature (left) and warming trend over time (right).", echo=F}
library(sf)
library(mapgl)

# get shapefiles for regions
sf <- readRDS("data/derived_data/azmp_area.rds") %>%
  filter(get(indicator) == T)

sf_merge <- sf %>%
  select(area, area_type, geometry) %>%
  rename(region = area) %>%
  # merge in modeled values
  inner_join(merge) 

# make color scale for trends - symmetrical around zero
maxtrend <- max(abs(sf_merge$trend))
scale_data <- data.frame(trend = c(sf_merge$trend, -maxtrend, maxtrend))
trend_color_scale <- interpolate_palette(
  data = scale_data,
  column = "trend",
  method = "equal",
  n = 5, 
  palette = function(n) rev(RColorBrewer::brewer.pal(n, "RdBu"))  
)


# make color scale for means
mean_color_scale <- interpolate_palette(
  data = sf_merge,
  column = "mean",   
  method = "equal",
  n = 4,
  palette = viridisLite::viridis
)

# make popup label
sf_merge <- sf_merge %>%
  mutate(label = paste(
      "<span style='font-size: 12px;'><b>",region, "</b><br>",
      "<b>", region_group_label, "</b><br>",
      "<b>Timeseries Mean:</b>", round(mean,2),"°C<br>",
      "<b>Temperature Trend:</b>", round(trend,3),"°C/year<br>",
      "<b>p_val:</b>", ifelse(p_val<0.05,"<0.05", round(p_val,2)),
      "<br>"
    )) %>%
    mutate(label2 = paste0(
    '<div style="',
    'font-size: 11px; ',
    'line-height: 1.5em',
    '">',label,
    '</div>'))


m1 <- maplibre(center = c(-62, 44),
         zoom = 4.5)  |> 
  # add filled polygons
  add_fill_layer(
    id = "warming_trend",
    source = sf_merge,
    fill_color = trend_color_scale$expression,
    fill_opacity = 0.6,
    hover_options = list(fill_opacity = 1,
                         fill_color =  trend_color_scale$expression),
    tooltip = "label2"
  ) |>
  # add solid outlines for significant trends
  add_line_layer(
    "sig_polygons",
    source = sf_merge %>% filter(p_val <= 0.05),
    line_color = "black",
    line_width = 1
  ) |>
  # add dashed outlines for nonsig trends
  add_line_layer(
    "nonsig_polygons",
    source = sf_merge %>% filter(p_val > 0.05),
    line_color = "black",
    line_dasharray = c(5,2),
    line_width = 1
  ) |>
  # add legend
  add_legend(
    "Warming Trend (°C/year)<br><small>Solid outlines indicate significance</small>",
    values = get_legend_labels(trend_color_scale, digits = 2),
    colors = get_legend_colors(trend_color_scale),
    type = "continuous",
    position = "top-right"
  ) 


m2 <- maplibre(center = c(-62, 44),
               zoom = 4.5)  |> 
  # add filled polygons
  add_fill_layer(
    id = "mean_temp",
    source = sf_merge,
    fill_color = mean_color_scale$expression,
    fill_opacity = 0.5,
    hover_options = list(fill_opacity = 1,
                         fill_color =  mean_color_scale$expression),
    tooltip = "label2"
  ) |>
  # add outlines (all same)
  add_line_layer(
    "mean_temp_polygons",
    source = sf_merge %>% filter(p_val <= 0.05),
    line_color = "black",
    line_width = 1
  ) |>
  # add legend
  add_legend(
    legend_title = "Mean Temperature (°C)<br><small>Average value across timeseries</small>",
    values = get_legend_labels(mean_color_scale, digits = 2),
    colors = get_legend_colors(mean_color_scale),
    type = "continuous",
    position = "top-left"
  ) 


# make widget showing both maps with a slider
compare(m2, m1)
  

```

<div style="height: 2rem;"></div>

### Summary Table by Region

Summary values for the AZMP Bottom Temperature within each individual NAFO zone in the Eastern and Western Scotian Shelf are shown below (Table \@ref(tab:azmp-bottom-temperature-trends-table))

Table: (\#tab:azmp-bottom-temperature-trends-table) Summary values of AZMP Bottom Temperature for NAFO zones in the Eastern Scotian Shelf (4Vn, 4Vs, 4W), and Western Scotian Shelf (4X) values through time. For warming trends, asterisks indicate significant trends over time.

|Metric      |Eastern Scotian Shelf     |Western Scotian Shelf  |
|:-----------|:------------------|:------------------|
| Mean Value |`r ess_means`      |`r wss_means`      |
| Warming Trend (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) |`r ess_trends`  |`r wss_trends`     |
| Hottest Year |`r ess_max_year`  |`r wss_max_year`     |
| Coldest Year |`r ess_min_year`  |`r wss_min_year`     |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Sea bottom temperature is highly relevant to population dynamics and spatial distributions, particularly for benthic or bottom-associated species. The effects of sea bottom temperature on fisheries is expected to be variable between regions and species, since thermal envelopes vary across species and shifts in temperature can increase or decrease preferred thermal habitat across regions. 

In the Scotian Shelf-Bay of Fundy region, warming bottom temperatures have been associated with negative impacts and/or northward distribution shifts to key decapod species, northern shrimp and snow crab, for which this region represents the southernmost extent of both fisheries [@zisserson2017impact; @chang2025temperature]. Alternatively, recent increases in bottom temperature across the Northwestern Atlantic have increased suitable thermal habitat for halibut in most Eastern Canadian NAFO zones, and models project further increases in abundance within the Canadian range as bottom temperatures continue to increase [@shackell2022spatial; @czich2023recent].



<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:azmp-bottom-temperature-variable-definitions) Column names and definitions in the AZMP_bottom_temperature dataset.


|variable   |description |unit |
|:----------|:-----------|:----|
|year       |Year of data collection         |   |
|region     |Region over which records are summarized         |   |
|mean_value |Regionally-aggregated annual mean temperature         |°C  |


<!----------- Additional Data Section -------------->
## Additional Data

AZMP Bottom Temperature also contains data for `r glue::glue_collapse(unique(azmp_bottom_temperature@data$region)[!unique(azmp_bottom_temperature@data$region) %in% merge$region], sep = ", ", last = " and ")`, which are not displayed on this page. 


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-azmp_bottom_temperature, include=T, echo=T, eval=F}
library(marea)
data('azmp_bottom_temperature')
plot(azmp_bottom_temperature)
```


```{r}
# remove all created objects
rm(list = ls())
```
