
<!-- Indicator Chapter for azmp_stratification: Stratification Index -->


```{r setup-azmp_stratification, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "azmp_stratification"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Stratification Index (AZMP)

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Scotian Shelf (4X, 4V, 4W)

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Stratification Index describes the vertical density structure of the water column, typically driven by differences in temperature and salinity between surface and deeper waters. AZMP calculates stratification index as the difference in density between water at 0 and 50m depth, where density is derived from CTD profiles of temperature and salinity [@harrison2004oceanographic]. 

A stratification index of 0 indicates no difference in density between surface and deeper waters, and implies strong vertical mixing in the water column. Higher stratification values indicate larger differences between surface and deep water, suggesting that surface water is warmer and/or fresher than deeper water. 



<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-azmp-stratification, fig.cap=paste0("AZMP Stratification for Scotian Shelf; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}

plotly_df <- data@data %>%
  mutate(smooth_10yr = zoo::rollapply(mean_value, mean, width = 10, partial = TRUE),
         overall_mean = mean(mean_value))

p <- plot_ly(plotly_df, x = ~year) %>%
  add_lines(y = ~mean_value,
            name = "Yearly Anomaly",
            line = list(color = "lightgrey"),
            hovertemplate = "Monthly anomaly: %{y:.2f}<extra></extra>"
           ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "red", width = 2),
            hovertemplate = "10-yr smoothed: %{y:.2f}<extra></extra>",
           ) %>%
    add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Overall Mean: %{y:.2f}<extra></extra>",
           ) %>%
  layout(
    title = "Stratification Index for Scotian Shelf (4X, 4V, 4W)<br><sup>Source: DFO Atlantic Zone Monitoring Program via azmpdata</sup>",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Stratification Index (kg/m<sup>3</sup>)",
                 fixedrange = TRUE),
    hovermode = "x unified",
    margin = list( t = 80)
    
  ) %>%
  config(displayModeBar = FALSE) 

p

```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-azmp_stratification, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
current_vs_mean <- last_val - overall_mean
current_vs_mean_text <- ifelse(sign(current_vs_mean) == 1,"above","below")
last_different_sign <- switch(current_vs_mean_text,
                                  "above" = last(which(data@data[value_col] < overall_mean)),
                                  "below" = last(which(data@data[value_col] > overall_mean)))

streak_start <- last_different_sign + 1
streak_start_date <- ifelse(has_months,
                                       paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
                                       paste0(data@data[streak_start,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the Stratification Index is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The Stratification Index is currently **`r current_vs_mean_text` the long-term mean**, and has been **since `r streak_start_date`**.
Stratification in the Scotian Shelf region has been **`r current_trend` trend** in recent years, and broadly throughout the time series. 



<!--------- Trends Table--------->
### Summary Table

Summary values for the AZMP Stratification in the Scotian Shelf region are shown below (Table \@ref(tab:azmp-stratification-trends-table))

Table: (\#tab:azmp-stratification-trends-table) Summary values of AZMP Stratification in the Scotian Shelf region through time. 

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. The most recent value is `r abs(round(current_vs_mean,2))` `r current_vs_mean_text` the long-term mean.|
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,5)` per year | In the last ten years, azmp_stratification has **`r last_ten_direction`** by `r round(last_ten_slope,5)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,5)` per year | Since the beginning of the timeseries, azmp_stratification has **`r full_mod_direction`** by `r round(full_mod_slope,5)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Highest Value | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Lowest Value | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Stratification can affect fisheries and marine ecosystems through bottom-up influence on primary productivity. Stratification can influence phytoplankton in the water column in complex ways. Stratified conditions can benefit phytoplankton by retaining them in the light-rich upper layer the water column.  Alternatively, stratified water columns experience weaker vertical mixing, potentially reducing nutrient influx from deeper waters which can limit algal growth [@araujo2012effects]. Expected effects of stratification change can therefore depend on the context of the target environment. 

In the Scotian Shelf region, stratification index has shown a negative relationship with primary productivity [@araujo2012effects], and influence on the timing of season algae blooms [@song2011interannual]. Stratification can also affect water retention, thus potentially planktonic dispersal and larval transport.


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:azmp-stratification-variable-definitions) Column names and definitions in the AZMP_stratification_temperature dataset.


|variable   |description |unit |
|:----------|:-----------|:----|
|year       |Year of data collection         |   |
|region     |Region over which samples are summarized (only scotian_shelf_box)          |   |
|mean_value |Difference in water density between 0 and 50m depth         | kg/m<sup>3</sup>  |

<!----------- Additional Data Section -------------->
## Additional Data

No additional data for AZMP Stratification. 


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-azmp_stratification, include=T, echo=T, eval=F}
library(marea)
data('azmp_stratification')
plot(azmp_stratification)
```


```{r}
# remove all created objects
rm(list = ls())
```
