
<!-- Indicator Chapter for azmp_surface_temperature: Surface Temperature -->


```{r setup-azmp_surface_temperature, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "azmp_surface_temperature"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Surface Temperature (AZMP)

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Scotian Shelf (4X, 4V, 4W)

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

AZMP Surface Temperature contains empirical temperature measurements at 0m depth, averaged across NAFO zones (4V, 4X, 4W) and regional boundaries (Bay of Fundy + Gulf of Maine). Data from this indicator is not subdivided regionally, but see Chapter \@ref(azmp-satellite-temperature) for spatially-stratified surface temperature data. 

Sea surface temperature is influenced by a variety of conditions spanning multiple spatial and temporal scales. In this region, long-term trends in SST are consistent with expectations of anthropogenic climate change [@greenan2019changes]. At the same time, decadal-to-interannual regional SST trends are influenced by atmospheric indicators like those described in this book, and short-term local values can be influenced by many oceanographic and climate factors, such as freshwater input, ocean currents, and weather events.




<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-azmp-surface-temperature, fig.cap=paste0("AZMP Surface Temperature in Scotian Shelf; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}
library(plotly)
plotly_df <- data@data %>%
  mutate( 
    smooth_10yr = zoo::rollapply(mean_value, mean, width = 10, partial = TRUE),
    overall_mean = mean(mean_value))

p <- plot_ly(plotly_df, x = ~year) %>%
  add_lines(y = ~mean_value,
            name = "Annual Mean",
            line = list(color = "grey"),
            hovertemplate = "Monthly anomaly: %{y:.2f}<extra></extra>"
           ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "blue", width = 2),
            hovertemplate = "10-yr smoothed: %{y:.2f}<extra></extra>",
            ) %>%
    add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Overall Mean: %{y:.2f}<extra></extra>",
           ) %>%
  layout(
    title = "AZMP Surface Temperature for Scotian Shelf (4X, 4V, 4W)<br><sup>Source: DFO Atlantic Zone Monitoring Program via azmpdata</sup>",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Surface Temperature (°C)",
                 fixedrange = TRUE),
    hovermode = "x unified",
    margin = list( t = 80)
    
  ) %>%
  config(displayModeBar = FALSE) 

p


```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-azmp_surface_temperature, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
current_vs_mean <- last_val - overall_mean
current_vs_mean_text <- ifelse(sign(current_vs_mean) == 1,"above","below")
last_different_sign <- switch(current_vs_mean_text,
                                  "above" = last(which(data@data[value_col] < overall_mean)),
                                  "below" = last(which(data@data[value_col] > overall_mean))) + 1
last_date_different_sign <- ifelse(has_months,
                                       paste0(data@data[last_different_sign,"month"],"/",data@data[last_different_sign,"year"]),
                                       paste0(data@data[last_different_sign,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the surface temperature value is `r last_val` °C, which is **`r current_magnitude`** among values in the timeseries.
Surface temperature is currently **`r current_vs_mean_text` the long-term mean**, and has been **since `r last_date_different_sign`**.
Surface temperature has followed a(n) **`r current_trend` trend** in recent years (Fig. \@ref(fig:plot-azmp-surface-temperature)).



<!--------- Trends Table--------->
### Summary Table

Summary values for the AZMP Surface Temperature in the Scotian Shelf region are shown below (Table \@ref(tab:azmp-surface-temperature-trends-table))

Table: (\#tab:azmp-surface-temperature-trends-table) Summary values of AZMP Stratification in the Scotian Shelf region through time. 

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. The most recent value is `r abs(round(current_vs_mean,2))` `r current_vs_mean_text` the long-term mean.|
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, azmp_surface_temperature has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, azmp_surface_temperature has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Highest Value | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Lowest Value | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Changes to sea surface temperature could have great impact on fisheries and stocks from both direct and indirect effects. 

Sea surface temperatures can directly impact fisheries via changes to the distribution and phenology of key fisheries species [@hutchings2012climate]. These changes are likely to result in the loss or decline of some commercially important species, but also the gain or increase of others. Atlantic Canada is projected to experience decreases in species and maximum catch in the Scotian Shelf and surrounding bioregions, but projected gains towards the Arctic [@cheung2010large]. These changes are consistent with expected poleward shifts if species track climatic envelopes towards the poles. Changes to surface temperature might also affect phenological process of key species, including spawning times and predator/prey encounters.

Sea surface temperatures can also affect the physical and chemical properties of marine ecosystems, in turn, affecting species. Warming and freshening of surface waters can lead to increased stratification in coastal oceans, influencing primary productivity and nutrient cycling.



<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:azmp-surface-temperature-variable-definitions) Column names and definitions in the AZMP_surface_temperature dataset.

|variable   |description |unit |
|:----------|:-----------|:----|
|year       |Year of data observations         |  |
|region     |Area over which observations were summarized (only one: `scotian_shelf_box`)        |  |
|mean_value |Averaged surface temperature in a given year       | °C  |

<!----------- Additional Data Section -------------->
## Additional Data

No additional data for AZMP Surface Temperature.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-azmp_surface_temperature, include=T, echo=T, eval=F}
library(marea)
data('azmp_surface_temperature')
plot(azmp_surface_temperature)
```


```{r}
# remove all created objects
rm(list = ls())
```
