
<!-- Indicator Chapter for eco_indicators: Fishing Pressure -->


```{r setup-eco_indicators_fp, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("FP"))
data@data %>% colnames()
```



<!------- Header Section -------------->
<!-- Title -->
# Fishing Pressure

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Fishing pressure is a relative metric of intensity of exploitation on fished groups. Here, fishing pressure is presented per guild and calculated as a ration: fishing pressure = total landings / total biomass [@Bundy2017_ecological_indicators].

Because fishing pressure in a metric, changes in fishing pressure can either result from changes to total landings, changes to total biomass, or both. Conversely, stability in fishing pressure does not guarantee that total landings and total biomass are also stable; if each value changes in the same direction, fishing pressure might remain stable. 



<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-fp, fig.cap="Fishing pressure in all NAFO regions and Scotian Shelf regions over time.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_biomass_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # assign regions to consistent colors
 # region_colors <- setNames(hcl.colors(length(regions), palette = "Dark 3"), regions)

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("fp_", "", x)) 

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, linetype, linewidth, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text =
            ifelse(region_i == "4VN" & group_name == "ESS",
                   "Eastern Scotian Shelf Zones",
                   ifelse(region_i == "4X",
                          "Western Scotian Shelf Zones",
                          NA))),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Fishing Pressure")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = "Fishing Pressure", fixedrange = TRUE),
      legend = list(
        title = list(text = str_to_title(region_col)),
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem"
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = 0, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80),
            # add horizontal line at 1
      shapes = list(
        type = "line",
        xref = "paper", 
        yref = "y",     
        x0 = 0,        
        x1 = 1,        
        y0 = 1,         
        y1 = 1,         
        line = list(color = "black", width = 2, dash = "dash") 
      )
    )
}

# usage:
p <- make_biomass_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
# use html to pause scrolling on main page while scrolling in dropdown tab
p


```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-eco-indicators-fp}

fp_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
fp_long_split <- fp_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
fp_means <- purrr::map(
  .x = fp_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
fp_mods <- purrr::map(
  .x = fp_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
fp_mods_slopes <- purrr::map(
  .x = fp_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
fp_mods_confint <- purrr::map(
  .x = fp_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
fp_mods_p <-  purrr::map(
  .x = fp_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = fp_mods,
  .y = fp_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- fp_means %>%
  left_join(fp_mods_slopes) %>%
  left_join(fp_mods_confint) %>%
  left_join(fp_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("fp_","") %>%
           str_replace_all("_"," ") %>%
           str_trim() %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_hline(yintercept = 0) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Fishing Pressure (Landings/Biomass)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, fp = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))



# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")

strongest_decrease <- merge %>% filter(trend == min(trend)) %>%
  select(region, label)

strongest_increase <- merge %>% filter(trend == max(trend)) %>%
  select(region, label)

# get just the trend for "all"
trend_all_ess <- merge[merge$region == "ESS" & merge$label == "All",]
trend_all_ess <- paste(ifelse(trend_all_ess$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_ess$trend > 0,"increased","decreased"))

trend_all_wss <- merge[merge$region == "WSS" & merge$label == "All",]
trend_all_wss <- paste(ifelse(trend_all_wss$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_wss$trend > 0,"increased","decreased"))


```

Estimates of fishing pressure are highly variable across time and between regions, and often exceed theoretical bounds (upper limit of 1), likely due to inaccurate estimates of total biomass (Fig. \@ref(fig:plot-eco-indicators-fp)). Still, interpreting trends can reveal some insights into changes in fishing pressure to Scotian Shelf communities.

When fitting fishing pressure to linear models over time across all taxa ("All" group), fishing pressure **`r trend_all_ess` in the Eastern Scotian Shelf** and **`r trend_all_wss` in the Western Scotian Shelf**.

Within specific guilds, fishing pressure increased in `r n_increasing_ess` groups in the Eastern Scotian Shelf (`r n_increasing_sig_ess` significant) and decreased in `r n_decreasing_ess` groups (`r n_decreasing_sig_ess` significant). In the Western Scotian Shelf, fishing pressure increased in `r n_increasing_wss` groups (`r n_increasing_sig_wss` significant) and decreased in `r n_decreasing_wss` groups (`r n_decreasing_sig_wss` significant) (Fig. \@ref(fig:plot-trends-eco-indicators-fp)).


The strongest increase in fishing pressure was on the `r strongest_increase$label` group in the `r strongest_increase$region` region, and the strongest decrease was on the `r strongest_decrease$label` group in the `r strongest_decrease$region` region (Fig. \@ref(fig:plot-trends-eco-indicators-fp)). 


```{r plot-trends-eco-indicators-fp, echo=F, fig.cap = "Fishing Pressure for guilds over time in Scotian Shelf."}
p_slopes
```


<!--------- Trends Table--------->


### Summary Table by Region and Guild

Trends of Fishing Pressure over time for each guild and NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown in the table below (Table \@ref(tab:make-table-per-zone-fp)). 


```{r make-table-per-zone-fp}

fp_long2 <- data@data %>%
  inner_join(global_cols2) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "taxon",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and taxon
fp_long_split2 <- fp_long2 %>% split(f = interaction(.$region,.$taxon))
# remove empty list items
sums <- purrr::map(.x = fp_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
fp_long_split2 <- fp_long_split2[sums != 0]

fp_mods_all <- purrr::map(.x = fp_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
fp_mods_all_sums <- purrr::map(.x = fp_mods_all,
                                    .f = ~summary(.x))
fp_mods_all_coefs <- purrr::map(.x = fp_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
fp_mods_all_p_val <- purrr::map(.x = fp_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

fp_mods_all_coefs_labels <- fp_mods_all_coefs %>%
  left_join(fp_mods_all_p_val) %>%
  mutate(region = str_extract(id, "4.*?(?=\\.)"),
         taxon = str_replace(id, paste0(region,".FP_"),""),
         taxon = str_replace(taxon,"\\._value",""),
         taxon = str_replace(taxon,"_value",""),
         taxon = str_replace(taxon,"_"," "),
         taxon = str_to_title(taxon))  %>%
  relocate(region, taxon)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

fp_mods_all_coefs_labels_collapse <- fp_mods_all_coefs_labels %>%
  left_join(global_cols2) %>%
  group_by(region_group_label, taxon) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
    rename(`Eastern Scotian Shelf` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(fp_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of fishing pressure in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Fishing pressure affects populations both by direct removal of individuals from the population, and by alteration of the size-structure and demographic properties of populations. Fishing pressure generally targets larger, older fish, thus truncating the age of the remaining population by removing mature and fecund individuals [@tu2018fishing]. High fishing pressure can contribute to the decline or collapse of exploited fisheries [@pinsky2015fishing; @essington2015fishing]. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-----------|:----|
|year                           |Year of data collection         |  |
|region                         |Area over which observations are summarized         |  |
|FP_{TAXON}_value       |Fishing Pressure, calculated as Landings/Biomass         | Proportion  |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions. Invertebrates:Groundfish ratios are not available at the Scotian Shelf region scale (ESS, WSS).

<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_fp, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
