
<!-- Indicator Chapter for eco_indicators: biomass of Fished groups -->


```{r setup-eco_indicators_biomassComm, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("biomass"))
data@data %>% colnames()
data@data <- data@data %>%
  select(year,
         region,
         BIOMASS_ALL_value,
         BIOMASS_FINFISH_value,
         BIOMASS_INVERTEBRATES_value
         )

timespan <- marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)

```



<!------- Header Section -------------->
<!-- Title -->
# Resource Potential: Community

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r timespan`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->


<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Resource potential describes the total production capacity of the community, which can change as the community is exploited. Resource potential of the overall community serves as an indicator of productivity in the ecosystem. [@Bundy2017_ecological_indicators]. 

Resource potential of community is quantified by the biomass of grouped species:

* All
* Finfish
* Invertebrates


<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-biomassComm,  fig.cap=paste0("Biomass of the Community in Scotian Shelf regions; ",timespan, ". Use dropdown box to select a target taxonomic or functional group, and click legend to isolate regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols2)

# function to create plot with dropdown menu ------------------------------
make_biomassComm_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) %>%  gsub("Lb","Large B",.) %>% gsub("Mb","Medium B",.)

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]

    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
        group_name <- unique(dat$region_group)
        color <- unique(dat$color)


      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }

      p <- p %>% add_bars(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        marker = list(color = color),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:,.2s}<extra></extra>")      )
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Biomass (tonnes)")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col), type = "category"),  # keep one bar per year
      yaxis = list(title = "Biomass (tonnes)", fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = 0, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_biomassComm_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p

```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco-indicators-biomassComm}

biomassComm_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
biomassComm_long_split <- biomassComm_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
biomassComm_means <- purrr::map(
  .x = biomassComm_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
biomassComm_mods <- purrr::map(
  .x = biomassComm_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
biomassComm_mods_slopes <- purrr::map(
  .x = biomassComm_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
biomassComm_mods_confint <- purrr::map(
  .x = biomassComm_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
biomassComm_mods_p <-  purrr::map(
  .x = biomassComm_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = biomassComm_mods,
  .y = biomassComm_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- biomassComm_means %>%
  left_join(biomassComm_mods_slopes) %>%
  left_join(biomassComm_mods_confint) %>%
  left_join(biomassComm_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("biomass_","") %>%
           str_replace_all("_"," ") %>%
           str_replace("lb","lg b") %>%
           str_replace("mb", "med b") %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Biomass (tonnes)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))


ess_all_label <- paste0(
  ifelse(merge %>% filter(region == "ESS", label == "All") %>% pull(trend) > 0, "increased","decreased"),
  " ",
    ifelse(merge %>% filter(region == "ESS", label == "All") %>% pull(p_val) < 0.05, "significantly","nonsignificantly"))

ess_finfish_label <- paste0(
  ifelse(merge %>% filter(region == "ESS", label == "Finfish") %>% pull(trend) > 0, "increased","decreased"),
  " ",
    ifelse(merge %>% filter(region == "ESS", label == "Finfish") %>% pull(p_val) < 0.05, "significantly","nonsignificantly"))

wss_all_label <- paste0(
  ifelse(merge %>% filter(region == "WSS", label == "All") %>% pull(trend) > 0, "increased","decreased"),
  " ",
    ifelse(merge %>% filter(region == "WSS", label == "All") %>% pull(p_val) < 0.05, "significantly","nonsignificantly"))


wss_finfish_label <- paste0(
  ifelse(merge %>% filter(region == "WSS", label == "Finfish") %>% pull(trend) > 0, "increased","decreased"),
  " ",
    ifelse(merge %>% filter(region == "WSS", label == "Finfish") %>% pull(p_val) < 0.05, "significantly","nonsignificantly"))


```


Resource potential of the community shows different trends on the Scotian Shelf between regions (Fig. \@ref(fig:plot-eco-indicators-biomassComm)). 

In the Eastern Scotian Shelf, biomass of all species `r ess_all_label`, and biomass of finfish `r ess_finfish_label`. In the Western Scotian Shelf, biomass of all species `r wss_all_label` and biomass of finfish `r wss_all_label` (Fig. \@ref(fig:plot-trends-eco-indicators-biomassComm)). Invertebrate data were not available at the "ESS" and "WSS" region scale. 

Because biomass of finfish is much higher than biomass of invertebrates across regions (Fig. \@ref(fig:plot-eco-indicators-biomassComm)), trends in "all" are likely driven primarily by trends in finfish  (Fig. \@ref(fig:plot-trends-eco-indicators-biomassComm)).

```{r plot-trends-eco-indicators-biomassComm, fig.cap="Overall Biomass trends for taxonomic and functional groups over time.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in biomass over time for each speciesgroup and each NAFO region within the Eastern and Western Scotian Shelf (`r timespan`) are shown in the table below (Table \@ref(tab:make-table-per-zone-biomassComm)). 

```{r make-table-per-zone-biomassComm}

biomassComm_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "taxon",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and taxon
biomassComm_long_split2 <- biomassComm_long2 %>% split(f = interaction(.$region,.$taxon))
# remove empty list items
sums <- purrr::map(.x = biomassComm_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
biomassComm_long_split2 <- biomassComm_long_split2[sums != 0]

biomassComm_mods_all <- purrr::map(.x = biomassComm_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
biomassComm_mods_all_sums <- purrr::map(.x = biomassComm_mods_all,
                                    .f = ~summary(.x))
biomassComm_mods_all_coefs <- purrr::map(.x = biomassComm_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
biomassComm_mods_all_p_val <- purrr::map(.x = biomassComm_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

biomassComm_mods_all_coefs_labels <- biomassComm_mods_all_coefs %>%
  left_join(biomassComm_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*(?=\\.)"),
         taxon = str_replace(id, paste0(region,".BIOMASS_"),""),
         taxon = str_replace(taxon,"_value",""),
         taxon = str_to_title(taxon))  %>%
  relocate(region, taxon)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

biomassComm_mods_all_coefs_labels_collapse <- biomassComm_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, taxon) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Biomass Trends (tonnes/year)` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Biomass Trends (tonnes/year)` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(biomassComm_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = paste0("Change trends for biomass of the community in NAFO regions within the Eastern and Western Scotian Shelf regions; ",timespan,". Asterisks indicate significant trends over time (p &lt; 0.05).")) %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Changes in resource potential of the community indicate the balance between community productivity and removal by harvest. In Atlantic Canada, resource potential metrics are heterogenous between regions and taxa, and can be useful indicators of change at the local scale [@Bundy2017_ecological_indicators; @irvine2025spatial].


<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|biomass_{GROUP}_value      |Biomass of community groups within regions         | tonnes |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_biomassComm, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
