
<!-- Indicator Chapter for eco_indicators: eco-resistance -->


```{r setup-eco_indicators_eco-resistance, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))

data@data  <- data@data %>%
  select(year, region,
         MeanLifespan_value,
         MMLength_BIOMASS_value,
         MMLength_ABUNDANCE_value,
         IVILandings_value
  )


# resouce potential:
# fishing balance
# abundance all?

# fishing pressure:
# DiversityTargetSpp_ALL_value
# MeanTL.Landings_value


# unknown:
# MTI.Landings_3.25_value
# PELAGIC_GROUNDFISH_value
# eco-resistance_INVERTEBRATES_value
# INVERTEBRATES_GROUNDFISH_value
#MeanLengthBIOMASS_value,
#MeanLengthABUNDANCE_value,
```



<!------- Header Section -------------->
<!-- Title -->
# Stability & Resistance: Ecosystem Resistance

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The `eco_indicators` dataset provides indicators to describe the resistance of the marine ecosystem, or its ability to withstand disturbances, whether Anthropogenic or natural [@Bundy2017_ecological_indicators]. 

* **Mean Lifespan**: Mean of the maximum longevity of each species observed in a survey.
* **Mean Maximum Length (Abundance)**:  Mean maximum length of fish species in a survey, weighted by abundance.
* **Mean Maximum Length (Biomass)**: Mean maximum length of fish species in a survey, weighted by biomass.  
* **Intrinsic Vulnerability Index (IVI) of Landings**: Intrinsic Vulnerability is a measure of a species' vulnerability to fishing, driven by life history traits. IVI of landings is the total intrinsic vulnerability of species in a catch. 

The first three of these indicators are measures of the "potential" of fish species to resist perturbation. Species that are longer-lived and larger can be more resistant to change, and are the same species that are generally targeted by fishing activities. These proxies can indicate the changes to the resistance potential of the community as a result of fishing activities. 


Intrinsic Vulnerability Index of the catch is a measure of the average vulnerability of harvested species. Species with larger body size, greater longevity, higher age at maturity, and lower growth rates are generally more vulnerable to fishing pressure. Therefore, IVI is a measure of the average of the Intrinsic Vulnerability Index per fish species, weighted by their annual catch. IVI of landings is expected to increase with increased fishing pressure [@Bundy2017_ecological_indicators]. Note that decreases in IVI could be the result of a decrease in the catch of these species due to overexploitation or reductions in fishing pressure to avoid overexploitation.


<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-eco-resistance, fig.cap="eco-resistance Structure Indicators in all NAFO regions and Scotian Shelf regions overall. Use dropdown menu to select indicator, and click legend to toggle regions.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_resistance_dropdown_plot <- function(df,
year_col   = "year",
region_col = "region",
value_suffix = "_value") {
  
  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)
  
  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])
  
  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) %>% str_replace("Ivi","IVI") %>% str_replace("Mm","Mean Max")
  
  # build plot -----------------
  p <- plot_ly()
  
  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, linetype, linewidth, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }
  
  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions
  
  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)
    
    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE
    
    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title =  pretty_label(metrics[metric_i]))
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })
  
  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = pretty_label(metrics[1]), fixedrange = TRUE),
      legend = list(
        title = list(text = str_to_title(region_col)),
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_resistance_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)

p


```




<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco_indicators_eco-resistance}

resistance_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
resistance_long_split <- resistance_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
resistance_means <- purrr::map(
  .x = resistance_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
resistance_mods <- purrr::map(
  .x = resistance_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
resistance_mods_slopes <- purrr::map(
  .x = resistance_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
resistance_mods_confint <- purrr::map(
  .x = resistance_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
resistance_mods_p <-  purrr::map(
  .x = resistance_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = resistance_mods,
  .y = resistance_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- resistance_means %>%
  left_join(resistance_mods_slopes) %>%
  left_join(resistance_mods_confint) %>%
  left_join(resistance_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace_all("_"," ") %>%
           str_to_title() %>%
           str_replace("Ivi","IVI") %>%
           str_replace("Mm","Mean Max")

  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y =  NULL,
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("ESS" = "purple",
                               "WSS" = "orange")) +
  scale_color_manual(values = c("ESS" = "purple",
                                "WSS" = "orange")) +
  guides(color = guide_legend(override.aes = list(alpha = 1, resistance = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm")))) 


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")



```

Ecological Resistance variables have changed through time in the Eastern and Western Scotian Shelf regions over the sampling span, with greater variation in ESS regions (Fig. \@ref(fig:plot-eco-indicators-eco-resistance)).

Assessing trends over time reveals generally opposing patterns between the Eastern and Western Scotian Shelf, with the **Eastern Scotian Shelf `r general_trend_ess` in resistance** and the **Western Scotian Shelf `r general_trend_wss` in resistance** over time (Fig. \@ref(fig:plot-trends-eco-indicators-eco-resistance)). Both regions experienced significant increases in IVI of Landings over time, but ESS significantly decreased while WSS significantly increased in all other variables (Fig. \@ref(fig:plot-trends-eco-indicators-eco-resistance)).


```{r plot-trends-eco-indicators-eco-resistance, echo=F, fig.cap = "Ecosystem Resistance trends for ESS and WSS"}
p_slopes
```


### Summary Table by NAFO Region and eco-resistance Variable

Trends of each Ecological Resistance variable for each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown below (Table \@ref(tab:make-table-per-zone-eco-resistance)).

```{r make-table-per-zone-eco-resistance}

resistance_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and variable
resistance_long_split2 <- resistance_long2 %>% split(f = interaction(.$region,.$variable))
# remove empty list items
sums <- purrr::map(.x = resistance_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
resistance_long_split2 <- resistance_long_split2[sums != 0]

resistance_mods_all <- purrr::map(.x = resistance_long_split2,
                            .f = ~lm(data = .x,
                                     value ~ year))
resistance_mods_all_sums <- purrr::map(.x = resistance_mods_all,
                                 .f = ~summary(.x))
resistance_mods_all_coefs <- purrr::map(.x = resistance_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
resistance_mods_all_p_val <- purrr::map(.x = resistance_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

resistance_mods_all_coefs_labels <- resistance_mods_all_coefs %>%
  left_join(resistance_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*?(?=\\.)"),
         variable = str_replace(id, paste0(region,"."),""),
         variable = str_replace(variable,"\\_ALL_value",""),
         variable = janitor::make_clean_names(variable, allow_dupes=T),
         variable = str_replace_all(variable,"_"," "),
         variable = str_to_title(variable))  %>%
  relocate(region, variable)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

resistance_mods_all_coefs_labels_collapse <- resistance_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Trends` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(resistance_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of resistance variables in NAFO regions in the Eastern and Western Scotian Shelf. Asterisks indicate significant trends (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```


<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Resistance variables offer proximal indicators of a community's ability to withstand change, and have long-term implications for fisheries and ecosystems. 

While negative changes in mean lifespan and mean max length could be the result of fishing pressure targeting larger, long-lived species, positive changes in these variables might also reflect effective fisheries management through time. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:eco-resistance-variable-table) Definitions of eco_indicators eco-resistance structure variables.

|variable                    |description |unit |
|:---------------------------|:-----------|:----|
|year                           |Year of trawl survey         |  |
|region                         |Region over which values are summarized         |  |
|MeanLifespan_value       | Mean estimated lifespan of species in a survey  | years |
| MMLength_BIOMASS_value      |Mean maximum length of species in a survey (weighted by biomass) | cm |
| MMLength_ABUNDANCE_value      |Mean maximum length of species in a survey (weighted by abundance) | cm |
| IVILandings_value      | Mean intrinsic vulnerability index of species in a survey |  |


         

<!----------- Additional Data Section -------------->
## Additional Data

eco-resistance Structure variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO divisions are nested within Scotian Shelf regions.

<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_eco-resistance, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
