# Chapter Title

**Description**: Short description of indicator

**Data Type**: Spatial or tabular

**Spatial Scope**: [Choose one: Coastwide, Regional, Specific]

**Temporal Breadth**: Duration of variable

**Contact(s)**: Name of data contributor including email. 

**Source**:

```{r echo=FALSE, message=F}
knitr::opts_chunk$set(echo = F)
library(marea)
```


## Introduction to Indicator
Paragraph(s) on indicator history and values.


## Trends
Basic figures summarizing trends in indicator data

```{r}
# Plots of data trends for indicator
```


## Relevance to Research and Stock Assessments
Proceed as though this were a short summary of a typical discussion section in a paper.


## Get the data
code to access data from marea R package

```{r, include=T, echo=T, eval=F}
library(marea)
data("var")
plot(var)
```


## Variable Definitions
```{r}
varDefs <- tibble::tribble(
  ~variable, ~definition, ~unit,
  "var1",     "varDef1", "unit1",
  "var2",    "varDef2", "unit2",
)

knitr::kable(varDefs)

```
