
<!-- Indicator Chapter for eco_indicators: Fishing in Balance -->


```{r setup-eco_indicators_fishingInBalance, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))

data@data  <- data@data %>%
  select(year, region,
         FishinginBalance_value
  )

```



<!------- Header Section -------------->
<!-- Title -->
# Resource Potential: Fishing Strategies and System Productivity -- Fishing in Balance

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Fishing in Balance describes how "balanced" a fishery is in ecological terms, relating fisheries catches and the trophic level of catches with the transfer efficiency of the community in relation to some baseline year (here, 1968-1970) [@Bundy2017_ecological_indicators; @pauly2000ecopath]. Fishing in balance captures the impact of fishing on system productivity. 

Positive trends in Fishing in Balance indicate that a fishery is expanding, that bottom-up effects are occurring in the community, or that catches are higher than expected. Alternatively, negative fishing in balance trends suggest a deterioration of underlying food webs, that fishing pressure is impacting ecosystem function, or that a system is less productive as a result of fishing activity [@cury2005trophodynamic; @Bundy2017_ecological_indicators].


<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-fishingInBalance, fig.cap="Fishing in Balance for all NAFO regions and Scotian Shelf regions overall. Use dropdown menu to select indicator, and click legend to toggle regions.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

p <- plot_ly()

regions <- unique(plotly_df$region)

for(r in regions){

  df_sub <- plotly_df %>% filter(region == r)

  group_name <- unique(df_sub$region_group)
  
  linewidth <- unique(df_sub$linewidth)
  linetype <- unique(df_sub$linetype)

  p <- p %>%
    add_lines(
      data = df_sub,
      x = ~year,
      y = ~FishinginBalance_value,
      name = r,
      legendgroup = group_name,
      legendgrouptitle = list(text =
        ifelse(group_name == "ESS",
               "Eastern Scotian Shelf Zones",
               "Western Scotian Shelf Zones"
               )
      ),
      line = list(color = unique(df_sub$color),
                  dash = linetype,
                  width = linewidth),
      text = ~paste0("<b>",region,"</b>: ",round(FishinginBalance_value,2),"°C"),
      hoverinfo = "x+text"
    )
}

p %>%
  layout(
    title = "Fishing Strategies and System Productivity",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Fishing in Balance Index", fixedrange = TRUE),
    hovermode = "x unified",
    legend = list(
      tracegroupgap = 5,
      groupclick = "toggleitem",
      itemdoubleclick = FALSE
    )
  ) %>%
  config(displayModeBar = FALSE)


```




<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco_indicators_fishingInBalance}

fishingInBalance_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
fishingInBalance_long_split <- fishingInBalance_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
fishingInBalance_means <- purrr::map(
  .x = fishingInBalance_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
fishingInBalance_mods <- purrr::map(
  .x = fishingInBalance_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
fishingInBalance_mods_slopes <- purrr::map(
  .x = fishingInBalance_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
fishingInBalance_mods_confint <- purrr::map(
  .x = fishingInBalance_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
fishingInBalance_mods_p <-  purrr::map(
  .x = fishingInBalance_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = fishingInBalance_mods,
  .y = fishingInBalance_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- fishingInBalance_means %>%
  left_join(fishingInBalance_mods_slopes) %>%
  left_join(fishingInBalance_mods_confint) %>%
  left_join(fishingInBalance_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = "Fishing in Balance"
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_hline(yintercept = 0) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Fishing in Balance Index",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, fishingInBalance = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm")))) 


wss_trend <- paste(
  ifelse(merge$p_val[merge$region == "WSS"] < 0.05,
         "significantly",
         "non-significantly"),
  ifelse(merge$trend[merge$region == "WSS"] < 0,
         "decreased","increased")
)

ess_trend <- paste(
  ifelse(merge$p_val[merge$region == "ESS"] < 0.05,
         "significantly",
         "non-significantly"),
  ifelse(merge$trend[merge$region == "ESS"] < 0,
         "decreased","increased")
)



```


Within NAFO divisions in the Scotian Shelf, trends in Fishing in Balance were fairly consistent and showed general decreases across time (Fig. \@ref(fig:plot-eco-indicators-fishingInBalance)).

**Fishing in Balance `r wss_trend` in the Western Scotian Shelf** and **`r ess_trend` in the Eastern Scotian Shelf**, both with signficant trends (Fig. \@ref(fig:plot-trends-eco-indicators-fishingInBalance)) . 



```{r plot-trends-eco-indicators-fishingInBalance, echo=F, fig.cap = "Fishing in Balance trends for ESS and WSS"}
p_slopes
```


### Summary Table by NAFO Region

Trends of Fishing in Balance for each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown below (Table \@ref(tab:make-table-per-zone-fishingInBalance)).

```{r make-table-per-zone-fishingInBalance}

fishingInBalance_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and variable
fishingInBalance_long_split2 <- fishingInBalance_long2 %>% split(f = interaction(.$region,.$variable))
# remove empty list items
sums <- purrr::map(.x = fishingInBalance_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
fishingInBalance_long_split2 <- fishingInBalance_long_split2[sums != 0]

fishingInBalance_mods_all <- purrr::map(.x = fishingInBalance_long_split2,
                            .f = ~lm(data = .x,
                                     value ~ year))
fishingInBalance_mods_all_sums <- purrr::map(.x = fishingInBalance_mods_all,
                                 .f = ~summary(.x))
fishingInBalance_mods_all_coefs <- purrr::map(.x = fishingInBalance_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
fishingInBalance_mods_all_p_val <- purrr::map(.x = fishingInBalance_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

fishingInBalance_mods_all_coefs_labels <- fishingInBalance_mods_all_coefs %>%
  left_join(fishingInBalance_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*?(?=\\.)"),
         variable = "Fishing in Balance")  %>%
  relocate(region, variable)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

fishingInBalance_mods_all_coefs_labels_collapse <- fishingInBalance_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Trends` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(fishingInBalance_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of biomass ratio variables in NAFO regions in the Eastern and Western Scotian Shelf. Asterisks indicate significant trends (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```


<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Decreases in Fishing in Balance detected here are consistent with previously reported trends in the Northwest Atlantic since the 1970s [@pauly2000ecopath]. Consistent decreases in the Fishing in Balance Index through time suggest that fishing pressure has negatively affected ecosystem productivity in the Scotian Shelf region [@Bundy2017_ecological_indicators]. 

<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:fishingInBalance-variable-table) Definitions of fishing in balance variables within eco_indicators in `marea`.

|variable                    |description |unit |
|:---------------------------|:-----------|:----|
|year                           |Year of trawl survey         |  |
|region                         |Region over which values are summarized         |  |
|FishinginBalance_value       |Fishing in Balance Index|  |



<!----------- Additional Data Section -------------->
## Additional Data

Fishing in Balance Index (within `eco_indicators`) contains data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO divisions are nested within Scotian Shelf regions.


<!----------- Data Access Section -------------->
## Get the Data

```{r get-data-eco_indicators_fishingInBalance, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
