

```{r setup-eco_indicators, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```

# `r data@meta$data_type`
**Data Type:** `r ifelse('sf' %in% class(data@data), 'Spatial Data','Tabular Data')` 

**Spatial Scope:** `r data@meta$region` 

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->


## Introduction to Indicator
<!-- ADD DESCRIPTION -->



## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-eco_indicators, echo=TRUE}
library(tidyr)
library(plotly)
library(stringr)

# function to create plot with dropdown menu ------------------------------
make_metric_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # assign regions to consistent colors
  region_colors <- setNames(hcl.colors(length(regions), palette = "Dark 3"), regions)

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x))

  # build plot -----------------
  p <- plot_ly()

  for (metric_i in seq_along(metrics)) {

    m <- metrics[metric_i]

    # loop to add traces for each region
    for (region_i in regions) {

      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i)

      # add lines
      p <- p %>% add_lines(
        data = dat,
        x = ~get(year_col),
        y = ~value,
        name = as.character(region_i),
        legendgroup = as.character(region_i),
        showlegend = (metric_i == 1), # add legend for first layer
        visible = (metric_i == 1), # keep only first layer visible
        line = list(color = region_colors[[region_i]]),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:,.2f}<extra></extra>")
       #hovertemplate = paste0(
       #  "<b>", pretty_label(metrics[metric_i]), "</b><br>",
       #  year_col, ": %{x}<br>",
       #  region_col, ": ", region_i, "<br>",
       #  "value: %{y:.2f}<extra></extra>"
       #)
      )
    }
  }

  # set keys to make regions visible/invisible
  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  # make buttons ------------------------------------------------
  # each button will toggle visibility of traces 1-6, 7-12, and so on...
  buttons <- purrr::map(
    .x = 1:length(metrics),
    .f = ~{
      # make everything false
      visible <- rep(FALSE, n_traces)
      showlegend <- rep(FALSE, n_traces)

      # index the traces to show for the given layer
      trace_n_start <- (.x - 1) * n_regions + 1
      trace_n_end <- .x * n_regions

      # make traces for the given layer visible
      visible[trace_n_start:trace_n_end] <- TRUE
      showlegend[trace_n_start:trace_n_end] <- TRUE

      # update the layout to show these layers
      list(
        method = "update",
        args = list(
          list(visible = visible, showlegend = showlegend),
          list(
            title = pretty_label(metrics[.x]),
            yaxis = list(title = pretty_label(metrics[.x]))
          )
        ),
        label = pretty_label(metrics[.x])
      )

    }
  )

  # add legend and dropdown to plot
  p %>%
    layout(
      # original title layout for layer 1
      title = pretty_label(metrics[1]),
      xaxis = list(title = year_col),
      yaxis = list(title = pretty_label(metrics[1]),
                   fixedrange = T),
      legend = list(
        title = list(text = "Region"),
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top"
      ),
     hovermode = "x unified",
      
      # add update menus for each dropdown selection
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.18, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}



# usage:
p <- make_metric_dropdown_plot(data@data)
p <- p %>% config(displayModeBar = F)


# use html to pause scrolling on main page while scrolling in dropdown tab
p %>% htmlwidgets::onRender("
function(el, x) {

  const gd = el.querySelector('.js-plotly-plot, .plotly');
  if (!gd) return;

  function hasPlotlyMenuAncestor(node) {
    const menuClasses = new Set([
      'updatemenu', 'updatemenus', 'updatemenu-container',
      'dropdown', 'dropdown-container',
      'scrollbar', 'scrollbar-container'
    ]);

    while (node && node !== gd && node !== document) {
      if (node.classList) {
        for (const c of node.classList) {
          if (menuClasses.has(c)) return true;
        }
      }
      node = node.parentNode;
    }
    return false;
  }

  function wheelHandler(e) {
    if (!gd.contains(e.target)) return;

    if (hasPlotlyMenuAncestor(e.target)) {
      // IMPORTANT:
      e.preventDefault();
    }
  }

  document.addEventListener('wheel', wheelHandler, { capture: true, passive: false });

  document.addEventListener('touchmove', function(e) {
    if (!gd.contains(e.target)) return;
    if (hasPlotlyMenuAncestor(e.target)) e.preventDefault();
  }, { capture: true, passive: false });

}
")

```


## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->




## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

## Variable Definitions

|variable                         |description |unit |
|:--------------------------------|:-----------|:----|
|year                             |ADD         |ADD  |
|region                           |ADD         |ADD  |
|SpeciesRichness_ALL_value        |ADD         |ADD  |
|ShannonDiversity_ALL_value       |ADD         |ADD  |
|MargalefRichness_ALL_value       |ADD         |ADD  |
|PielouEvenness_ALL_value         |ADD         |ADD  |
|HillDiversity_ALL_value          |ADD         |ADD  |
|HillDominance_ALL_value          |ADD         |ADD  |
|Heips_ALL_value                  |ADD         |ADD  |
|SpeciesRichness_ALL_s_value      |ADD         |ADD  |
|ShannonDiversity_ALL_s_value     |ADD         |ADD  |
|MargalefRichness_ALL_s_value     |ADD         |ADD  |
|PielouEvenness_ALL_s_value       |ADD         |ADD  |
|HillDiversity_ALL_s_value        |ADD         |ADD  |
|HillDominance_ALL_s_value        |ADD         |ADD  |
|Heips_ALL_s_value                |ADD         |ADD  |
|ABUNDANCE_ALL_value              |ADD         |ADD  |
|BIOMASS_ALL_value                |ADD         |ADD  |
|BIOMASS_CLUPEIDS_value           |ADD         |ADD  |
|BIOMASS_FINFISH_value            |ADD         |ADD  |
|BIOMASS_FLATFISH_value           |ADD         |ADD  |
|BIOMASS_FORAGE_value             |ADD         |ADD  |
|BIOMASS_GADOIDS_value            |ADD         |ADD  |
|BIOMASS_GROUNDFISH_value         |ADD         |ADD  |
|BIOMASS_PELAGIC_value            |ADD         |ADD  |
|BIOMASS_SKATES_value             |ADD         |ADD  |
|FishinginBalance_value           |ADD         |ADD  |
|ABUNDANCE_ALL_s_value            |ADD         |ADD  |
|BIOMASS_ALL_s_value              |ADD         |ADD  |
|BIOMASS_CLUPEIDS_s_value         |ADD         |ADD  |
|BIOMASS_FINFISH_s_value          |ADD         |ADD  |
|BIOMASS_FLATFISH_s_value         |ADD         |ADD  |
|BIOMASS_FORAGE_s_value           |ADD         |ADD  |
|BIOMASS_GADOIDS_s_value          |ADD         |ADD  |
|BIOMASS_GROUNDFISH_s_value       |ADD         |ADD  |
|BIOMASS_PELAGIC_s_value          |ADD         |ADD  |
|BIOMASS_SKATES_s_value           |ADD         |ADD  |
|FishinginBalance_s_value         |ADD         |ADD  |
|DiversityTargetSpp_ALL_value     |ADD         |ADD  |
|MeanTL.Landings_value            |ADD         |ADD  |
|MTI.Landings_3.25_value          |ADD         |ADD  |
|landings_ALL_value               |ADD         |ADD  |
|landings_CLUPEIDS.L_value        |ADD         |ADD  |
|landings_FINFISH.L_value         |ADD         |ADD  |
|landings_FLATFISH.L_value        |ADD         |ADD  |
|landings_FORAGE.L_value          |ADD         |ADD  |
|landings_GADOIDS.L_value         |ADD         |ADD  |
|landings_GROUNDFISH.L_value      |ADD         |ADD  |
|landings_INVERTEBRATES.L_value   |ADD         |ADD  |
|landings_LARGE_PELAGIC.L_value   |ADD         |ADD  |
|landings_SKATES.L_value          |ADD         |ADD  |
|FP_ALL_value                     |ADD         |ADD  |
|FP_CLUPEIDS_value                |ADD         |ADD  |
|FP_FINFISH_value                 |ADD         |ADD  |
|FP_FLATFISH_value                |ADD         |ADD  |
|FP_FORAGE_value                  |ADD         |ADD  |
|FP_GADOIDS_value                 |ADD         |ADD  |
|FP_GROUNDFISH_value              |ADD         |ADD  |
|FP_SKATES_value                  |ADD         |ADD  |
|DiversityTargetSpp_ALL_s_value   |ADD         |ADD  |
|MeanTL.Landings_s_value          |ADD         |ADD  |
|MTI.Landings_3.25_s_value        |ADD         |ADD  |
|landings_ALL_s_value             |ADD         |ADD  |
|landings_CLUPEIDS.L_s_value      |ADD         |ADD  |
|landings_FINFISH.L_s_value       |ADD         |ADD  |
|landings_FLATFISH.L_s_value      |ADD         |ADD  |
|landings_FORAGE.L_s_value        |ADD         |ADD  |
|landings_GADOIDS.L_s_value       |ADD         |ADD  |
|landings_GROUNDFISH.L_s_value    |ADD         |ADD  |
|landings_INVERTEBRATES.L_s_value |ADD         |ADD  |
|landings_LARGE_PELAGIC.L_s_value |ADD         |ADD  |
|landings_SKATES.L_s_value        |ADD         |ADD  |
|FP_ALL_s_value                   |ADD         |ADD  |
|FP_CLUPEIDS_s_value              |ADD         |ADD  |
|FP_FINFISH_s_value               |ADD         |ADD  |
|FP_FLATFISH_s_value              |ADD         |ADD  |
|FP_FORAGE_s_value                |ADD         |ADD  |
|FP_GADOIDS_s_value               |ADD         |ADD  |
|FP_GROUNDFISH_s_value            |ADD         |ADD  |
|FP_SKATES_s_value                |ADD         |ADD  |
|CVBiomass_value                  |ADD         |ADD  |
|MeanLifespan_value               |ADD         |ADD  |
|BIOMASS_TL2_value                |ADD         |ADD  |
|BIOMASS_TL3_value                |ADD         |ADD  |
|BIOMASS_TL4_value                |ADD         |ADD  |
|MMLength_BIOMASS_value           |ADD         |ADD  |
|MMLength_ABUNDANCE_value         |ADD         |ADD  |
|IVILandings_value                |ADD         |ADD  |
|CVBiomass_s_value                |ADD         |ADD  |
|MeanLifespan_s_value             |ADD         |ADD  |
|BIOMASS_TL2_s_value              |ADD         |ADD  |
|BIOMASS_TL3_s_value              |ADD         |ADD  |
|BIOMASS_TL4_s_value              |ADD         |ADD  |
|MMLength_BIOMASS_s_value         |ADD         |ADD  |
|MMLength_ABUNDANCE_s_value       |ADD         |ADD  |
|IVILandings_s_value              |ADD         |ADD  |
|BIOMASS_LBENTHIVORE_value        |ADD         |ADD  |
|BIOMASS_MBENTHIVORE_value        |ADD         |ADD  |
|BIOMASS_PISCIVORE_value          |ADD         |ADD  |
|BIOMASS_PLANKTIVORE_value        |ADD         |ADD  |
|BIOMASS_ZOOPISCIVORE_value       |ADD         |ADD  |
|PELAGIC_GROUNDFISH_value         |ADD         |ADD  |
|PREDATORS_ALL_value              |ADD         |ADD  |
|LargeFishIndicator_value         |ADD         |ADD  |
|MeanLengthBIOMASS_value          |ADD         |ADD  |
|MeanLengthABUNDANCE_value        |ADD         |ADD  |
|CCondition_FINFISH_value         |ADD         |ADD  |
|CCondition_LBENTHIVORE_value     |ADD         |ADD  |
|CCondition_MBENTHIVORE_value     |ADD         |ADD  |
|CCondition_PISCIVORE_value       |ADD         |ADD  |
|CCondition_PLANKTIVORE_value     |ADD         |ADD  |
|CCondition_ZOOPISCIVORE_value    |ADD         |ADD  |
|BIOMASS_LBENTHIVORE_s_value      |ADD         |ADD  |
|BIOMASS_MBENTHIVORE_s_value      |ADD         |ADD  |
|BIOMASS_PISCIVORE_s_value        |ADD         |ADD  |
|BIOMASS_PLANKTIVORE_s_value      |ADD         |ADD  |
|BIOMASS_ZOOPISCIVORE_s_value     |ADD         |ADD  |
|PELAGIC_GROUNDFISH_s_value       |ADD         |ADD  |
|PREDATORS_ALL_s_value            |ADD         |ADD  |
|LargeFishIndicator_s_value       |ADD         |ADD  |
|MeanLengthBIOMASS_s_value        |ADD         |ADD  |
|MeanLengthABUNDANCE_s_value      |ADD         |ADD  |
|CCondition_FINFISH_s_value       |ADD         |ADD  |
|CCondition_LBENTHIVORE_s_value   |ADD         |ADD  |
|CCondition_MBENTHIVORE_s_value   |ADD         |ADD  |
|CCondition_PISCIVORE_s_value     |ADD         |ADD  |
|CCondition_PLANKTIVORE_s_value   |ADD         |ADD  |
|CCondition_ZOOPISCIVORE_s_value  |ADD         |ADD  |
|BIOMASS_INVERTEBRATES_value      |ADD         |ADD  |
|INVERTEBRATES_GROUNDFISH_value   |ADD         |ADD  |
|FP_INVERTEBRATES_value           |ADD         |ADD  |


## Get the Data
```{r get-data-eco_indicators, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot('eco_indicators')
```
