
<!-- Indicator Chapter for eco_indicators: size -->


```{r setup-eco_indicators_size, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))

data@data  <- data@data %>%
  select(year, region,
         MeanLengthBIOMASS_value,
         MeanLengthABUNDANCE_value
  )

# stability and resistance:
# CVsize_value, # coefficient of variation size
# size_TL2_value
# size_TL3_value
# size_TL4_value

# resouce potential:
# fishing balance
# abundance all?

# fishing pressure:
# DiversityTargetSpp_ALL_value
# MeanTL.Landings_value


# unknown:
# MTI.Landings_3.25_value
# PELAGIC_GROUNDFISH_value
# size_INVERTEBRATES_value
# INVERTEBRATES_GROUNDFISH_value
```



<!------- Header Section -------------->
<!-- Title -->
# Structure & Function: Size Structure of Fish Community

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The `eco_indicators` dataset provides indicators of ecosystem structure with focus on the size structure of the fish community [@Bundy2017_ecological_indicators].

* **Mean Length (Biomass)**: Mean length of fish, weighted by biomass.
* **Mean Length (Abundance)**:  Mean length of fish, weighted by abundance.

Changes in mean length are indicative of changes to community structure and function. Decreases in mean length might indicate pressure from harvesting of large individuals or resource limitations. Alternatively, increases in mean length could indicate increases in resource availability for resident species, increases in metabolism due to environmental change, or filtering of small individuals through recruitment-limiting factors.  

<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-size, fig.cap="Size Structure Indicators in all NAFO regions and Scotian Shelf regions overall. Use dropdown menu to select indicator, and click legend to isolate regions.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_size_dropdown_plot <- function(df,
year_col   = "year",
region_col = "region",
value_suffix = "_value") {
  
  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)
  
  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])
  
  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x))
  
  # build plot -----------------
  p <- plot_ly()
  
  # Add line traces
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) 
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }
  
  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions
  
  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)
    
    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE
    
    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title =  pretty_label(metrics[metric_i]),
                       fixedrange=TRUE)
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })
  
  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = pretty_label(metrics[1]), 
                   fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_size_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)

p


```




<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-eco_indicators_size}

size_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
size_long_split <- size_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
size_means <- purrr::map(
  .x = size_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
size_mods <- purrr::map(
  .x = size_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
size_mods_slopes <- purrr::map(
  .x = size_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
size_mods_confint <- purrr::map(
  .x = size_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
size_mods_p <-  purrr::map(
  .x = size_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = size_mods,
  .y = size_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- size_means %>%
  left_join(size_mods_slopes) %>%
  left_join(size_mods_confint) %>%
  left_join(size_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_all_value","") %>%
           str_replace(.,"_value","") %>%
           str_replace_all("_"," ") %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label) +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Mean Length (cm)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm")))) 


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")



```


Across the Scotian Shelf, fish size has decreased over time. Despite some variation between NAFO zone subunits (Fig. \@ref(fig:plot-eco-indicators-size)), mean size has significantly decreased by both metrics in both Scotian Shelf regions (ESS, WSS) over time (Fig. \@ref(fig:plot-trends-eco-indicators-size)).

```{r plot-trends-eco-indicators-size, echo=F, fig.cap = "Size Structure trends for ESS and WSS"}
p_slopes
```


### Summary Table by NAFO Region and size Variable

Trends of each size variable for each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown below (Table \@ref(tab:make-table-per-zone-size)).

```{r make-table-per-zone-size}

size_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and variable
size_long_split2 <- size_long2 %>% split(f = interaction(.$region,.$variable))
# remove empty list items
sums <- purrr::map(.x = size_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
size_long_split2 <- size_long_split2[sums != 0]

size_mods_all <- purrr::map(.x = size_long_split2,
                            .f = ~lm(data = .x,
                                     value ~ year))
size_mods_all_sums <- purrr::map(.x = size_mods_all,
                                 .f = ~summary(.x))
size_mods_all_coefs <- purrr::map(.x = size_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
size_mods_all_p_val <- purrr::map(.x = size_mods_all_sums,
                                  .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

size_mods_all_coefs_labels <- size_mods_all_coefs %>%
  left_join(size_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*?(?=\\.)"),
         variable = str_replace(id, paste0(region,"."),""),
         variable = str_replace(variable,"\\_ALL_value",""),
         variable = janitor::make_clean_names(variable, allow_dupes=T),
         variable = str_replace_all(variable,"_"," "),
         variable = str_to_title(variable))  %>%
  relocate(region, variable)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

size_mods_all_coefs_labels_collapse <- size_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Trends` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(size_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of size structure variables in NAFO regions in the Eastern and Western Scotian Shelf. Asterisks indicate significant trends (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```


<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

Decreases in mean length of fish are directly relevant to fisheries and stock assessments, although they can be interpreted as both an effect of fisheries activities, and having consequences on future fisheries. 

Decreasing size structure of the fish community indicates that large individuals are not as common as they once were. This trend is, in part, linked to harvesting activities [@bell2018changes], but could also be related to resource limitations and changes in suitability across space. Because large fish have disporportionately higher reproductive output than smaller fish [@barneche2018fish], decreasing average size also has implications for community replenishment and resilience. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:size-variable-table) Definitions of eco_indicators size structure variables.

|variable                    |description |unit |
|:---------------------------|:-----------|:----|
|year                           |Year of trawl survey         |  |
|region                         |Region over which values are summarized         |  |
|MeanLengthBIOMASS_value       |Average length of fish, weighted by biomass| cm |
|MeanLengthABUNDANCE_value   |Average length of fish, weighted by abundance        |cm  |



<!----------- Additional Data Section -------------->
## Additional Data

Size Structure variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_size, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
