
<!-- Indicator Chapter for eco_indicators: Distribution of Fishing Pressure -->


```{r setup-eco_indicators_distFP, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data %>% colnames()
data@data <- data@data %>%
  select(year,
         region,
         DiversityTargetSpp_ALL_value,
         MeanTL.Landings_value,
         MTI.Landings_3.25_value)
```



<!------- Header Section -------------->
<!-- Title -->
# Fishing Pressure: Distribution of Fishing Pressure

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The distribution of fishing pressure is characterized by three proximal indicators.

* **Diversity of Target Species**: the number of species targeted by commercial fisheries; proximal variable for fisheries effort.
* **Mean Trophic Level of Fisheries Landings**: the trophic position of harvested fish.
* **Marine Trophic Index of Fisheries Landings**: tropic average of landed fish, normally with a cutoff (here, 3.5) to capture change in fisheries effort in high-trophic-level species. 




<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-distFP,  fig.cap=paste0("Distribution of Fishing Pressure Variables in Scotian Shelf regions; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan), ". Use dropdown box to select a target trophic level, and click legend to toggle regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_distFP_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- sort(unique(long$metric))
  regions <- unique(long[[region_col]])

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_replace(x, "diversity_target_spp_all","Diversity of Target Species") %>%
    str_replace(.,"mean_tl_landings", "Mean TL of Landings") %>% str_replace(.,"mti_landings_3_25","MTI of Landings")

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) 
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title =pretty_label(metrics[metric_i]))
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = pretty_label(metrics[1]), fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_distFP_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p 

```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco-indicators-distFP}

distFP_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
distFP_long_split <- distFP_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
distFP_means <- purrr::map(
  .x = distFP_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
distFP_mods <- purrr::map(
  .x = distFP_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
distFP_mods_slopes <- purrr::map(
  .x = distFP_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
distFP_mods_confint <- purrr::map(
  .x = distFP_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
distFP_mods_p <-  purrr::map(
  .x = distFP_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = distFP_mods,
  .y = distFP_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- distFP_means %>%
  left_join(distFP_mods_slopes) %>%
  left_join(distFP_mods_confint) %>%
  left_join(distFP_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)

merge <- merge %>%
  mutate(label = 
           recode(group,
                  "DiversityTargetSpp_ALL_value" = "Diversity of Target Species",
                  "MTI.Landings_3.25_value" = "MTI of Landings",
                  "MeanTL.Landings_value" = "Mean TL of Landings")) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = NULL,
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))


```


Although trends between NAFO zones varied over time (Fig. \@ref(fig:plot-eco-indicators-distFP)), overall trends in the Eastern and Western Scotian Shelf regions were mostly consistent (Fig. \@ref(fig:plot-trends-eco-indicators-distFP)). 

In both regions:

* **Diversity of Target Species increased**
* **Mean Trophic Level of Landings significantly decreased**
* **Marine Trophic Index of Landings decreased**, but with a non-significant trend for ESS. 

The increase in diversity of target species indicates a growing distribution of fishing pressure throughout the marine community. Decreases in Mean TL and MTI of landings indicate that harvest activities have shifted away from high trophic level species, although this could be due to previous overharvesting of species at high trophic levels. 

```{r plot-trends-eco-indicators-distFP, fig.cap="Trends in distribution of fishing pressure variables over time in the Eastern and Western Scotian Shelf regions.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in distribution of fishing pressure variables for each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown in the table below (Table \@ref(tab:make-table-per-zone-distFP)). 

```{r make-table-per-zone-distFP}

distFP_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "Variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and TL
distFP_long_split2 <- distFP_long2 %>% split(f = interaction(.$region,.$Variable))
# remove empty list items
sums <- purrr::map(.x = distFP_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
distFP_long_split2 <- distFP_long_split2[sums != 0]

distFP_mods_all <- purrr::map(.x = distFP_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
distFP_mods_all_sums <- purrr::map(.x = distFP_mods_all,
                                    .f = ~summary(.x))
distFP_mods_all_coefs <- purrr::map(.x = distFP_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
distFP_mods_all_p_val <- purrr::map(.x = distFP_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

distFP_mods_all_coefs_labels <- distFP_mods_all_coefs %>%
  left_join(distFP_mods_all_p_val) %>%
  mutate(region = str_extract(id, "^[^.]+"),
         Variable = str_replace(id, paste0(region,"."),""))  %>%
  mutate(Variable = 
           recode(Variable,
                  "DiversityTargetSpp_ALL_value" = "Diversity of Target Species",
                  "MTI.Landings_3.25_value" = "MTI of Landings",
                  "MeanTL.Landings_value" = "Mean TL of Landings")) %>%
  relocate(region, Variable)  %>% 
  mutate(trend_label = 
           paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) ,
                  " ", ifelse(p_val < 0.05,"\u2217","")))

distFP_mods_all_coefs_labels_collapse <- distFP_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, Variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Trends` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(distFP_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = "Change trends for Distribtuion of Fishing Pressure variables across time in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Increases in the diversity of target species represent expansions of fisheries effort across the marine community [@Bundy2017_ecological_indicators]. Expansions of fishing effort across the community could be a result of decreased abundance of historically targeted species. 

Decreases in mean trophic level and Marine Trophic Index oflandings represent shifts from fisheries efforts targeted towards high trophic level species towards species lower in the food chain. These shifts are consistent with the "fishing down the food web" process, where the depletion of large predatory fish has resulted in fishieries expansions towards lower trophic level groups [@pauly1998fishing].

<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|DiversityTargetSpp_ALL_value   |Number of species targeted by commercial fishing           | Number of Species |
|MeanTL.Landings_value   |Mean Trophic Level of landings          | Trophic Level |
|MTI.Landings_3.25_value   |Marine Trophic Index of landings           | Trophic Level |


<!----------- Additional Data Section -------------->
## Additional Data

Distribution of Fishing Pressure (within `eco_indicators`) contains data for `r glue::glue_collapse(data@data %>% distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO divisions are nested within Scotian Shelf regions.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_distFP, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
