
<!-- Indicator Chapter for eco_indicators: CCondition -->


```{r setup-eco_indicators_ccondition, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("CCondition"))
data@data %>% colnames()
```



<!------- Header Section -------------->
<!-- Title -->
# Fish Community Condition

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Fulton's Condition factor [@tw1904rate] is a ratio of body weight to length, or the relative "plumpness" of fish. Fulton's condition is calculated as condition = 100 * weight / length^3, and is used for a proximal for fish health. Higher condition factors represent fish that are heavier at a given length, and are assumed to be in better condition. However, Fulton's condition does not always correlate with more precise measures of fish body condition, especially when the assumotion of isometric growth is violated. 

While comparisons of Fulton's condition between species or subgroups might be hindered by inherent differences in mass to length ratios between group, comparisons of Fulton's condition can be made between individuals within a species in in aggregate for a species over time to capture overall trends or indicate resource limitations or excesses [@ricker1975computation]. Fulton's condition can also change seasonally, representing variation in nutritional resources available throughout the year in seasonal environments.

Although Fulton's condition will depend on the size and shape of individual species, a condition of 1 is generally considered a "normal" condition. Values above 1 represent fish in good condition, and below 1, fish in poor condition. 


<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->

```{r plot-eco-indicators-ccondition, fig.cap="Community condition in all NAFO regions and Scotian Shelf regions overall.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_biomass_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # assign regions to consistent colors
 # region_colors <- setNames(hcl.colors(length(regions), palette = "Dark 3"), regions)

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) %>% gsub("Lb","Large B",.) %>% gsub("Mb","Medium B",.) %>% str_replace("C Condition ","")

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, linetype, linewidth, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text =
            ifelse(region_i == "4VN" & group_name == "ESS",
                   "Eastern Scotian Shelf Zones",
                   ifelse(region_i == "4X",
                          "Western Scotian Shelf Zones",
                          NA))),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Community Condition")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = "Community Condition", fixedrange = TRUE),
      legend = list(
        title = list(text = str_to_title(region_col)),
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem"
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80),
            # add horizontal line at 1
      shapes = list(
        type = "line",
        xref = "paper", 
        yref = "y",     
        x0 = 0,        
        x1 = 1,        
        y0 = 1,         
        y1 = 1,         
        line = list(color = "black", width = 2, dash = "dash") 
      )
    )
}

# usage:
p <- make_biomass_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p

```




<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-eco_indicators_ccondition-2}

ccondition_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
ccondition_long_split <- ccondition_long %>% 
  split(f = interaction(.$region,.$group))



# get overall means
ccondition_means <- purrr::map(
  .x = ccondition_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
    bind_rows(.id = "group")


# model all
ccondition_mods <- purrr::map(
  .x = ccondition_long_split,
  .f = ~lm(data = .x, 
           formula = "value ~ year")
)

# extract slopes
ccondition_mods_slopes <- purrr::map(
  .x = ccondition_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
ccondition_mods_confint <- purrr::map(
  .x = ccondition_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group") 

# extract p values
ccondition_mods_p <-  purrr::map(
  .x = ccondition_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
      bind_rows(.id = "group") %>%
      rename(p_val = value)

augment <- purrr::map2(
  .x = ccondition_mods,
  .y = ccondition_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- ccondition_means %>%
    left_join(ccondition_mods_slopes) %>%
    left_join(ccondition_mods_confint) %>%
    left_join(ccondition_mods_p) %>%
    arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dup = T) %>%
           str_replace(.,"_value","") %>%
           str_replace_all("_"," ") %>%
           str_replace("lb","large b") %>%
           str_replace("mb","medium b") %>%
           str_replace("c condition ","") %>%
           str_to_title()
           ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>% 
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                  "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") + 
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Community Condition",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("ESS" = "purple",
                               "WSS" = "orange")) +
  scale_color_manual(values = c("ESS" = "purple",
                               "WSS" = "orange")) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")

strongest_decrease <- merge %>% filter(trend == min(trend)) %>%
  select(region, label)

strongest_increase <- merge %>% filter(trend == max(trend)) %>%
  select(region, label)

```


Community condition changes through time have differed between regions (Fig. \@ref(fig:plot-eco-indicators-ccondition). Across 6 functional groups, **condition in the Eastern Scotian Shelf is generally `r general_trend_ess`**, and **community condition in the Western Scotian Shelf is more often `r general_trend_wss`**. In the Eastern Scotian Shelf, `r n_increasing_ess` group had an increasing trend (`r n_increasing_sig_ess` significant), and `r n_decreasing_ess` groups had decreasing trends (`r n_decreasing_sig_ess` significant). In the Western Scotian Shelf, `r n_increasing_wss` groups increased in condition through time (`r n_increasing_sig_wss` significant), and `r n_decreasing_wss` groups decreased in condition over time (`r n_decreasing_sig_wss`) (Fig. \@ref(fig:plot-trends-eco-indicators-ccondition)).

The strongest decrease in condition was in the `r strongest_decrease$label` group in the `r strongest_decrease$region` region, and the strongest increase in condition was in the `r strongest_increase$label` in the `r strongest_increase$region` region (Fig. \@ref(fig:plot-trends-eco-indicators-ccondition)). 

```{r plot-trends-eco-indicators-ccondition, echo=F, fig.cap = "Fulton's Condition for fish groups over time in Scotian Shelf"}
p_slopes
```


### Summary Table by NAFO Region and Functional Group

Trends in Community Condition for each functional group in each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) region are shown in the table below (Table \@ref(tab:make-table-per-zone-ccondition)). 

```{r make-table-per-zone-ccondition}

ccondition_long2 <- data@data %>%
  inner_join(global_cols2) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and variable
ccondition_long_split2 <- ccondition_long2 %>% split(f = interaction(.$region,.$variable))
# remove empty list items
sums <- purrr::map(.x = ccondition_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
ccondition_long_split2 <- ccondition_long_split2[sums != 0]

ccondition_mods_all <- purrr::map(.x = ccondition_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
ccondition_mods_all_sums <- purrr::map(.x = ccondition_mods_all,
                                    .f = ~summary(.x))
ccondition_mods_all_coefs <- purrr::map(.x = ccondition_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
ccondition_mods_all_p_val <- purrr::map(.x = ccondition_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)



ccondition_mods_all_coefs_labels <- ccondition_mods_all_coefs %>%
  left_join(ccondition_mods_all_p_val) %>%
  mutate(region = str_extract(id, "4.*?(?=\\.)"),
         variable = str_replace(id, paste0(region,"."),"") %>%
           str_replace("CCondition_","") %>%
           str_replace("_value","") %>%
           str_replace("LB","Large B") %>%
           str_replace("MB", "Medium B") %>%
           str_to_title())  %>%
  relocate(region, variable)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217",""))) 

ccondition_mods_all_coefs_labels_collapse <- ccondition_mods_all_coefs_labels %>%
  left_join(global_cols2) %>%
  group_by(region_group_label, variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>% 
    rename(`Eastern Scotian Shelf ccondition Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf ccondition Trends` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(ccondition_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of ccondition variables in NAFO regions in the Eastern and Western Scotian Shelf. Asterisks indicate significant trends (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Fulton's condition at the scale of fish stocks reveals several insights into stock health. For cod, stocks with higher Fulton's condition factors have higher recruitment potential, and higher biological management reference points  for safe exploitation without collapse (F<sub>med</sub>) [@ratz2003variation]. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:cconditions-variable-table) Definitions of eco_indicators CConditions variables.

|Variable                       |Description |Unit |
|:------------------------------|:-----------|:----|
|year                           |Year of trawl surveys         |  |
|region                         |Region over which data are summarized         |  |
|CCondition_{taxon}_value       |Fulton's Condition Index         |100*Weight/Length<sup>3</sup>  |


<!----------- Additional Data Section -------------->
## Additional Data

CCondition (within `eco_indicators`) contains data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_ccondition, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
