
<!-- Indicator Chapter for eco_indicators: topfoodweb -->


```{r setup-eco_indicators_topfoodweb, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))

data@data  <- data@data %>%
  select(year, region,
        LargeFishIndicator_value, #- structure (top of food web)
        PREDATORS_ALL_value # - structure (top of food web)
         )

```



<!------- Header Section -------------->
<!-- Title -->
# Structure & Function: Top of the Food Web

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The `eco_indicators` dataset provides indicators of ecosystem structure with focus on the top of the food web [@Bundy2017_ecological_indicators].

* **Large Fish Indicator**: Proportion of large fish (>35cm) in the community.
* **Proportion of Predators**: Proportion of predatory fish (feeding on fish or non-planktonic invertebrates).

These indicators provide data on how the marine community is structured, and have relevance to top-down controls on the ecosystem.

<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-topfoodweb, fig.cap="Top of the food web indicators in all NAFO regions and Scotian Shelf regions overall. Use dropdown menu to select indicator, and click legend to toggle regions.", echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols3)

# function to create plot with dropdown menu ------------------------------
make_topfoodweb_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # assign regions to consistent colors
 # region_colors <- setNames(hcl.colors(length(regions), palette = "Dark 3"), regions)

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) 

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]
    
    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, linetype, linewidth, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
      group_name <- unique(dat$region_group)
      color <- unique(dat$color)
      linetype <- unique(dat$linetype)
      width <- unique(dat$linewidth)
      
      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }
      
      p <- p %>% add_lines(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text =
            ifelse(group_name == "ESS",
                   "Eastern Scotian Shelf Zones",
                   "Western Scotian Shelf Zones"
                    )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        line = list(color = color, dash = linetype),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:.3f}<extra></extra>")      )
      
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title =  pretty_label(metrics[metric_i]))
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col)),  # keep one bar per year
      yaxis = list(title = metrics[1], fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = -.1, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_topfoodweb_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p


```




<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco_indicators_topfoodweb}

topfoodweb_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
topfoodweb_long_split <- topfoodweb_long %>% 
  split(f = interaction(.$region,.$group))



# get overall means
topfoodweb_means <- purrr::map(
  .x = topfoodweb_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
    bind_rows(.id = "group")


# model all
topfoodweb_mods <- purrr::map(
  .x = topfoodweb_long_split,
  .f = ~lm(data = .x, 
           formula = "value ~ year")
)

# extract slopes
topfoodweb_mods_slopes <- purrr::map(
  .x = topfoodweb_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
topfoodweb_mods_confint <- purrr::map(
  .x = topfoodweb_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group") 

# extract p values
topfoodweb_mods_p <-  purrr::map(
  .x = topfoodweb_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
      bind_rows(.id = "group") %>%
      rename(p_val = value)

augment <- purrr::map2(
  .x = topfoodweb_mods,
  .y = topfoodweb_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- topfoodweb_means %>%
    left_join(topfoodweb_mods_slopes) %>%
    left_join(topfoodweb_mods_confint) %>%
    left_join(topfoodweb_mods_p) %>%
    arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_all_value","") %>%
           str_replace(.,"_value","") %>%
           str_replace_all("_"," ") %>%
           str_to_title()
           ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>% 
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                  "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label) + 
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Top of the Food Web Values\n(Proportion of Community)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("ESS" = "purple",
                               "WSS" = "orange")) +
  scale_color_manual(values = c("ESS" = "purple",
                               "WSS" = "orange")) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm")))) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_cartesian(ylim = c(0,1))


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")



```


Top of the food web variables are changing across the Scotian Shelf region with noisy trends through time (Fig. \@ref(fig:plot-eco-indicators-topfoodweb)).

The proportion of top of the food web species is **`r general_trend_ess` in the Eastern Scotian Shelf** with `r add_s(n_decreasing_ess,"variable")` showing negative trends through time (`r n_decreasing_sig_ess` significant). In the Western Scotian Shelf, the proportion of large fish has decreased significantly through time, but the proportion of predators has increased (Fig. \@ref(fig:plot-trends-eco-indicators-topfoodweb)).

```{r plot-trends-eco-indicators-topfoodweb, echo=F, fig.cap = "Top of the Food Web trends for ESS and WSS"}
p_slopes
```


### Summary Table by NAFO Region and topfoodweb Variable

Trends of each top of the food web variable for each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown below (Table \@ref(tab:make-table-per-zone-topfoodweb)). 

```{r make-table-per-zone-topfoodweb}

topfoodweb_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "variable",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and variable
topfoodweb_long_split2 <- topfoodweb_long2 %>% split(f = interaction(.$region,.$variable))
# remove empty list items
sums <- purrr::map(.x = topfoodweb_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
topfoodweb_long_split2 <- topfoodweb_long_split2[sums != 0]

topfoodweb_mods_all <- purrr::map(.x = topfoodweb_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
topfoodweb_mods_all_sums <- purrr::map(.x = topfoodweb_mods_all,
                                    .f = ~summary(.x))
topfoodweb_mods_all_coefs <- purrr::map(.x = topfoodweb_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
topfoodweb_mods_all_p_val <- purrr::map(.x = topfoodweb_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

topfoodweb_mods_all_coefs_labels <- topfoodweb_mods_all_coefs %>%
  left_join(topfoodweb_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*?(?=\\.)"),
         variable = str_replace(id, paste0(region,"."),""),
         variable = str_replace(variable,"\\_ALL_value",""),
         variable = janitor::make_clean_names(variable, allow_dupes=T), 
         variable = str_replace_all(variable,"_"," "),
         variable = str_to_title(variable))  %>%
  relocate(region, variable)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217",""))) 

topfoodweb_mods_all_coefs_labels_collapse <- topfoodweb_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, variable) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>% 
    rename(`Eastern Scotian Shelf Trends` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Trends` = `Western Scotian Shelf`,)


library(kableExtra)
knitr::kable(topfoodweb_mods_all_coefs_labels_collapse,
             format = "html", escape = F,
             caption = "Change trends of top of the food web variables in NAFO regions in the Eastern and Western Scotian Shelf. Asterisks indicate significant trends (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```


<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Fishing activities disproportionately target large fish and predator species, so top of the food web indicators are expected to respond to fishing activities in exploited systems. 

In Atlantic Canada, removal of predator species via fishing activity can lead to trophic cascades, affecting marine ecosystems through top-down change [@frank2005trophic]. Declines of large-bodied predator species can result in "fishing down the food web" strategies in marine fisheries, which might result in a phase of increased landings or revenue from lower-trophic level species, but ultimately unsustainability for the marine ecosystem [@pauly1998fishing].



<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                    |description |unit |
|:---------------------------|:-----------|:----|
|year                           |Year of trawl survey         |  |
|region                         |Region over which values are summarized         |  |
|LargeFishIndicator_value       |Proportion of fish >35cm         |Proportion  |
|PREDATORS_ALL_value   |Proportion of predators         |Proportion  |


<!----------- Additional Data Section -------------->
## Additional Data

Top of Food Web Variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_topfoodweb, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
