
<!-- Indicator Chapter for eco_indicators: biomass of Fished groups -->


```{r setup-eco_indicators_biomassFG, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("biomass"))
data@data %>% colnames()
data@data <- data@data %>%
  select(year,
         region,
         BIOMASS_CLUPEIDS_value,
         BIOMASS_FORAGE_value,
         BIOMASS_GADOIDS_value,
         BIOMASS_FLATFISH_value,
         BIOMASS_GROUNDFISH_value
         )
```



<!------- Header Section -------------->
<!-- Title -->
# Resource Potential: Fished Groups

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Resource potential describes the total production capacity of the community, which can be further separated into resource potential of target groups. Harvesting removes resources from a community without replacement, and diminishes the community's ability to replenish resources to previous levels. Thus, resource potential of fished groups can be used as an indicator of the effect of fishing on the community [@Bundy2017_ecological_indicators]. 

Resource potential of fished groups is quantified by the biomass of harvested species, quantified from fisheries-independent surveys. Here, resource potential is quantified from biomass with focus on five harvested groups: 

* Clupeids
* Forage Fish
* Gadoids
* Flatfish
* Groundfish 


<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-biomassFG,  fig.cap=paste0("Biomass of Fished Groups in Scotian Shelf regions; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan), ". Use dropdown box to select a target taxonomic or functional group, and click legend to isolate regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols2)

# function to create plot with dropdown menu ------------------------------
make_biomassFG_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) %>%  gsub("Lb","Large B",.) %>% gsub("Mb","Medium B",.)

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]

    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
        group_name <- unique(dat$region_group)
        color <- unique(dat$color)


      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }

      p <- p %>% add_bars(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        marker = list(color = color),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:,.2s}<extra></extra>")      )
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Biomass (tonnes)")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col), type = "category"),  # keep one bar per year
      yaxis = list(title = "Biomass (tonnes)", fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = 0, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_biomassFG_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p

```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-eco-indicators-biomassFG}

biomassFG_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
biomassFG_long_split <- biomassFG_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
biomassFG_means <- purrr::map(
  .x = biomassFG_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
biomassFG_mods <- purrr::map(
  .x = biomassFG_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
biomassFG_mods_slopes <- purrr::map(
  .x = biomassFG_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
biomassFG_mods_confint <- purrr::map(
  .x = biomassFG_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
biomassFG_mods_p <-  purrr::map(
  .x = biomassFG_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = biomassFG_mods,
  .y = biomassFG_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- biomassFG_means %>%
  left_join(biomassFG_mods_slopes) %>%
  left_join(biomassFG_mods_confint) %>%
  left_join(biomassFG_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("biomass_","") %>%
           str_replace_all("_"," ") %>%
           str_replace("lb","lg b") %>%
           str_replace("mb", "med b") %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Biomass (tonnes)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")

strongest_decrease <- merge %>% filter(trend == min(trend)) %>%
  select(region, label)

strongest_increase <- merge %>% filter(trend == max(trend)) %>%
  select(region, label)

# get just the trend for "all"
trend_all_ess <- merge[merge$region == "ESS" & merge$label == "All",]
trend_all_ess <- paste(ifelse(trend_all_ess$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_ess$trend > 0,"increased","decreased"))

trend_all_wss <- merge[merge$region == "WSS" & merge$label == "All",]
trend_all_wss <- paste(ifelse(trend_all_wss$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_wss$trend > 0,"increased","decreased"))

```

Resource potential of fished groups on the Scotian Shelf has different trends for different target groups and regions (Fig. \@ref(fig:plot-eco-indicators-biomassFG)).

In the Eastern Scotian Shelf, resource potential of fished groups (biomass) decreased for `r add_s(n_decreasing_ess,"group")` (`r n_decreasing_sig_ess` significant) and increased for `r add_s(n_increasing_ess,"group")` (`r n_increasing_sig_ess` significant) In the Western Scotian Shelf, `r add_s(n_decreasing_wss,"group")` decreased (`r n_decreasing_sig_wss` significant), and `r add_s(n_increasing_wss,"group")` increased (`r n_increasing_sig_wss` significant) over time (Fig. \@ref(fig:plot-trends-eco-indicators-biomassFG)). 

The strongest biomass decrease was observed in the `r strongest_decrease$label` group in the `r strongest_decrease$region` region, and the strongest increase was in the `r strongest_increase$label` in the `r strongest_increase$region` region (Fig. \@ref(fig:plot-trends-eco-indicators-biomassFG)).


```{r plot-trends-eco-indicators-biomassFG, fig.cap="Overall Biomass trends for taxonomic and functional groups overtime.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in Biomass over time for each guild and each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown in the table below (Table \@ref(tab:make-table-per-zone-biomassFG)). 

```{r make-table-per-zone-biomassFG}

biomassFG_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "taxon",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and taxon
biomassFG_long_split2 <- biomassFG_long2 %>% split(f = interaction(.$region,.$taxon))
# remove empty list items
sums <- purrr::map(.x = biomassFG_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
biomassFG_long_split2 <- biomassFG_long_split2[sums != 0]

biomassFG_mods_all <- purrr::map(.x = biomassFG_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
biomassFG_mods_all_sums <- purrr::map(.x = biomassFG_mods_all,
                                    .f = ~summary(.x))
biomassFG_mods_all_coefs <- purrr::map(.x = biomassFG_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
biomassFG_mods_all_p_val <- purrr::map(.x = biomassFG_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

biomassFG_mods_all_coefs_labels <- biomassFG_mods_all_coefs %>%
  left_join(biomassFG_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*(?=\\.)"),
         taxon = str_replace(id, paste0(region,".BIOMASS_"),""),
         taxon = str_replace(taxon,"_value",""),
         taxon = str_to_title(taxon))  %>%
  relocate(region, taxon)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

biomassFG_mods_all_coefs_labels_collapse <- biomassFG_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, taxon) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Biomass Trends (tonnes/year)` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Biomass Trends (tonnes/year)` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(biomassFG_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = "Change trends for biomass of fished groups in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Changes in resource potential for for fished groups may indicate the effects of harvesting, harvest management, and ecological factors, but can also indicate the future potential of fisheries. In Atlantic Canada, resource potential metrics are heterogenous between regions and taxa, and can be useful indicators of change at the local scale [@irvine2025spatial].


<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|biomass_{GROUP}_value      |Biomass of select fished groups within regions         | tonnes |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_biomassFG, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
