
<!-- Indicator Chapter for eco_indicators: biomass of Trophic Levels -->


```{r setup-eco_indicators_biomassTL, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("biomass"))
data@data %>% colnames()
data@data <- data@data %>%
  select(c(year, region, 
            contains("TL")))
```



<!------- Header Section -------------->
<!-- Title -->
# Stability & Resistance: Stability of Trophic Community

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Stability of the trophic community is indicative of an ecosystem's stability and resistance [@Bundy2017_ecological_indicators]. Stability of the trophic community is quantified by biomass of fish captured in RV surveys, classified into trophic levels 2, 3 and 4.

<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-biomassTL,  fig.cap=paste0("Biomass of Trophic Levels in Scotian Shelf regions; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan), ". Use dropdown box to select a target trophic level, and click legend to isolate regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols2)

# function to create plot with dropdown menu ------------------------------
make_biomassTL_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value) %>%
    arrange(region,value)

  # find all metrics and regions
  metrics <- sort(unique(long$metric))
  regions <- unique(long[[region_col]])

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("biomass_", "", x)) %>% gsub("Tl","Trophic Level ",.)

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]

    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
        group_name <- unique(dat$region_group)
        color <- unique(dat$color)


      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }

      p <- p %>% add_bars(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones")),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        marker = list(color = color),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:,.2s}<extra></extra>")      )
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Biomass (tonnes)")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col), type = "category"),  # keep one bar per year
      yaxis = list(title = "Biomass (tonnes)", fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = 0, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_biomassTL_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p 

```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*

```{r calc-trends-eco-indicators-biomassTL}

biomassTL_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
biomassTL_long_split <- biomassTL_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
biomassTL_means <- purrr::map(
  .x = biomassTL_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
biomassTL_mods <- purrr::map(
  .x = biomassTL_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
biomassTL_mods_slopes <- purrr::map(
  .x = biomassTL_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
biomassTL_mods_confint <- purrr::map(
  .x = biomassTL_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
biomassTL_mods_p <-  purrr::map(
  .x = biomassTL_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = biomassTL_mods,
  .y = biomassTL_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- biomassTL_means %>%
  left_join(biomassTL_mods_slopes) %>%
  left_join(biomassTL_mods_confint) %>%
  left_join(biomassTL_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("biomass_","") %>%
           str_replace("tl", "Trophic Level ")
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Biomass (tonnes)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")

strongest_decrease <- merge %>% filter(trend == min(trend)) %>%
  select(region, label)

strongest_increase <- merge %>% filter(trend == max(trend)) %>%
  select(region, label)

# get just the trend for "all"
trend_all_ess <- merge[merge$region == "ESS" & merge$label == "All",]
trend_all_ess <- paste(ifelse(trend_all_ess$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_ess$trend > 0,"increased","decreased"))

trend_all_wss <- merge[merge$region == "WSS" & merge$label == "All",]
trend_all_wss <- paste(ifelse(trend_all_wss$p_val < 0.05,"significantly","non-significantly"), ifelse(trend_all_wss$trend > 0,"increased","decreased"))

```


Biomass of Trophic Levels changed over time with opposing patterns in the Eastern and Western Scotian Shelf, indicating structural change (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTL)). 

In the Western Scotian Shelf, the trophic community shifted towards higher trophic levels through time, with TL2 experiencing little change, TL3 significantly decreasing, and TL4 increasing, although with a non-significant linear slope (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTL)). 

In the Eastern Scotian Shelf, the opposite pattern occurred, with TL2 and TL3 increasing significantly over time, and biomass of TL4 decreasing, indicating a shift towards lower trophic level species (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTL)). 

Patterns of trophic level biomass were fairly consistent within individual NAFO regions (Fig. \@ref(fig:plot-eco-indicators-biomassTL)). 



```{r plot-trends-eco-indicators-biomassTL, fig.cap="Overall Biomass trends for Trophic Levels over time.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in Biomass over time for each guild and each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown in the table below (Table \@ref(tab:make-table-per-zone-biomassTL)). 

```{r make-table-per-zone-biomassTL}

biomassTL_long2 <- data@data %>%
  inner_join(global_cols3) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "Trophic Level",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and TL
biomassTL_long_split2 <- biomassTL_long2 %>% split(f = interaction(.$region,.$`Trophic Level`))
# remove empty list items
sums <- purrr::map(.x = biomassTL_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
biomassTL_long_split2 <- biomassTL_long_split2[sums != 0]

biomassTL_mods_all <- purrr::map(.x = biomassTL_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
biomassTL_mods_all_sums <- purrr::map(.x = biomassTL_mods_all,
                                    .f = ~summary(.x))
biomassTL_mods_all_coefs <- purrr::map(.x = biomassTL_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
biomassTL_mods_all_p_val <- purrr::map(.x = biomassTL_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

biomassTL_mods_all_coefs_labels <- biomassTL_mods_all_coefs %>%
  left_join(biomassTL_mods_all_p_val) %>%
  mutate(region = str_extract(id, ".*(?=\\.)"),
         `Trophic Level` = str_replace(id, paste0(region,".BIOMASS_"),""),
         `Trophic Level` = str_replace(`Trophic Level`,"_value",""),
         `Trophic Level` = str_to_title(`Trophic Level`))  %>%
  relocate(region, `Trophic Level`)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

biomassTL_mods_all_coefs_labels_collapse <- biomassTL_mods_all_coefs_labels %>%
  left_join(global_cols3) %>%
  group_by(region_group_label, `Trophic Level`) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Biomass Trends (tonnes/year)` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Biomass Trends (tonnes/year)` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(biomassTL_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = "Change trends for biomass of Trophic Levels in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Assessment of patterns of trophic structure through time reveal patterns of stability in marine food webs. In this region, trophic structure has shifted in both the Eastern and Western Scotian Shelf regions since the 1970s, indicating structural change in the community. Because fishing activities generally target species at higher trophic levels, shifts towards communities dominated by lower trophic level species could be the result of harvest pressure at the higher trophic levels, and shifts in the opposite direction could indicate fisheries recovery or effective management. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|biomass_TL{X}_value      |Biomass of select Trophic Levels within regions         | tonnes |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO divisions are nested within Scotian Shelf regions.


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_biomassTL, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
