
<!-- Indicator Chapter for eco_indicators: biomass of Fished groups -->


```{r setup-eco_indicators_biomassInd, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("biomass"))
data@data %>% colnames()
data@data <- data@data %>%
  select(year,
         region,
         BIOMASS_SKATES_value
         )

timespan <- marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)

```



<!------- Header Section -------------->
<!-- Title -->
# Resource Potential: Indicator Species

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

Biomass of indicator species represent long-term changes to an ecosystem [@Bundy2017_ecological_indicators]. In the Scotian Shelf region, skates can serve as an indicator species for community resource potential. Because are long-lived, and not directly targeted by fisheries, changes in skate populations represent broader shifts in community productivity occurring across long timescales. 



<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-biomassInd,  fig.cap=paste0("Biomass of indicator species in Scotian Shelf regions; ", timespan, ". Click legend to toggle regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols2)



p <- plot_ly()

regions_order <- c("4VN","4VS","4W","4X")

for(r in regions_order){

  df_sub <- plotly_df %>% filter(region == r)

  group_name <- unique(df_sub$region_group)

  p <- p %>%
    add_lines(
      data = df_sub,
      x = ~year,
      y = ~BIOMASS_SKATES_value,
      name = r,
      legendgroup = group_name,
      legendgrouptitle = list(
        text = ifelse(group_name == "ESS",
                     "Eastern Scotian Shelf Zones",
                     "Western Scotian Shelf Zones"
        )
      ),
      line = list(color = unique(df_sub$color), width = 2),
      hovertemplate = paste0("<b>", r,":</b> ","%{y:,.2s}<extra></extra>"
    ))
}

p %>%
  layout(
    title = "Biomass of Skates by Region",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Biomass Skates (tonnes)", fixedrange = TRUE),
    hovermode = "x unified",
    legend = list(
      tracegroupgap = 5,
      groupclick = "toggleitem",
      itemdoubleclick = FALSE
    )
  ) %>%
  config(displayModeBar = FALSE)

```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-eco-indicators-biomassInd}

biomassInd_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
biomassInd_long_split <- biomassInd_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
biomassInd_means <- purrr::map(
  .x = biomassInd_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
biomassInd_mods <- purrr::map(
  .x = biomassInd_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
biomassInd_mods_slopes <- purrr::map(
  .x = biomassInd_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
biomassInd_mods_confint <- purrr::map(
  .x = biomassInd_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
biomassInd_mods_p <-  purrr::map(
  .x = biomassInd_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = biomassInd_mods,
  .y = biomassInd_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- biomassInd_means %>%
  left_join(biomassInd_mods_slopes) %>%
  left_join(biomassInd_mods_confint) %>%
  left_join(biomassInd_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("biomass_","") %>%
           str_replace_all("_"," ") %>%
           str_replace("lb","lg b") %>%
           str_replace("mb", "med b") %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Biomass (tonnes)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010)) +
  geom_hline(yintercept = 0)




ess_trend <- paste0(
  ifelse(merge %>% filter(region == "ESS") %>% pull(p_val) < 0.05,
         "significantly",
         "non-significantly")," ",
  ifelse(merge %>% filter(region == "ESS") %>% pull(trend) > 0,
         "increased", "decreased"))



wss_trend <- paste0(
  ifelse(merge %>% filter(region == "WSS") %>% pull(p_val) < 0.05,
         "significantly",
         "non-significantly")," ",
  ifelse(merge %>% filter(region == "WSS") %>% pull(trend) > 0,
         "increased", "decreased"))
  


```


Resource potential of indicator groups (skates) has shown different trends by region (Fig. \@ref(fig:plot-eco-indicators-biomassInd)). 

Biomass of skates has **`r ess_trend` in the Eastern Scotian Shelf** and **`r wss_trend` in the Western Scotian Shelf** over the sampling duration (`r timespan`) (Fig. \@ref(fig:plot-trends-eco-indicators-biomassInd)).. 


```{r plot-trends-eco-indicators-biomassInd, fig.cap="Overall Biomass trends for taxonomic and functional groups overtime.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in biomass of indicator species (skates) over time for each guild and each NAFO region within the Eastern and Western Scotian Shelf (`r timespan`) are shown in the table below (Table \@ref(tab:make-table-per-zone-biomassInd)). 

```{r make-table-per-zone-biomassInd}

biomassInd_long2 <- data@data %>%
  inner_join(global_cols2) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "taxon",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and taxon
biomassInd_long_split2 <- biomassInd_long2 %>% split(f = interaction(.$region,.$taxon))
# remove empty list items
sums <- purrr::map(.x = biomassInd_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
biomassInd_long_split2 <- biomassInd_long_split2[sums != 0]

biomassInd_mods_all <- purrr::map(.x = biomassInd_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
biomassInd_mods_all_sums <- purrr::map(.x = biomassInd_mods_all,
                                    .f = ~summary(.x))
biomassInd_mods_all_coefs <- purrr::map(.x = biomassInd_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
biomassInd_mods_all_p_val <- purrr::map(.x = biomassInd_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

biomassInd_mods_all_coefs_labels <- biomassInd_mods_all_coefs %>%
  left_join(biomassInd_mods_all_p_val) %>%
  mutate(region = str_extract(id, "4.*(?=\\.)"),
         taxon = str_replace(id, paste0(region,".BIOMASS_"),""),
         taxon = str_replace(taxon,"_value",""),
         taxon = str_to_title(taxon))  %>%
  relocate(region, taxon)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

biomassInd_mods_all_coefs_labels_collapse <- biomassInd_mods_all_coefs_labels %>%
  left_join(global_cols2) %>%
  group_by(region_group_label, taxon) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Biomass Trends (tonnes/year)` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Biomass Trends (tonnes/year)` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(biomassInd_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = "Change trends for biomass of indicator species in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

While skates are not currently targeted as a major fishery in Atlantic Canada, their population changes may serve as a "canary in the coalmine" for the greater ecosystem [Bundy2017_ecological_indicators]. Changes to skate populations may therefore be relevant to monitoring the overall health of marine ecosystems in Atlantic Canada. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|BIOMASS_SKATES_value     |Biomass of skates within regions         | tonnes |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>%distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_biomassInd, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
