
<!-- Indicator Chapter for eco_indicators: biomass of Trophic Guilds -->


```{r setup-eco_indicators_biomassTG, include=FALSE}
library(marea)
library(dplyr)
source("global.R")

knitr::opts_chunk$set(echo = F, message = F)
indicator <- "eco_indicators"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
data@data <- data@data %>%
  select(year, region,contains("biomass"))
data@data %>% colnames()
data@data <- data@data %>%
  select(c(year, region, 
            contains("vore")))
```



<!------- Header Section -------------->
<!-- Title -->
# Structure & Function: Biomass of Trophic Guilds

<!-- Metadata -->
**Data Type:** Tabular Data (within `eco_indicators`)

**Spatial Scope:** Maritimes

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** [`r data@meta$source_citation`](https://publications.gc.ca/collections/collection_2018/mpo-dfo/Fs97-6-3232-eng.pdf)

<!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

One proximal indicator of changes in ecosystem structure is the biomass of communities in trophic guilds [@Bundy2017_ecological_indicators]. Trophic guild bionass is quantified through RV surveys, wherein species are categorized into guilds by their prey. 

Changes in biomass of trophic guilds can indicate changes in predation pressure for prey species, and indicate large-scale patterns of change in the resource flow within ecosystems. 


<!----------- Plots Section -------------->
## View Data

```{r plot-eco-indicators-biomassTG,  fig.cap=paste0("Biomass of Trophic Guilds in Scotian Shelf regions; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan), ". Use dropdown box to select a target taxonomic or functional group, and click legend to toggle regions." ),  echo=TRUE}

library(tidyr)
library(plotly)
library(stringr)

plotly_df <- data@data %>% inner_join(global_cols2)

# function to create plot with dropdown menu ------------------------------
make_biomassTG_dropdown_plot <- function(df,
                                      year_col   = "year",
                                      region_col = "region",
                                      value_suffix = "_value") {

  # convert to long format
  long <- df %>%
    janitor::clean_names() %>%
    pivot_longer(
      cols = ends_with(value_suffix),
      names_to = "metric",
      values_to = "value"
    ) %>%
    # remove suffix
    mutate(
      metric = str_remove(metric, "_value")
    ) %>%
    # drop NAs (some regions don't have data for some variables or years)
    tidyr::drop_na(value)

  # find all metrics and regions
  metrics <- unique(long$metric)
  regions <- unique(long[[region_col]])

  # clean names for dropdown panels, helper
  pretty_label <- function(x) str_to_title(gsub("_", " ", x)) %>%  gsub("Lb","Large B",.) %>% gsub("Mb","Medium B",.) %>% gsub("Biomass ","",.)

  # build plot -----------------
  p <- plot_ly()

  # Add bar traces: metric1 has region1..K, metric2 has region1..K, ...
  for (metric_i in seq_along(metrics)) {
    m <- metrics[metric_i]

    for (region_i in regions) {
      dat <- long %>%
        filter(metric == m, .data[[region_col]] == region_i) %>%
        group_by(.data[[year_col]],color, region_group, region_group_label) %>%     # in case you have multiple rows per year
        summarise(value = sum(value), .groups = "drop") %>%
        arrange(.data[[year_col]])
      
        group_name <- unique(dat$region_group)
        color <- unique(dat$color)

      # If a region truly has no data for that metric, add an empty trace
      # (keeps trace indexing stable)
      if (nrow(dat) == 0) {
        dat <- tibble::tibble(!!year_col := integer(0), value = numeric(0))
      }

      p <- p %>% add_bars(
        data = dat,
        x = ~.data[[year_col]],
        y = ~value,
        name = as.character(region_i),
        legendgroup = group_name,
        legendgrouptitle = list(
          text = ifelse(group_name == "ESS",
                        "Eastern Scotian Shelf Zones",
                        "Western Scotian Shelf Zones"
          )),
        showlegend = (metric_i == 1),
        visible = (metric_i == 1),
        marker = list(color = color),
        hovertemplate = paste0("<b>", region_i,":</b> ","%{y:,.2s}<extra></extra>")      )
    }
  }

  n_regions <- length(regions)
  n_traces  <- length(metrics) * n_regions

  buttons <- lapply(seq_along(metrics), function(metric_i) {
    vis <- rep(FALSE, n_traces)
    shl <- rep(FALSE, n_traces)

    idx_start <- (metric_i - 1) * n_regions + 1
    idx_end   <- metric_i * n_regions
    vis[idx_start:idx_end] <- TRUE
    shl[idx_start:idx_end] <- TRUE

    list(
      method = "update",
      args = list(
        list(visible = vis, showlegend = shl),
        list(
          title = pretty_label(metrics[metric_i]),
          yaxis = list(title = "Biomass (tonnes)")
        )
      ),
      label = pretty_label(metrics[metric_i])
    )
  })

  p %>%
    layout(
      barmode = "stack",
      hovermode = "x unified",
      title = pretty_label(metrics[1]),
      xaxis = list(title = str_to_title(year_col), type = "category"),  # keep one bar per year
      yaxis = list(title = "Biomass (tonnes)", fixedrange = TRUE),
      legend = list(
        x = 1.02, xanchor = "left",
        y = 1,    yanchor = "top",
        groupclick = "toggleitem",
        itemdoubleclick = FALSE
      ),
      updatemenus = list(list(
        type = "dropdown",
        x = 0, xanchor = "left",
        y = 1.15, yanchor = "top",
        buttons = buttons
      )),
      margin = list(r = 180, t = 80)
    )
}

# usage:
p <- make_biomassTG_dropdown_plot(plotly_df)
p <- p %>% config(displayModeBar= F)
  
p 

```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-eco-indicators-biomassTG}

biomassTG_long <- data@data %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "group",
                      values_to = "value") %>%
  left_join(global_cols3) %>%
  filter(region %in% c("ESS","WSS")) %>%
  tidyr::drop_na(value)


# split into list
biomassTG_long_split <- biomassTG_long %>%
  split(f = interaction(.$region,.$group))



# get overall means
biomassTG_means <- purrr::map(
  .x = biomassTG_long_split,
  .f = ~.x %>% summarize(mean = mean(value,na.rm=T))
) %>%
  bind_rows(.id = "group")


# model all
biomassTG_mods <- purrr::map(
  .x = biomassTG_long_split,
  .f = ~lm(data = .x,
           formula = "value ~ year")
)

# extract slopes
biomassTG_mods_slopes <- purrr::map(
  .x = biomassTG_mods,
  .f = ~coefficients(.x)["year"] %>% as_tibble
) %>%
  bind_rows(.id = "group") %>%
  rename(trend = value)

# extract confint
biomassTG_mods_confint <- purrr::map(
  .x = biomassTG_mods,
  .f = ~confint(.x)[2,] %>% setNames(c("lower","upper"))
) %>%
  bind_rows(.id = "group")

# extract p values
biomassTG_mods_p <-  purrr::map(
  .x = biomassTG_mods,
  .f = ~summary(.x)$coefficients[2,"Pr(>|t|)"] %>% as_tibble) %>%
  bind_rows(.id = "group") %>%
  rename(p_val = value)

augment <- purrr::map2(
  .x = biomassTG_mods,
  .y = biomassTG_long_split,
  .f = ~broom::augment(.x, .y, interval = "confidence")
) %>%
  bind_rows()

# merge
merge <- biomassTG_means %>%
  left_join(biomassTG_mods_slopes) %>%
  left_join(biomassTG_mods_confint) %>%
  left_join(biomassTG_mods_p) %>%
  arrange(mean)  %>%
  mutate(region = str_sub(group, 1,3),
         group = str_sub(group, 5,nchar(group)))  %>%
  relocate(region, group) %>%
  left_join(global_cols3)


merge <- merge %>%
  mutate(label = janitor::make_clean_names(group, allow_dupes = T) %>%
           str_replace(.,"_value","") %>%
           str_replace("biomass_","") %>%
           str_replace_all("_"," ") %>%
           str_replace("lb","lg b") %>%
           str_replace("mb", "med b") %>%
           str_to_title()
  ) %>%
  relocate(label)

p_slopes <- augment %>%
  left_join(merge) %>%
  mutate(significant = ifelse(p_val < 0.05,"Significant","Not Significant")) %>%
  ggplot(aes(x = year,
             y = .fitted,
             linetype = significant,
             color = region,
             group = region)) +
  geom_point(aes(y = value), shape = 16, alpha = .3) +
  geom_ribbon(aes(ymin = .lower,
                  ymax = .upper,
                  fill = region),
              alpha = .15, linewidth = 0, show.legend = F) +
  geom_line( linewidth = .75) +
  scale_linetype_manual(breaks = c("Significant","Not Significant"), values= c("Significant" = "solid",
                                                                               "Not Significant"= "longdash"),
                        labels = c("Significant Trend","Non-significant Trend")) +
  facet_wrap(~label, scales = "free_y") +
  ggthemes::theme_few() +
  labs(x = "Year",
       y = "Biomass (tonnes)",
       linetype = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = ss_plot_cols) +
  scale_color_manual(values = ss_plot_cols) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)),
         linetype = guide_legend(theme = theme(legend.key.width = unit(1,"cm"))))  +
  scale_x_continuous(breaks = c(1970, 1990, 2010))


# find increasing trends
n_increasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] > 0)
n_increasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )
n_increasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] > 0)
n_increasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] > 0 )

# find decreasing trends
n_decreasing_wss <- sum(merge$trend[merge$region == "WSS" & merge$label != "All"] < 0)
n_decreasing_sig_wss <- sum(merge$trend[merge$region == "WSS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )
n_decreasing_ess <- sum(merge$trend[merge$region == "ESS" & merge$label != "All"] < 0)
n_decreasing_sig_ess <- sum(merge$trend[merge$region == "ESS" & merge$p_val < 0.05 & merge$label != "All"] < 0 )

general_trend_wss <- ifelse(n_increasing_wss > n_decreasing_wss,"increasing","decreasing")
general_trend_ess <- ifelse(n_increasing_ess > n_decreasing_ess,"increasing","decreasing")

strongest_decrease <- merge %>% filter(trend == min(trend)) %>%
  select(region, label)

strongest_increase <- merge %>% filter(trend == max(trend)) %>%
  select(region, label)

# get just the trend for "all"
# EDIT: biomass_all is not part of this indicator
# trend_all_ess <- merge[merge$region == "ESS" & merge$label == # "All",]
# trend_all_ess <- paste(ifelse(trend_all_ess$p_val < # 0.05,"significantly","non-significantly"), # ifelse(trend_all_ess$trend > 0,"increased","decreased"))
# 
# trend_all_wss <- merge[merge$region == "WSS" & merge$label == # "All",]
# trend_all_wss <- paste(ifelse(trend_all_wss$p_val < # 0.05,"significantly","non-significantly"), # ifelse(trend_all_wss$trend > 0,"increased","decreased"))

```


Biomass values of trophic guilds varied greatly over time and between management areas (Fig. \@ref(fig:plot-eco-indicators-biomassTG)), with biomass of different groups changing and remaining stable between the Eastern and Western Scotian Shelf (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTG)).

For individual guilds, biomass decreased for `r n_decreasing_ess` groups (`r n_decreasing_sig_ess` significant trends) and increased for `r n_increasing_ess` groups (`r n_increasing_sig_ess` significant) in the Eastern Scotian Shelf. In the Western Scotian Shelf, `r n_decreasing_wss` groups had decreasing trends (`r n_decreasing_sig_wss` significant), and `r n_increasing_wss` groups increased (`r n_increasing_sig_wss` significant) over time (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTG)). 

The strongest biomass decrease was observed in the `r strongest_decrease$label` group in the `r strongest_decrease$region` region, and the strongest increase was in the `r strongest_increase$label` in the `r strongest_increase$region` region (Fig. \@ref(fig:plot-trends-eco-indicators-biomassTG)).


```{r plot-trends-eco-indicators-biomassTG, fig.cap="Overall Biomass trends for taxonomic and functional groups overtime.", echo=F}
p_slopes
```


### Summary Table by Region and Functional Group

Trends in Biomass over time for each guild and each NAFO region within the Eastern and Western Scotian Shelf (`r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`) are shown in the table below (Table \@ref(tab:make-table-per-zone-biomassTG)). 

```{r make-table-per-zone-biomassTG}

biomassTG_long2 <- data@data %>%
  inner_join(global_cols2) %>%
  tidyr::pivot_longer(cols = c(3:ncol(data@data)),
                      names_to = "taxon",
                      values_to = "value") %>%
  tidyr::replace_na(list(value = 0))

# split by region and taxon
biomassTG_long_split2 <- biomassTG_long2 %>% split(f = interaction(.$region,.$taxon))
# remove empty list items
sums <- purrr::map(.x = biomassTG_long_split2,
                   .f = ~sum(.x$value)) %>% unlist() %>% unname()
biomassTG_long_split2 <- biomassTG_long_split2[sums != 0]

biomassTG_mods_all <- purrr::map(.x = biomassTG_long_split2,
                               .f = ~lm(data = .x,
                                       value ~ year))
biomassTG_mods_all_sums <- purrr::map(.x = biomassTG_mods_all,
                                    .f = ~summary(.x))
biomassTG_mods_all_coefs <- purrr::map(.x = biomassTG_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Estimate"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(estimate = value)
biomassTG_mods_all_p_val <- purrr::map(.x = biomassTG_mods_all_sums,
                                     .f = ~.x$coefficients[2,"Pr(>|t|)"] %>% as_tibble()) %>%
  bind_rows(.id = "id") %>%
  rename(p_val = value)

biomassTG_mods_all_coefs_labels <- biomassTG_mods_all_coefs %>%
  left_join(biomassTG_mods_all_p_val) %>%
  mutate(region = str_extract(id, "4.*(?=\\.)"),
         taxon = str_replace(id, paste0(region,".BIOMASS_"),""),
         taxon = str_replace(taxon,"_value",""),
         taxon = str_to_title(taxon))  %>%
  relocate(region, taxon)  %>%
  mutate(trend_label = paste0("<b>",region,":</b> ",formatC(estimate, format = "e", digits = 2) , " ", ifelse(p_val < 0.05,"\u2217","")))

biomassTG_mods_all_coefs_labels_collapse <- biomassTG_mods_all_coefs_labels %>%
  left_join(global_cols2) %>%
  group_by(region_group_label, taxon) %>%
  summarize(trend_label = glue::glue_collapse(trend_label, sep = "<br>")) %>%
  tidyr::pivot_wider(names_from = region_group_label,
                     values_from = trend_label) %>%
  rename(`Eastern Scotian Shelf Biomass Trends (tonnes/year)` = `Eastern Scotian Shelf`,
         `Western Scotian Shelf Biomass Trends (tonnes/year)` = `Western Scotian Shelf`,)

library(kableExtra)
knitr::kable(biomassTG_mods_all_coefs_labels_collapse,
             format = "html", escape = FALSE,
             caption = "Change trends for biomass of Trophic Guilds in NAFO regions within the Eastern and Western Scotian Shelf regions. Asterisks indicate significant trends over time (p &lt; 0.05).") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "20%") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "40%")

```



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->



<!----------- Column Definitions Section -------------->
## Variable Definitions

|variable                       |description |unit |
|:------------------------------|:-------------|:----|
|year                           |Year of data collection         |  |
|region                         |Region over which data are summarized         |  |
|biomass_{GROUP}_value      |Biomass of select Trophic Guilds within regions         | tonnes |


<!----------- Additional Data Section -------------->
## Additional Data

Structural Change variables (within `eco_indicators`) contain data for `r glue::glue_collapse(data@data %>% distinct(region) %>% pull(), sep = ", ", last = ", and ")`. All are shown on this page, but note that NAFO zones are nested within Scotian Shelf regions.


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-eco_indicators_biomassTG, include=T, echo=T, eval=F}
library(marea)
data('eco_indicators')
plot(eco_indicators)
```


```{r}
# remove all created objects
rm(list = ls())
```
