

```{r setup-glorys_bottom_temperature, include=FALSE}
library(marea)
library(dplyr)
library(sf)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "glorys_bottom_temperature"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```

# `r data@meta$data_type`
**Data Type:** `r ifelse('sf' %in% class(data@data), 'Spatial Data','Tabular Data')` 

**Spatial Scope:** `r data@meta$region` 

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->


## Introduction to Indicator
<!-- ADD DESCRIPTION -->



## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-glorys_bottom_temperature, echo=TRUE}
#plot(glorys_bottom_temperature, style = 'default')
```


## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-glorys_bottom_temperature, include=F,eval=F}
# calculate most recent value
last_data <-  paste0(last(data@data)$month,'/',last(data@data)$year)
last_val <- last(data@data)[[3]]
# calculate trend over the last 10 years
max_year <- as.numeric(max(data@data$year))
last10 <- data@data[data@data$year %in% c((max_year-10):max_year),]
trend <- lm(last10[[3]] ~ c(1:nrow(last10)))$coefficients[[2]]
# calculate how long value has been same sign as most recent value
same_sign <- data@data %>%
      mutate(
same_sign = sign(.[[3]]) == sign(last_val),
# new group each time value switches signs
        run_id = cumsum(same_sign != lag(same_sign, default = first(same_sign)))
      )
# run_id of the last row 
current_run_id <- same_sign$run_id[nrow(same_sign)]
# first row of *that* run where values are negative
streak_start_row <- same_sign %>%
      filter(run_id == current_run_id) %>%
      slice_head(n = 1)
streak_start_date <- paste0(streak_start_row$month,'/',streak_start_row$year)
```

As of `r max_year`, the glorys_bottom_temperature value was `r last_val`, with an `r ifelse(trend < 0,'decreasing','increasing')` trend over the last 10 years. The value has been `r ifelse(last_val < 0, 'negative','positive')` since `r streak_start_date`.



## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

## Variable Definitions

|variable        |description |unit |
|:---------------|:-----------|:----|
|geometry        |ADD         |ADD  |
|time_descriptor |ADD         |ADD  |
|value           |ADD         |ADD  |
|year            |ADD         |ADD  |
|month           |ADD         |ADD  |


## Get the Data
```{r get-data-glorys_bottom_temperature, include=T, echo=T, eval=T}
library(marea)
data('glorys_bottom_temperature')
glorys_2024 <- ea.subset.spatial(glorys_bottom_temperature, 'year', 2024)

plot(glorys_2024, style = "fill")
```
