
<!-- Indicator Chapter for GLORYS bottom temperature -->


```{r setup-glorys-bottom-temperature, include=FALSE, cache=FALSE}
library(marea)
library(dplyr)
library(purrr)
library(sf)
library(terra)
library(tidyterra)
library(patchwork)

knitr::opts_chunk$set(echo = F, message = F)
source("global.R")
indicator <- "glorys_bottom_temperature"

# get metadata & data for this variable
data("glorys_bottom_temperature")

data <- glorys_bottom_temperature
# convert from sf squares to terra taster
vect <- vect(glorys_bottom_temperature@data)
res <- 0.08333

# make empty raster template
empty_rast <- rast(ext(vect), resolution = res, crs = crs(vect))

time_vals <- unique(vect[["time_descriptor"]][,1])
# loop and rasterize
r_list <- vector(mode = "list", length = length(time_vals))
for(i in 1:length(time_vals)){
 
  r_list[[i]] <- rasterize( vect[vect[["time_descriptor"]] == time_vals[i]],
             empty_rast, field = "value")
  print(i)
  
  names(r_list[[i]]) <- time_vals[i]
   
}

r_stack <- rast(r_list)
rm(r_list)

```



<!------- Header Section -------------->
<!-- Title -->
# GLORYS Gridded Sea Bottom Temperature

<!-- Metadata -->
**Data Type:** Gridded Data

**Spatial Scope:** Northwest Atlantic

**Duration:** 2020-2025

**Source:** `r data@meta$source_citation` 

 <!--**Contact:** ADD CONTACT-->


<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The GLORYS12V1 dataset is a product of the Copernicus Marine Environment Monitoring Service (CMEMS), providing spatiotemporal modeled estimates of environmental data for global oceans at 1/12° horizontal resolution and 50 vertical levels [@copernicus_global_ocean_physics_reanalysis]. Here, we use the bottom temperature estimate for the Northwest Atlantic, encompassing the Scotian Shelf region. 

GLORYS bottom temperature estimates are generated using a numerical ocean model with assimilated satellite and in-situ observations, but see Chapter \@ref(azmp-bottom-temperature) for summarized empirical temperature measurements. 

<!----------- Plots Section -------------->
## View Data

```{r plot-glorys-bottom-temperature, fig.cap="GLORYS bottom temperature data for Scotian Shelf region; 2020-2025. Use pause and play button or slider bar to filter through month/years, and click legend to isolate target NAFO divisions.", echo=TRUE}

library(terra)
library(plotly)

# prepare sf shapes as coordinates
# NAFO regions
#sf <- readRDS("data/derived_data/azmp_area.rds") %>%
#  filter(area %in% c("4X","4Vn","4Vs","4W")) %>%
#  inner_join(global_cols2 %>% rename("area" = "region"))
#sf <- sf %>% st_transform(crs = crs(r_stack))
#sf_coords <- st_coordinates(st_cast(sf, "MULTILINESTRING") ) %>%
#  as.data.frame() %>% 
#  split(.,f = .$L2) %>%
#  purrr::map(.x = .,
#             .f = ~.x %>% rbind(c(NA,NA,NA,NA))) 
#

sf <- readRDS("data/derived_data/nafo_regions_correct.rds") %>%
  filter(region %in% c("4X","4Vn","4Vs","4W")) %>%
  inner_join(global_cols2) %>%
  rename(area = region)
sf <- sf %>% st_buffer(-.05)
sf <- sf %>% st_transform(crs = crs(r_stack)) 
# add NA after multipolygons so that plotly draws lines correctly
sf_coords <- st_coordinates(st_cast(sf, "MULTILINESTRING") ) %>%
  as.data.frame() %>% 
  split(.,f = interaction(.$L2,.$L1), drop = T) %>%
  purrr::map(.x = .,
             .f = ~.x %>% rbind(c(NA,NA,NA,NA))) 
# bind back together by region
name_ids <- sub("\\..*$", "",  names(sf_coords))
grouped <- split(sf_coords, name_ids)
grouped <- purrr::map(.x = grouped, 
                      .f = ~.x %>% bind_rows)

data("coastline")
coast_coords <- st_coordinates(coastline) %>%
    as.data.frame() %>% 
  split(.,f = interaction(.$L1,.$L2,.$L3, drop = T)) %>% 
  purrr::map(.x = .,
             .f = ~.x %>% rbind(c(NA,NA,NA,NA,NA))) %>%
  bind_rows()   


times <- names(r_stack)

# consistent color scale
zmin <- range(values(r_stack), na.rm = TRUE)[1]
zmax <- range(values(r_stack), na.rm = TRUE)[2]

x_vals <- xFromCol(r_stack, 1:ncol(r_stack))
y_vals <- yFromRow(r_stack, 1:nrow(r_stack))
ext <- ext(r_stack)

# Base plot
p <- plot_ly(
  x = x_vals, y = y_vals,
  z =as.matrix(r_stack[[1]], wide = TRUE),
  type = "heatmap",
  colorscale = "YlGnBu",
  hoverinfo = "none",
  colorbar = list(title = "Bottom<br>Temperature<br>(°C)"),
  zmin = zmin, zmax = zmax,
  showscale = TRUE
)

# add polygons
p <- p %>%
    add_polygons(
    x = coast_coords[,1],
    y = coast_coords[,2],
    fillcolor = "#dcdedc",
    showlegend = F,
    line = list(color = "black", width= .2),
    hoverinfo = "none",
    inherit = FALSE
  ) %>% 
  add_paths(x = grouped[[1]][,1],
            y = grouped[[1]][,2],
            line = list(color = unique(sf$color[1]),
                        width = 3),
            hoverinfo = "none",
            inherit = F,
            name = unique(sf$area[1])) %>%
    add_paths(x = grouped[[2]][,1],
            y = grouped[[2]][,2],
            line = list(color = unique(sf$color[2]),
                        width = 3),
            hoverinfo = "none",
            inherit = F,
            name = unique(sf$area[2])) %>%
    add_paths(x = grouped[[3]][,1],
            y = grouped[[3]][,2],
            line = list(color = unique(sf$color[3]),
                        width = 3),
            hoverinfo = "none",
            inherit = F,
            name = unique(sf$area[3])) %>%
    add_paths(x = grouped[[4]][,1],
            y = grouped[[4]][,2],
            line = list(color = unique(sf$color[4]),
                        width = 3),
            hoverinfo = "none",
            inherit = F,
            name = unique(sf$area[4]))


# loop through raster layers to make frames
frames <- purrr::map(
  .x = c(1:nlyr(r_stack)),
  .f = ~list(
    name = times[.x],
    data = list(list(
      z = as.matrix(r_stack[[.x]], wide = TRUE),
      type = "heatmap",
      zmin = zmin,
      zmax = zmax,
      hoverinfo = "none"
    ))
  )
)

# add frames to map
p$x$frames <- frames

# make animation list
steps <- purrr::map(
  .x = c(1:nlyr(r_stack)),
  .f = ~list(
    method = "animate",
    label = times[.x],
    args = list(
      list(times[.x]),
      list(
        mode = "immediate",
        frame = list(duration = 0, redraw = T),
        transition = list(duration = 0)
      )
    )
  )
)

# Slider steps (scrub)
steps <- lapply(seq_along(times), function(i) {
  list(
    method = "animate",
    label  = times[i],
    args   = list(
      list(times[i]),
      list(
        mode = "immediate",
        frame = list(duration = 0, redraw = TRUE),
        transition = list(duration = 0)
      )
    )
  )
})

# Layout + aspect lock + slider + play/pause
p <- p %>%
  layout(
    xaxis = list(
      title = "",
      showgrid = FALSE, zeroline = FALSE,
      range = c(xmin(ext), xmax(ext)),
      constrain = "domain"
    ),
    yaxis = list(
      title = "",
      showgrid = FALSE, zeroline = FALSE,
      range = c(ymin(ext), ymax(ext)),
      scaleanchor = "x",
      scaleratio = 1.5
    ),
    margin = list(l = 0, r = 0, t = 50, b = 70),

    sliders = list(list(
      active = 0,
      y = -.05,
      currentvalue = list(prefix = "Time: "),
      pad = list(t = 10),
      steps = steps
    )),

    updatemenus = list(list(
      type = "buttons",
      direction = "left",
      x = 0.02, y = 1.15,
      xanchor = "left", yanchor = "top",
      pad = list(r = 10, t = 0),
      buttons = list(
        # PLAY: explicitly animate through *all frame names*
        list(
          label = "Play",
          method = "animate",
          args = list(
            as.list(times),
            list(
              mode = "immediate",
              fromcurrent = TRUE,
              frame = list(duration = 250, redraw = TRUE),
              transition = list(duration = 0)
            )
          )
        ),
        # PAUSE
        list(
          label = "Pause",
          method = "animate",
          args = list(
            list(NULL),
            list(
              mode = "immediate",
              frame = list(duration = 0, redraw = FALSE),
              transition = list(duration = 0)
            )
          )
        )
      )
    ))
  )

p %>% config(displayModeBar = F)


```




<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-glorys-bottom-temperature, message=FALSE}

# find mean temperature per cell
glorys_mean <- terra::app(r_stack, "mean")

# find slope of change per cell
slope_fun <- function(x) {
  if (all(is.na(x))) return(NA)
  lm(x ~ c(1:length(x)))$coefficients[2]
}
# since the time series is short (5 years), 
# I don't think it makes sense to regress across all values (which are cyclical).
# instead, filter to august temperatures, which are relevant (summer), 
# and also maximize retained data (e.g., only 4 years available for july)
glorys_august_slope <- terra::app(
  r_stack[[stringr::str_detect(names(r_stack),"-8")]],
                    slope_fun)
names(glorys_august_slope) <- "trend"

glorys_february_slope <- terra::app(
  r_stack[[stringr::str_detect(names(r_stack),"-2$")]],
                    slope_fun)
names(glorys_february_slope) <- "trend"

# extract mean temperature of every cell per NAFO region 
nafo_zone_mean_values <-
  exactextractr::exact_extract(
    glorys_mean, sf,
    progress=F
    ) %>%
  set_names(sf$area) %>%
  bind_rows(.id = "region") 
# find spatial mean within polyons
nafo_zones_mean_means <- nafo_zone_mean_values %>%
  group_by(region) %>%
  summarize(mean = weighted.mean(value,coverage_fraction,  na.rm=T),
            stdev = sd(value, na.rm=T))
  
# find warming trend for every cell within NAFO regions - august
nafo_zones_august_trend_values <-
  exactextractr::exact_extract(
    glorys_august_slope, sf,    
    progress=F
    ) %>%
  set_names(sf$area) %>%
  bind_rows(.id = "region") 
nafo_zones_august_trend_means <- nafo_zones_august_trend_values %>%
  group_by(region) %>%
  summarize(aug_mean_trend = weighted.mean(value,coverage_fraction,  na.rm=T),
            aug_stdev_trend = sd(value, na.rm=T))

# find warming trend for every cell within NAFO regions - february
nafo_zones_frbruary_trend_values <-
  exactextractr::exact_extract(
    glorys_february_slope, sf,    
    progress=F
    ) %>%
  set_names(sf$area) %>%
  bind_rows(.id = "region") 
nafo_zones_february_trend_means <- nafo_zones_frbruary_trend_values %>%
  group_by(region) %>%
  summarize(feb_mean_trend = weighted.mean(value,coverage_fraction,  na.rm=T),
            feb_stdev_trend = sd(value, na.rm=T))

merge <- nafo_zones_mean_means %>%
  left_join(nafo_zones_august_trend_means) %>%
  left_join(nafo_zones_february_trend_means) %>%
  left_join(global_cols2)


p_slopes <- merge %>%
  ggplot(aes(y = region, x = aug_mean_trend,
              xmin = aug_mean_trend - aug_stdev_trend, xmax = aug_mean_trend + aug_stdev_trend ,
              color = mean)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(linewidth = 1) +
  scale_color_viridis_c() +
  ggthemes::theme_few(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title.position = "plot") +
  labs(y = "NAFO Division",
       x = "Mean Temperature Trend of Cells (°C/year)",
       color = "Mean\nTemperature\n(°C)",
       title = "Bottom Temperature Trends per Region") +
    facet_wrap(region_group_label~.,
               ncol = 1,
               scales = "free_y",
               space = "free_y")  


p_slopes_feb <- merge %>%
  ggplot(aes(y = region, x = feb_mean_trend,
              xmin = feb_mean_trend - feb_stdev_trend, xmax = feb_mean_trend + feb_stdev_trend ,
              color = mean)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(linewidth = 1) +
  scale_color_viridis_c() +
  ggthemes::theme_few(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title.position = "plot") +
  labs(y = "NAFO Division",
       x = "Mean Temperature Trend of Cells (°C/year)",
       color = "Mean\nTemperature\n(°C)",
       title = "Bottom Temperature Trends per Region") +
    facet_wrap(region_group_label~.,
               ncol = 1,
               scales = "free_y",
               space = "free_y")  

p_slopes_merge <- (p_slopes +labs(title = "August Temp. Trend")|
                     p_slopes_feb + labs(title = "February Temp. Trend")) +
  plot_layout(guides = "collect",
              axes = "collect") &
  theme(plot.title = element_text(size = 16))


# find number of regions increasing and decreasing
n_regions <- length(unique(merge$region))
n_increasing <- sum(merge$aug_mean_trend > 0)
n_decreasing <- sum(merge$aug_mean_trend < 0)
majority_trend <- ifelse(n_increasing > n_decreasing, "increasing","decreasing")


coldest_nafo <- merge$region[merge$mean == min(merge$mean)]
coldest_nafo_value <- merge$mean[merge$region == coldest_nafo]
warmest_nafo <- merge$region[merge$mean == max(merge$mean)]
warmest_nafo_value <- merge$mean[merge$region == warmest_nafo]

# get summary values
ess_means <-  merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_means <-  merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(mean,2),"°C")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")


ess_trends_summer <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(aug_mean_trend,2)," ± ", round(aug_stdev_trend,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_trends_summer <- merge %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(aug_mean_trend,2)," ± ", round(aug_stdev_trend,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

ess_trends_winter <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(feb_mean_trend,2)," ± ", round(feb_stdev_trend,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

wss_trends_winter <- merge %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",region,"</b>: ", round(feb_mean_trend,2)," ± ", round(feb_stdev_trend,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

# find most recent values
last_date <- names(r_stack)[nlyr(r_stack)]
recent_means <- exactextractr::exact_extract(
  r_stack[[last_date]], sf, c("mean","stdev"), append_cols=T,    
    progress=F)

recent_means_ess <- recent_means %>%
  filter(region_group_label == "Eastern Scotian Shelf") %>%
  mutate(label = paste0("<b>",area,"</b>: ", round(mean,2)," ± ", round(stdev,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

recent_means_wss <- recent_means %>%
  filter(region_group_label == "Western Scotian Shelf") %>%
  mutate(label = paste0("<b>",area,"</b>: ", round(mean,2)," ± ", round(stdev,2),"°C/year ")) %>%
  pull(label) %>% 
  glue::glue_collapse(sep = "<br>")

```


<!---------- Trends Text ----------->

Sea bottom temperatures in the GLORYS dataset are highly heterogenous across space andtime in the Scotian Shelf region, with higher temperatures and greater temporal variability in coastal regions (Fig. \@ref(fig:plot-glorys-bottom-temperature)).

Monthly mean values in this dataset span only a short time period (2020-2025). Among NAFO divisions within the Scotian Shelf, **region `r coldest_nafo` had the lowest mean temperature (`r round(coldest_nafo_value,2)`°C)**, and **region `r warmest_nafo` had the highest mean temperature (`r round(warmest_nafo_value,2)`°C)** (average of individual cell means over time). Due to the short sampling duration and high seasonal variability in this dataset, trends of temperature change within regions are displayed here for summer (August) and winter (February) separately, with target months chosen to maximize data in each time series. In all cases, **temperature decreased over time** (Fig. \@ref(fig:plot-trends-glorys-bottom-temperature)).

```{r plot-trends-glorys-bottom-temperature, fig.cap="Coefficients of GLORYS bottom temperature trends over time (2020-2025) for summer (August) and winter (February) withing NAFO divisions in the Eastern and Western Scotian Shelf regions. Error bars indicate ± 1 standard deviation of individual raster cell trends within each region.", echo=F}
p_slopes_merge
```

Across the greater Scotian Shelf and Northwest Atlantic region, both mean temperature and temperature trend (here, shown as the slope of August temperatures over time) were highly variable across space, with coastal shelf regions cooling on average, and outer shelf regions experiencing little change (Fig. \@ref(fig:plot-trends-glorys-bottom-temperature-map)). 

```{r plot-trends-glorys-bottom-temperature-map, fig.cap = "Map of mean monthly temperatures (left), and summer (month of August) temperature change trends over time (right) for the Scotian Shelf/Northwest Atlantic region. Hover over NAFO divisions to view summary values per region."}


sf_merge <- sf %>%
  rename(region = area) %>%
  left_join(nafo_zones_mean_means) %>%
  left_join(nafo_zones_august_trend_means)

sf_merge <- sf_merge %>%
   mutate(label = paste(
      "<span style='font-size: 12px;'><b>",region, "</b><br>",
      "<b>Mean Monthly SBT:</b>", round(mean,2),"°C<br>",
      "<b>Temperature Trend:</b>", round(aug_mean_trend,2),"°C/year<br>"
    )) %>%
    mutate(label2 = paste0(
    '<div style="',
    'font-size: 11px; ',
    'line-height: 1.5em',
    '">',label,
    '</div>'))


library(mapgl)

# make color scale for means
mean_color_scale <- interpolate_palette(
  data = glorys_mean %>% as.data.frame(),
  column = "mean",    
  method = "equal",
  n = 4,
  palette = viridisLite::viridis
)


m1 <- maplibre(center = c(-62, 44),
         zoom = 3.5) |>
  add_image_source(
    id = "local-raster-source",
    data = glorys_mean %>% project( "epsg:3857")
  ) |>
  add_raster_layer(
    id = "local-raster-layer",
    source = "local-raster-source"
  ) %>%
  add_fill_layer(
    id = "mean_temp",
    source = sf_merge,
    fill_opacity = 0,
    fill_color = "white",
    hover_options = list(fill_opacity = 0.3),
    tooltip = "label2"
  ) %>%
    add_line_layer(
    "ess_poly",
    source = sf_merge %>% filter(region_group =="ESS"),
    line_color = "purple",
    line_width = 3
  ) %>%
      add_line_layer(
    "wss_poly",
    source = sf_merge %>% filter(region_group =="WSS"),
    line_color = "orange",
    line_width = 3
  ) %>%
    # add legend
  mapgl::add_legend(
    "Mean Monthly<br>Sea Bottom Temperature (°C)",
    values = get_legend_labels(mean_color_scale, digits = 2),
    colors = get_legend_colors(mean_color_scale),
    type = "continuous",
    position = "top-left",
    style = list(background_opacity = .8)
  ) 

# repeat for yearly trends ----------------------
# note: there's not a good way to make a diverging palette at zero
# in maplibre, so we're going to have to hack one together.

# find slope limits
trend_range <- range(values(glorys_august_slope), na.rm=T)
trend_min <- trend_range[1]
trend_max <- trend_range[2]

# make a full, equally spaced, blue-white-red palette
trend_ramp_original <- 
  grDevices::colorRampPalette( rev(RColorBrewer::brewer.pal(11, "RdBu")) )(256)
# subset to the cold half, all of which we will be using
trend_ramp_cold <- trend_ramp_original[1:128]
# subset to the warm half, but then we're going to cut it down 
# since the warm values don't go as high in this dataset
trend_ramp_warm <- trend_ramp_original[129:256]
# find how many warm colors we'll need in raster cells
# (the proportion of the palette's magnitude in the warm direction
n_warm <- round(length(trend_ramp_warm) * abs(trend_max) / abs(trend_min))
trend_ramp_warm <- trend_ramp_warm[1:32]
# clip the palette back together, which we'll use to color the image
# (note that this isn't a "real" diverging palette at 0, it's just
# scaled correctly to our data. 
trend_ramp_clipped <- c(trend_ramp_cold, trend_ramp_warm)


m2 <- maplibre(center = c(-62, 44), 
         zoom = 3.5) |> 
  # add the raster, colored by our "fake" diverging palette
  add_image_source( id = "trend-raster", 
                    data = glorys_august_slope, 
                    colors = trend_ramp_clipped ) |>
  add_raster_layer( id = "trend-layer", 
                    source = "trend-raster" ) %>%
    add_fill_layer(
    id = "mean_temp",
    source = sf_merge,
    fill_opacity = 0,
    fill_color = "white",
    hover_options = list(fill_opacity = 0.3),
    tooltip = "label2"
  ) %>%
    add_line_layer(
    "ess_poly",
    source = sf_merge %>% filter(region_group =="ESS"),
    line_color = "purple",
    line_width = 3
  ) %>%
      add_line_layer(
    "wss_poly",
    source = sf_merge %>% filter(region_group =="WSS"),
    line_color = "orange",
    line_width = 3
  ) %>%
  # add a legend using the REAL diverging palette
  mapgl::add_legend( 
    values = c(trend_min, round(trend_min/2), 0, round(abs(trend_min)/2), abs(trend_min)), 
    legend_title = "(August) Bottom Temperature<br>Trend Over Time (°C/year)", 
    colors = trend_ramp_original,
    position = "top-right") 

mapgl::compare(m1, m2)


```

<div style="height: 2rem;"></div>

### Summary Table by Region

Summary values for the GLORYS Sea Bottom Temperature within each individual NAFO division in the Eastern and Western Scotian Shelf are shown below (Table \@ref(tab:glorys-bottom-temperature-trends-table))

Table: (\#tab:glorys-bottom-temperature-trends-table) Summary values of GLORYS Sea Bottom Temperature for NAFO divisions in the Eastern Scotian Shelf (4Vn, 4Vs, 4W), and Western Scotian Shelf (4X).

|Metric      |Eastern Scotian Shelf     |Western Scotian Shelf  |
|:-----------|:------------------|:------------------|
| Overall Mean  |`r ess_means`      |`r wss_means`      |
| Warming Trend<br>(August 2020-2025) |`r ess_trends_summer`  |`r wss_trends_summer`     |
| Warming Trend<br>(Winter 2020-2025)|`r ess_trends_winter`  |`r wss_trends_winter`     |
| Most Recent Value (`r last_date`)|`r recent_means_ess`  |`r recent_means_wss`     |


<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

Sea bottom temperature can strongly affect the populations, distribtuions, and productivity of marine species. See Chapter \@ref(azmp-bottom-temperature) for an overview of known effects in Atlantic Canada. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

GLORYS bottom temperature is stored as a spatial dataframe in `marea`. See variable names and definitions in the table below  (Table \@ref(tab:glorys-bottom-temperature-variable-definitions))

Table: (\#tab:glorys-bottom-temperature-variable-definitions) Column names and definitions in the glorys bottom temperature dataset.


|variable   |description |unit |
|:----------|:-----------|:----|
|geometry       | Spatial polygon denoting the boundaries of 12° grid cells            | Spatial Polygon     |
|time_descriptor     |  Year and month of modeled value          |     |
|value | Modeled sea bottom temperature          | °C    |
|year | Year of modeled value          |   |
|month | Month of modeled value          |   |

<!----------- Additional Data Section -------------->
## Additional Data

Data in the GLORYS Bottom Temperature dataset are gridded with no regional groupings. Summary values displayed on this page were generated, not explicitly provided in the `marea` package. 


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-glorys-bottom-temperature, include=T, echo=T, eval=F}
library(marea)
data('glorys_bottom_temperature')
plot(glorys_bottom_temperature, style = "fill")
```


```{r}
# remove all created objects
rm(list = ls())
```
