# Glorys Bottom Temperature

**Description**: Short description of indicator

**Data Type**: Spatial or tabular

**Spatial Scope**: [Choose one: Coastwide, Regional, Specific]

**Temporal Breadth**: Duration of variable

**Contact(s)**: Name of data contributor including email. 

```{r echo=FALSE, message=F}
knitr::opts_chunk$set(echo = F)
library(marea)
```

## Introduction to Indicator
See "Thermal Habitat Gridded" for cell-based calculations. Many deep water benthic and demersal exhibit thermal preferences for metabolic, reproductive, and growth processes. Temperatures above these thermal presences may impair these processes. Temperature thresholds between 0.5C and 30C at 0.5C intervals. Data from 1993-01-01 to 2024-09-10 were obtained from the CMEMS' GLORYS12V1 global reanalysis product. Daily cell-based metrics include: the number of days exceeding a temperature limit (nDays), the number of consecutive days exceeding a temperature limit (nDays_consecutive), the sum of temperatures above a temperature limit (degree_days), and the proportion of each area that exceeds a temperature limit (percent_area). These daily metrics are summarized with a minimum (_min), maximum (_max), mean (_mean), median (_median), and standard deviation (_sd).

## Key Results and Visualizations
Proceed as though this were a short summary of a typical results section in a paper.


## Indicator statistics 
Spatial scale: by EPU and full shelf (All)

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_glorys_bottom_temperature}
# Either from Contributor or ecodata
```

## Implications
Proceed as though this were a short summary of a typical discussion section in a paper.

## Get the data

**Point of contact**: [Joseph Caracappa (joseph.caracappa@noa.gov)](mailto:Joseph Caracappa (joseph.caracappa@noa.gov)){.email}

**ecodata name**: No dataset

**Variable definitions**

year: year temperature_limit: temperature cutoff used in calculations 
epu: EPU name or ALL for shelf-wide 
nDays_min: minimum number of days in a year where a cell's temperature exceeded temperature_limit 
nDays_max: maximum number of days in a year where a cell's temperature exceeded temperature_limit 
nDays_mean: mean number of days in a year where a cell's temperature exceeded temperature_limit 
nDays_median: median number of days in a year where a cell's temperature exceeded temperature_limit 
nDays_sd: standard deviation of number of days in a year where a cell's temperature exceeded temperature_limit 
nDays_consecutive_min: minimum consecutive days in a year where a cell's temperature exceeded temperature_limit 
nDays_consecutive_max: maximum consecutive days in a year where a cell's temperature exceeded temperature_limit 
nDays_consecutive_mean: mean consecutive days in a year where a cell's temperature exceeded temperature_limit 
nDays_consecutive_median: median consecutive days in a year where a cell's temperature exceeded temperature_limit 
nDays_consecutive_sd: standard deviation of consecutive days in a year where a cell's temperature exceeded temperature_limit 
degree_day_min: minimum degree days in a year where a cell's temperature exceeded temperature_limit 
degree_day_max: maximum degree days in a year where a cell's temperature exceeded temperature_limit 
degree_day_mean: mean degree days in a year where a cell's temperature exceeded temperature_limit 
degree_day_median: median degree days in a year where a cell's temperature exceeded temperature_limit 
degree_day_sd: standard deviation of degree days in a year where a cell's temperature exceeded temperature_limit 
percent_area_min: minimum daily percent of an area where cell's temperature exceeded temperature_limit 
percent_area_max: maximum daily percent of an area where cell's temperature exceeded temperature_limit 
percent_area_mean: mean daily percent of an area where cell's temperature exceeded temperature_limit 
percent_area_median: median daily percent of an area where cell's temperature exceeded temperature_limit 
percent_area_sd: standard deviation of daily percent of an area where cell's temperature exceeded temperature_limit


No Data

**Indicator Category**:

- [X] Extensive analysis, not yet published


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_


