
<!-- Indicator Chapter for npgo: North Pacific Gyre Oscillation -->


```{r setup-npgo, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "npgo"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# North Pacific Gyre Oscillation

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** North Pacific Gyre

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

 <!--**Contact:** ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The North Pacific Gyre Oscillation Index is a fairly new oceanographic index, created to describe changes in salinity, nutrients, and chlorophyll in the North Pacific that were not well-captured by existing indices such as PDO. NPGO is calculated as the second Empirical Orthogonal Function mode of Sea Surface Temperature and Sea Surface Height anomalies in the Northeast Pacific region, closely tracking the second principal component of Sea Surface Temperature anomalies [@di2008north].

A positive NPGO phase indicates the existence of two gyres in the North Pacific, rotating counter to each other, in contrast to the single gyre described by a positive PDO [@di2008north]. 


<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-npgo, echo=TRUE, fig.cap=paste0("NPGO Index Value; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan))}
library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1),
         smooth_1yr = zoo::rollapply(anomaly_value, mean, width = 12, partial = TRUE),
         smooth_10yr = zoo::rollapply(anomaly_value, mean, width = 120, partial = TRUE),
         overall_mean = mean(anomaly_value))


p <- plot_ly(plotly_df, x = ~date) %>%
  add_lines(y = ~anomaly_value,
            name = "Monthly Anomaly",
            line = list(color = "lightgrey"),
            hovertemplate = "Anomaly Value: %{y:.2f} <extra></extra>"
  ) %>%
  add_lines(y = ~smooth_1yr,
            name = "1-yr Smooth",
            line = list(color = "blue", width = 2),
            hovertemplate = "1-yr Smooth: %{y:.2f} <extra></extra>"
  ) %>%
  add_lines(y = ~smooth_10yr,
            name = "10-yr Smooth",
            line = list(color = "red", width = 2),
            hovertemplate = "10-yr Smooth: %{y:.2f} <extra></extra>") %>%
  add_lines(y = ~overall_mean,
            name = "Overall Mean",
            line = list(color = "black", width = 2, dash = "dash"),
            hovertemplate = "Timeseries Mean: %{y:.2f} <extra></extra>") %>%
  layout(
    title = "North Pacific Gyre Oscillation Index",
    xaxis = list(title = "Date",
                 hoverformat = "%b %Y" ),
    yaxis = list(title = "NPGO Index (normalized anomaly)",
                 fixedrange = T),
    hovermode = "x unified",
    margin = list(t = 80)
    
  ) %>%
  config(displayModeBar = FALSE)

p 
```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-npgo, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
last_anomaly <- last(data@data)[[value_col]]
current_phase <- ifelse(last_anomaly < 0,"negative","positive")

streak_start <- switch(
  current_phase,
  "negative" = last(which(data@data[value_col] > 0)),
  "positive" = last(which(data@data[value_col] < 0))) + 1
streak_start_date <- ifelse(
  has_months,
  paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
  paste0(data@data[streak_start,"year"]))


```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the NPGO value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The NPGO is **currently in a `r current_phase` phase**, and has been since `r streak_start_date`.
The NPGO value has followed a **`r current_trend` trend** in recent years.



<!--------- Trends Table--------->
### Summary Table
Summary values for the North Pacific Gyre Oscillation Index are found in the table below (Table \@ref(tab:npgo-trends-table))

Table: (\#tab:npgo-trends-table) Summary table of NPGO values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r round(last_val,3)` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. |
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, npgo has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, npgo has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Timeseries Record High | `r round(highest_val,3)` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Timeseries Record Low | `r round(lowest_val,3)` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments

In Canada, NPGO has shown positive relationships with salmon productivity in Pacific regions [@malick2017effects]. In Atlantic Canada, NPGO is not expected to be a major driver of fisheries productivity or ecological patterns.  


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:npgo-variable-definitions) Column names and definitions in the NPGO dataset.

|variable      |description |unit |
|:-------------|:-----------|:----|
|year          |Year of projected value         |  |
|month         |Month of projected value         |  |
|anomaly_value |Normalized NPGO anomaly     | Standard deviation units |


<!----------- Additional Data Section -------------->
## Additional Data

No additional data for NPGO.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-npgo, include=T, echo=T, eval=F}
library(marea)
data('npgo')
plot(npgo)
```


```{r}
# remove all created objects
rm(list = ls())
```
