
<!-- Indicator Chapter for oni: Oceanic Niño Index -->


```{r setup-oni, include=FALSE}
library(marea)
library(dplyr)
source("global.R")
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "oni"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Oceanic Niño Index

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Niño 3.4 Region (Pacific)

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The Oceanic Niño Index (ONI) is one measure of the El Niño Southern Oscillation pattern in the Pacific. ONI measures the 3-month running-mean anomaly of sea surface temperature in the Central Pacific from 30-year baselines that are updated every 5 years [@glantz2020reviewing]. 

High ONI anomaly values for prolonged periods indicate a strong likelihood that an El Niño event will occur. 




<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-oni, fig.cap=paste0("ONI index values; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}
library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1)) %>%
  mutate(color = ifelse(anomaly > 0,"#d63227","#3e5eb8"))

p <- plot_ly(plotly_df, x = ~date,y = ~anomaly) %>%
  add_bars(marker = list(color = ~color),
           hovertemplate = "Anomaly Value: %{y:.2f}°C <extra></extra>")  %>%
    layout(
    title = "Oceanic Niño Index for Niño 3.4 region",
    xaxis = list(title = "Date",
                 hoverformat = "%b %Y" ),
    yaxis = list(title = "3-month running mean anomaly (°C)",
                 fixedrange = T),
    hovermode = "x unified",
    margin = list(t = 80)
    
  ) %>%
  config(displayModeBar = FALSE)

p

```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*


```{r calc-trends-oni, fig.cap=paste0("ONI Index Value; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan))}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
last_anomaly <- last(data@data)[["anomaly"]]
current_phase <- ifelse(last_anomaly < 0,"negative","positive")

streak_start <- switch(
  current_phase,
  "negative" = last(which(data@data["anomaly"] > 0)),
  "positive" = last(which(data@data["anomaly"] < 0))) + 1
streak_start_date <- ifelse(
  has_months,
  paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
  paste0(data@data[streak_start,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the ONI index is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The ONI is **currently in a `r current_phase` phase**, and has been since `r streak_start_date`.
The ONI value has followed a(n) **`r current_trend` trend** in recent years (Fig. \@ref(fig:plot-oni)).



<!--------- Trends Table--------->
### Summary Table
Summary values for the Oceanic Niño Index are found in the table below (Table \@ref(tab:oni-trends-table))

Table: (\#tab:oni-trends-table) Summary table of ONI values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values.|
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, oni has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, oni has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Timeseries Record High | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Timeseries Record Low | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

ONI and related El Niño indicators have strong effects on fisheries and oceanic and atmopheric processes in the Pacific, but are not expected to be major drivers of change in Atlantic Canada.

<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:oni-variable-definitions) Column names and definitions in the ONI dataset.

|variable    |description |unit |
|:-----------|:-----------|:----|
|year        |Year of model-projected value   |  |
|month       |Month of model-projected value   |  |
|value_value |3-month running mean of sea surface temperature in Niño 3.4 region (Central Pacific)         | °C  |
|anomaly     |ONI anomaly         | °C (deviation from average) |

<!----------- Additional Data Section -------------->
## Additional Data

No additional data for ONI.



<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-oni, include=T, echo=T, eval=F}
library(marea)
data('oni')
plot(oni)
```
