
<!-- Indicator Chapter for pdo: Pacific Decadal Oscillation -->


```{r setup-pdo, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "pdo"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Pacific Decadal Oscillation

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** North Pacific

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The Pacific Decadal Oscillation is a recurring pattern of sea surface temperature and air pressure variability in the Pacific Ocean, north of 20° latitude. Like the ENSO, the PDO varies on an interannual-to-interdecadal scale, and its pattern has broad affects on oceanographic and atmospheric processes in the North Pacific [@mantua1997pacific]. 

<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-pdo, fig.cap=paste0("PDO Index Value; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=T, echo=TRUE}

library(plotly)
plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1)) %>%
  mutate(color = ifelse(anomaly_value > 0,"#d63227","#3e5eb8"))

p <- plot_ly(plotly_df, x = ~date,y = ~anomaly_value) %>%
  add_bars(marker = list(color = ~color),
           hovertemplate = "Anomaly Value: %{y:.2f}°C <extra></extra>")  %>%
    layout(
    title = "Pacific Decadal Oscillation for North Pacific",
    xaxis = list(title = "Date",
                 hoverformat = "%b %Y" ),
    yaxis = list(title = "PDO Index",
                 fixedrange = T),
    hovermode = "x unified",
    margin = list(t = 80)
    
  ) %>%
  config(displayModeBar = FALSE)

p


```



<!----------- Trends Section -------------->
## Summary and Trends
*Trend and summary values are automatically generated; data were last updated on marea package install on `r marea_install_date`*



```{r calc-trends-pdo, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether the current value is above or below the mean, and how long it has been
current_vs_mean <- last_val - overall_mean
current_vs_mean_text <- ifelse(sign(current_vs_mean) == 1,"above","below")
last_different_sign <- switch(current_vs_mean_text,
                                  "above" = last(which(data@data[value_col] < overall_mean)),
                                  "below" = last(which(data@data[value_col] > overall_mean)))
last_date_different_sign <- ifelse(has_months,
                                       paste0(data@data[last_different_sign,"month"],"/",data@data[last_different_sign,"year"]),
                                       paste0(data@data[last_different_sign,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the PDO value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The PDO value is currently **`r current_vs_mean_text` the long-term mean**, and has been **since `r last_date_different_sign`**.
The PDO value has followed a(n) **`r current_trend` trend** in recent years (Fig. \@ref(fig:plot-pdo)).



<!--------- Trends Table--------->
### Summary Table
Summary values for the Pacific Decadal Oscillation Index are found in the table below (Table \@ref(tab:pdo-trends-table))

Table: (\#tab:pdo-trends-table) Summary table of PDO values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. |
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, pdo has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, pdo has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Timeseries Record High | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Timeseries Record Low | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

The PDO affects several processes in the Pacific ocean and is linked to population dynamics of important harvested species such as salmon [@mantua1997pacific]. In the Canadian Atlantic, PDO is not expected to be a major driver of fisheries changes. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:pdo-variable-definitions) Column names and definitions in the SOI dataset.

|variable      |description |unit |
|:-------------|:-----------|:----|
|year          |Year of model-projected value         |  |
|month         |Month of model-projected value         |  |
|anomaly_value |PDO Index         |  |



<!----------- Additional Data Section -------------->
## Additional Data

No additional data are available for PDO.

<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-pdo, include=T, echo=T, eval=F}
library(marea)
data('pdo')
plot(pdo)
```


```{r}
# remove all created objects
rm(list = ls())
```
