
<!-- Indicator Chapter for soi: Southern Oscillation Index -->


```{r setup-soi, include=FALSE}
library(marea)
library(dplyr)
knitr::opts_chunk$set(echo = F, message = F)
indicator <- "soi"

# get metadata & data for this variable
data <- get(data(list = indicator, package = 'marea'))
```



<!------- Header Section -------------->
<!-- Title -->
# Southern Oscillation Index

<!-- Metadata -->
**Data Type:** Tabular Data

**Spatial Scope:** Equatorial Pacific

**Duration** `r marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)`

**Source:** `r data@meta$source_citation` 

**Contact:** <!--ADD CONTACT-->



<!----------- Intro Section -------------->
## Introduction to Indicator
<!-- ADD DESCRIPTION -->

The Southern Oscillation Index describes differences in sea level pressure in the Southern Pacific, between Tahiti and Darwin, Australia, normalized by by monthly standard deviations, and correlates with changes in ocean temperature in the South Pacific [@chen1982assessment]. 

Positive phases of the SOI describe lower-than-average pressure at Darwin and higher-than-average pressure at Tahiti, and negative phases describe the opposite. Negative phases are indicative of El Niño conditions including warm water in the Pacific, and positive phases of La Niña. 

<!----------- Plots Section -------------->
## View Data
<!-- USE DEFAULT MAREA PLOT OR REPLACE WITH CUSTOM -->
```{r plot-soi, fig.cap=paste0("SOI index values; ",marea_metadata() %>% filter(Dataset == indicator) %>% pull(TimeSpan)), echo=TRUE}
library(plotly)

plotly_df <- data@data %>%
  mutate(date = lubridate::make_date(year, month, day = 1)) %>%
  mutate(color = ifelse(anomaly_value < 0,"#d63227","#3e5eb8"))

p <- plot_ly(plotly_df, x = ~date,y = ~anomaly_value) %>%
  add_bars(marker = list(color = ~color),
           hovertemplate = "Anomaly Value: %{y:.2f} <extra></extra>")  %>%
    layout(
    title = "Southern Oscillation Index for Equatorial Pacific",
    xaxis = list(title = "Date",
                 hoverformat = "%b %Y" ),
    yaxis = list(title = "SOI Index",
                 fixedrange = T),
    hovermode = "x unified",
    margin = list(t = 80)
    
  ) %>%
  config(displayModeBar = FALSE)

p

```



<!----------- Trends Section -------------->
## Trends
<!-- ADD SPECIFICATION TO DEFAULTS -->

```{r calc-trends-soi, include=F}

# assess dataset
# find the most recent recoded value, and its relation to broader data:
last_row <- last(data@data)
value_col <- colnames(data@data)[first(which(stringr::str_detect(colnames(data@data), "value")))]
has_months <- "month" %in% colnames(last_row)
has_regions <- "region" %in% colnames(last_row)

# calculate most recent value
last_date <- ifelse(has_months,
                        paste0(last_row$month,"/",last_row$year),
                        paste0(last_row$year))
last_val <- last(data@data)[[value_col]]
last_percentile <- round(ecdf(data@data[[value_col]])(last_val) * 100,2)

# calculate trend over the last 10 years
max_year <- max(data@data$year)
max_year_minus_ten <- max_year - 10
min_year <- min(data@data$year)
last_ten <- data@data[data@data$year %in% c((max_year_minus_ten):max_year),]
last_ten_mod <- lm(last_ten[[value_col]] ~ c(1:nrow(last_ten)))
last_ten_slope <- last_ten_mod$coefficients[[2]]
last_ten_slope <- ifelse(has_months, last_ten_slope * 12, last_ten_slope)
last_ten_direction <- ifelse(last_ten_slope > 0,"increased","decreased")
last_ten_p <- summary(last_ten_mod)$coefficients[2,"Pr(>|t|)"]
last_ten_p_annotate <-
      if(last_ten_p <= .01) "Very Strong" else
        if(last_ten_p <= .05) "Strong" else
          if(last_ten_p <= .1) "Marginal" else
            "Nonsignificant"
last_ten_p_annotate2 <-
      if(last_ten_p <= .01) "p < 0.01" else
        if(last_ten_p <= .05) "p < 0.05" else
          if(last_ten_p <= .1) "p < 0.1" else
            "p > 0.1"



# calculate trend over full timeseries
full_mod <- lm(data@data[[value_col]] ~ c(1:nrow(data@data)))
full_mod_slope <- full_mod$coefficients[[2]]
full_mod_slope <- ifelse(has_months,full_mod_slope * 12, full_mod_slope)
full_mod_p <- summary(full_mod)$coefficients[2,"Pr(>|t|)"]
full_mod_p_annotate <-
      if(full_mod_p <= .01) "Very Strong" else
        if(full_mod_p <= .05) "Strong" else
          if(full_mod_p <= .1) "Marginal" else
            "Nonsignificant"
full_mod_p_annotate2 <-
      if(full_mod_p <= .01) "p < 0.01" else
        if(full_mod_p <= .05) "p < 0.05" else
          if(full_mod_p <= .1) "p < 0.1" else
            "p > 0.1"
full_mod_direction <- ifelse(full_mod_slope > 0,"increased","decreased")



# calculate highest point in time series
highest_row <- data@data[which.max(data@data[[value_col]]),]
highest_timepoint <- ifelse(has_months,
                                paste0(highest_row$month,"/",highest_row$year),
                                highest_row$year)
highest_val <- highest_row[[value_col]]



# calculate lowest point in time series
lowest_row <- data@data[which.min(data@data[[value_col]]),]
lowest_timepoint <- ifelse(has_months,
                               paste0(lowest_row$month,"/",lowest_row$year),
                               lowest_row$year)
lowest_val <- lowest_row[[value_col]]



# find current values
current_trend <- ifelse(last_ten_slope > 0,"increasing", "decreasing")
current_magnitude <-
      if(last_percentile > 70) "high" else
        if(last_percentile < 30) "low" else
          "moderate"



# find overall mean
overall_mean <-  mean(data@data[[value_col]])
highest_vs_mean <- round(highest_val - overall_mean,2)
lowest_vs_mean <- round(overall_mean - lowest_val,2)



# find whether SOI is currently in a positive or negative phase
last_anomaly <- last(data@data)[[value_col]]
current_phase <- ifelse(last_anomaly < 0,"negative","positive")

streak_start <- switch(
  current_phase,
  "negative" = last(which(data@data[value_col] > 0)),
  "positive" = last(which(data@data[value_col] < 0))) + 1
streak_start_date <- ifelse(
  has_months,
  paste0(data@data[streak_start,"month"],"/",data@data[streak_start,"year"]),
  paste0(data@data[streak_start,"year"]))

```



<!---------- Trends Text ----------->
As of the most recent data entry in **`r last_date`**, the SOI value is `r last_val`, which is **`r current_magnitude`** among values in the timeseries.
The AMO is **currently in a `r current_phase` phase**, and has been since `r streak_start_date`.
The SOI value has followed a(n) **`r current_trend` trend** in recent years (Fig. \@ref(fig:plot-soi)).



<!--------- Trends Table--------->
### Summary Table
Summary values for the Southern Oscillation Index are found in the table below (Table \@ref(tab:soi-trends-table))

Table: (\#tab:soi-trends-table) Summary table of SOI values through time.

|Metric      |Value     |Description          |
|:-----------|:---------|:--------------------|
|Most Recent Value (`r last_date`)  | `r last_val` |The most recent value is **`r current_magnitude`** within the timeseries, in the `r last_percentile` percentile of all values. |
|Recent Trend (`r max_year_minus_ten` -- `r max_year`)  | `r round(last_ten_slope,3)` per year | In the last ten years, soi has **`r last_ten_direction`** by `r round(last_ten_slope,3)` per year with a **`r last_ten_p_annotate` trend** (`r last_ten_p_annotate2`). |
|Overall Trend (`r min_year` -- `r max_year`)  | `r round(full_mod_slope,3)` per year | Since the beginning of the timeseries, soi has **`r full_mod_direction`** by `r round(full_mod_slope,3)` per year with a **`r full_mod_p_annotate` trend** (`r full_mod_p_annotate2`). |
|Highest Value | `r highest_val` | The highest value in the timeseries was recorded on **`r highest_timepoint`**, and was **`r highest_vs_mean`** higher than the overall timeseries mean. |
|Lowest Value | `r lowest_val` | The lowest value in the timeseries was recorded on **`r lowest_timepoint`**, and was **`r lowest_vs_mean`** lower than the overall timeseries mean. |



<!----------- Relevance Section -------------->
## Relevance to Research and Stock Assessments
<!-- ADD RELEVANCE-->

The SOI is linked to several oceanographic processes in the Pacific and has several potential impacts on tropical Pacific fisheries [@harrison2015multi]. However, it is not expected to be a major driver of fisheries in the Canadian Atlantic. 


<!----------- Column Definitions Section -------------->
## Variable Definitions

Table: (\#tab:soi-variable-definitions) Column names and definitions in the SOI dataset.

|variable      |description |unit |
|:-------------|:-----------|:----|
|year          |Year of model-projected value         |  |
|month         |Month of model-projected value         |  |
|anomaly_value |SOI Index         |  |



<!----------- Additional Data Section -------------->
## Additional Data

No additional data are available for SOI. 


<!----------- Data Access Section -------------->
## Get the Data
```{r get-data-soi, include=T, echo=T, eval=F}
library(marea)
data('soi')
plot(soi)
```


```{r}
# remove all created objects
rm(list = ls())
```
